# Linear Algebra for Fitting Models to Data {#sec-la-intro}

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo    = FALSE, 
  error   = FALSE, 
  message = FALSE, 
  warning = FALSE
)
```

```{r}
#| label: CRAN-libraries

library(assertthat)
library(dslabs)
library(GGally)
library(ggimg)
library(HistData)
library(here)
library(keras)
library(latex2exp)
library(OECD)
library(plotly)
library(rgl)
library(R.utils)
library(reticulate)
library(scatterplot3d)
library(tensorflow)
library(tidyverse)
library(tinytex)
library(UsingR)
```

```{r}
#| label: local-source

source(here("code", "galton_ht_data.R"))
source(here("code", "mnist_file_mgmt.R"))
source(here("code", "oecd_bli.R"))
source(here("code", "pgram_grid.R"))
source(here("code", "z_score.R"))
```

------------------------------------------------------------------------

Vectors and matrices are the central objects of linear algebra.  And central to data science and machine learning is the notion of a _data matrix_, in which each row is composed of different types of values that represent a single case of data.  For example, a single row of a data matrix might represent the recorded characteristics of an individual participant, item, or unit in some study.  In contrast, each column (known as a _feature vector_ or _data variable_) represents multiple instances of just one of these prescribed types of value. [^data-variable]  Let's consider some examples.

[^data-variable]: To be more precise, every _data variable_ qualifies as a _feature vector_, but a feature vector may also be some function of the data and other information.

## Data Examples

### Heights of Parents and Oldest Child

```{r}
#| label: galton_3d

galton_3d_lst <- get_galton_3d()

# tibble(father, mother, child, gender)
# (child, gender) refers to oldest child
galton_3d <- galton_3d_lst$ galton_3d

# cite the following values in the narrative
n_children_all      <- 
  galton_3d_lst$ counts_lst$ n_children_all
n_families          <- 
  galton_3d_lst$ counts_lst$ n_families
n_child_1_daughters <- 
  galton_3d_lst$ counts_lst$ n_child_1_daughters
n_child_1_sons      <- 
  galton_3d_lst$ counts_lst$ n_child_1_sons

```

In 1885 Sir Francis Galton examined the heights (in inches) of parents and their adult children to determine the strength of evidence to support height as a hereditary trait.  The corresponding `R` data set `HistData::GaltonFamilies` consists of `r n_children_all` adult children from a total of `r n_families` families.  Restricting attention to the oldest child in each family, there were `r n_child_1_daughters` daughters and `r n_child_1_sons` sons.

The table below shows a portion of this data matrix.  Each row represents a family and consists of: a family identifier, the father's height, the mother's height, the oldest child's height, and the oldest child's gender.

```{r}
#| label: tbl-galton-3d
#| tbl-cap: "Family heights in inches: father, mother, oldest child"

galton_3d |> 
  dplyr::slice_head(n = 6) |> 
  knitr::kable(
  caption = "Family heights: father, mother, oldest child"
)
```

The figure below represents all the families, with the gender of the oldest child distinguished by color: red for daughters and blue for sons.

```{r}
#| label: fig-mfc-scat3d
#| fig-cap: "Height of oldest child: daughters (red), sons (blue)"

g_mfc_s3d <- scatterplot3d::scatterplot3d(
  x = galton_3d$ mother, xlab = "mother", 
  y = galton_3d$ father, ylab = "father", 
  z = galton_3d$ child,  zlab = "child", 
  pch = 16, 
  color = dplyr::if_else(
    galton_3d$ gender == "male", "steelblue", "red"), 
  main = "Family heights: mother, father, child", 
  sub  = "daughters in red, sons in blue"
)
```

In @sec-conditioning we regressed the son's height on the father's height.  We obtained the regression line, which approximates the graph of averages: the average son's height per father's height.  The linear regression can be interpreted as a linear prediction of the height of a son whose father is of some given height.

We can now expand on this idea by regressing the son's height on the heights of both the mother and the father.  This is a model in which the predicted son's height, $\hat{s}$, is some constant plus some linear combination of the parents' heights.

$$
\begin{align}   
  \hat{s} & = \mathcal{l}_{R}(m, f) \\   
  &= \beta_0 \; + \; \beta_m \times m \; + \; \beta_f \times f 
\end{align} 
$$ {#eq-s-per-mf-regression-plane}

where 

$$
\begin{align} 
  \hat{s} &= \text{predicted height of son} \\ 
  m &= \text{height of mother}  \\ 
  f &= \text{height of father} 
\end{align} 
$$ {#eq-s-per-mf-2}

Each set of coefficient values determines some plane in the 3-dimensional space of (mother, father, son) heights.  The coefficients $(\hat{\beta}_0, \hat{\beta}_m, \hat{\beta}_f)$ obtained by linear regression determine the _regression plane_ (@fig-mfs-scat3d) that gives the best linear approximation $(\hat{s})$ to the son's height for a given pair of parent heights $(m, f)$. [^least-squares-approximation]

[^least-squares-approximation]: "Best" in the sense of minimizing the sum of squared residuals of actual minus predicted sons' heights.

```{r}
#| label: s-mf-lm

s_mf_lm <- lm(
  data    = galton_3d |> dplyr::filter(gender == "male"), 
  formula = child ~ mother + father)
```

```{r}
#| label: s-mf-tbl

# append fitted values (predictions)
s_mf_tbl <- galton_3d |> 
  dplyr::filter(gender == "male") |> 
  dplyr::rename(son = child) |> 
  dplyr::mutate(s_hat = s_mf_lm$ fitted.values)
```

```{r}
#| label: fig-mfs-scat3d
#| fig-cap: "Son's height given (mother, father) heights: predicted (plane) and observed (point)"

g_mfs_pt <- scatterplot3d(
  x = s_mf_tbl$ mother, xlab = "mother", 
  y = s_mf_tbl$ father, ylab = "father", 
  z = s_mf_tbl$ son,    zlab = "son", 
  pch = 16, 
  color = "steelblue",
  angle = 55, 
  main = "Family heights: mother, father, son", 
  sub  = "The regression plane contains the predicted heights of sons.")

# Add the regression plane ("g_mfs_plane")
s_mf_lm |> g_mfs_pt$ plane3d(
  lty          = "solid",
  lty.box      = "solid",
  draw_polygon = TRUE,
  col          = "grey")
```

In vector-matrix notation we are seeking a vector $(\hat{\beta}_0, \hat{\beta}_m, \hat{\beta}_f)$ of coefficient values that yields the least-squares solution to the following linear approximation problem.

$$
\begin{align} 
  s_\bullet &\approx (1_\bullet, m_\bullet, f_\bullet) \times 
    \begin{pmatrix} 
      \hat{\beta}_0 \\ \hat{\beta}_m \\ \hat{\beta}_f
    \end{pmatrix} 
\end{align} 
$$ {#eq-s-per-mf-3}

where 

$$
\begin{align} 
  s_\bullet &= \text{data column vector: heights of sons} \\ 
  1_\bullet &= \text{column vector } (1, \ldots, 1) \\ 
  m_\bullet &= \text{data column vector: heights of mothers}  \\ 
  f_\bullet &= \text{data column vector: heights of fathers} 
\end{align} 
$$ {#eq-s-per-mf-4}

This is a statistical estimation problem that corresponds to the following linear algebra problem and notation.

$$
\begin{align} 
  b_\bullet &\approx A_{\bullet, \bullet} \times x_\bullet
\end{align} 
$$ {#eq-b-Ax}

where 

$$
\begin{align} 
  b_\bullet &= s_\bullet \\ 
  A_{\bullet, \bullet} &= (1_\bullet, m_\bullet, f_\bullet) \\ 
  x_\bullet &= (\hat{\beta}_0, \hat{\beta}_m, \hat{\beta}_f) 
\end{align} 
$$ {#eq-b-Ax-2}

It turns out that the least squares solution $(\hat{\beta}_0, \hat{\beta}_m, \hat{\beta}_f)$ can be obtained as the vector of coefficients of an orthogonal projection of vector $s_\bullet$ onto the 3-dimensional subspace spanned by vectors $(1_\bullet, m_\bullet, f_\bullet)$.  More on this later.

### Survey Data: Better Life Index

```{r}
#| label: bli-long
bli_long <- get_bli_long()
```

```{r}
#| label: bli-wide

bli_wide <- assert_bli_wide(bli_long)

# record constants to be used in narrative
n_bli_loc   <- nrow(bli_wide)
n_bli_comps <- ncol(bli_wide) - 2L
```

We now turn to a data set having several data columns, namely the OECD's Better Life Index (BLI). [^OECD-about] The following table shows a portion of the data.

[^OECD-about]: The OECD (Organisation for Economic Co-operation and Development) works with 100+ countries to collect and analyze data in order to promote public policy.  The OECD's 38 Member countries span the world, from North America and South America to Europe and Asia-Pacific.

```{r}
#| label: tbl-bli-wide
#| tbl-cap: "Better Life Index (BLI)"

bli_wide |> print(n = 6)
```

Each row of this data matrix gives specified measurements of an identified country.  The first two columns give, respectively, each country's OECD code and name.  The remaining `r n_bli_comps` columns are measures pertaining to the well-being of the populace.

The column name of each measures consists of a two-letter prefix followed by a suffix.  The prefix is associated with a broad indicator of social well-being.  The suffix pertains to a particular component of this indicator.  Here is an expansion of these prefixes.

```{r}
#| label: bli-comp-prefix
bli_comp_prefix <- assert_bli_comp_prefix()
```

```{r}
#| label: tbl-bli-comp-prefix
#| tbl-cap: "BLI Indicators and Sub-Components"

bli_comp_prefix |> 
  dplyr::rename(components = comps) |> 
  knitr::kable(
    caption = "BLI Indicators and Sub-Components")
```

The component indicators (corresponding to the suffix of the column name) are elaborated in the following table.

```{r}
#| label: bli-components
bli_components <- assert_bli_component_indicators()
```

```{r}
#| label: tbl-bli-components
#| tbl-cap: "BLI Component Indicators"

bli_components |> 
  dplyr::select(prefix, suffix, unit, post_name, dscr) |> 
  dplyr::rename(
    name = post_name, 
    description = dscr
  ) |> 
  knitr::kable(
    caption = "BLI Component Indicators")
```

The `unit` column in the above table gives the unit of measure, with `PC` meaning percent, `YR` meaning number of years, and so on.

We now turn to a statistical and algebraic treatment of the BLI data matrix of @tbl-bli-wide.  Consider the indicator component `SW_LIFS` (Life Satisfaction) as a response variable, with the remaining `r n_bli_comps -1L` indicator components serving as explanatory variables.  As with the previous data example, we want to approximate or predict the response variable by a constant $\beta_0$ plus a linear combination of the explantory variables, as follows.

$$
\begin{align} 
  L_\bullet &\approx (1_\bullet, C_{1, \bullet}, \ldots, C_{d, \bullet}) \times 
    \begin{pmatrix} 
      \hat{\beta}_0 \\ \hat{\beta}_1 \\ \vdots \\ \hat{\beta}_d
    \end{pmatrix} 
\end{align} 
$$ {#eq-L-per-C}

where 

$$
\begin{align} 
  L_\bullet &= \text{life satisfaction indicator per country} \\ 
  1_\bullet &= \text{column vector } (1, \ldots, 1) \\ 
  C_{k, \bullet} &= k^{th} \text{ indicator component per country} \\ 
  d &= \text{number of explanatory indicators}
\end{align} 
$$ {#eq-L-per-C-2}

We now have more explanatory variables than in the previous example, a fact that merits some comment.

On the one hand, the approach to determining least-squares regression coefficients $\hat{\beta}_0, \ldots, \hat{\beta}_d$ is unchanged.  We project the response vector, now $L_\bullet$, onto the space spanned by the constant vector $1_\bullet$ along with the explanatory variables, that is onto the space spanned by $(1_\bullet, C_{1, \bullet}, \ldots, C_{d, \bullet})$.  The fitted coefficients yield a function of the explanatory variables that forms a regression hyperplane of dimension `r n_bli_comps - 1L` that passes through a cloud of data points, $(C_{1, \bullet}, \ldots, C_{d, \bullet}, L_\bullet)$, in a space of dimension `r n_bli_comps`.

On the other hand, we are now estimating `r n_bli_comps` regression coefficients based on observations from just `r n_bli_loc` countries.  From a statistical perspective, this paucity of observations relative to the number of estimates leads to large standard errors for the set of estimated coefficients.  From the perspective of numerical linear algebra, the vector of fitted coefficients $(\hat{\beta}_0, \ldots, \hat{\beta}_d)$ is less stable (more sensitive to error in the data) than it was in the previous example.

### MNIST: Images of Handwritten Digits

The MNIST database (Modified National Institute of Standards and Technology database) is a large database of handwritten decimal digits consisting of 60,000 training images and 10,000 testing images. [^MNIST-refs]

[^MNIST-refs]: See @LeCun_Cortes_Burges_2005 and @wiki_MNIST.

The history of this database goes back to 1988, when the US Postal Service constructed images of digits appearing on handwritten zip codes.  Around the same time the US Census Bureau requested NIST to evaluate optical character recognition (OCR) systems.  In 1992, NIST and the Census Bureau sponsored a competition in which participating teams were given images of Handwriting Sample Forms (HSFs), including handwritten decimal digits.  The initial version of MNIST was constructed sometime before summer 1994.

```{r}
#| label: read-xmpl-train
# Read a tibble (created in another project) that represents 
# a minimal subset of one (training set) image per distinct label.

xmpl_train_image_tbl <- read_tsv(here::here(
  "data", "mnist_xmpl_per_digit", "xmpl_train_image_tbl.txt"
))

```

Here's an example of each handwritten digit from the training set of images.

```{r}
#| label: fig-xmpl-train
#| fig-cap: "Example images of handwritten digits from the MNIST dataset"

g_xampl_train <- xmpl_train_image_tbl |> 
  arrange(label, img_dx, col_dx, row_dx) |> 
  ggplot2::ggplot(aes(x = col_dx, y = row_dx, fill = pixel)) +
  geom_raster() +  
  scale_fill_gradient(low = "white", high = "black") +
  scale_y_reverse() +
  facet_wrap(~ label, nrow = 2, ncol = 5,
             labeller = labeller(label = \(x) paste("Digit:", x))) +
  coord_equal() +
  theme_void() +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    legend.position = "none",
    panel.spacing = unit(0.5, "lines")
  )
g_xampl_train

```

Each image is represented by a $28 \times 28$ matrix of pixels, with each pixel represented as a grayscale integer value from 0 through 255.  That is, each image represents a single vector in a space of dimension 784 (since $28 \times 28 = 784$).

The 1992 competition prompted the development of algorithms to determine the decimal digit represented by any such image.  This is a classification problem: to label each case of data (image) as belonging to one of several possible categories (decimal digits).

One such method, multinomial logistic regression, assigns a probability that a given image represents a specified digit, resulting in a 10-element probability vector per image. [^wiki-multinomial-regression]

[^wiki-multinomial-regression]: See @wiki_multinomial_regression.

#### Multinomial Logistic Regression

To formulate the model, we convert the representation of an image from a $28 \times 28$ matrix of pixels into a vector of pixels of length 784. [^row-major-format]  We'll denote such a vector as $(P_1, \ldots, P_d)$, where $d = 784$.

[^row-major-format]: The conversion of a matrix of pixels to a vector of pixels is known as raster-to-vector (R2V) conversion, usually in row-major format, whereby the elements of the vector are taken from the top row and then from each succeeding row.  See @wiki_raster.

Let $D$ denote the digit represented by the image.  The ordering of the digits from 0 through 9 is not directly relevant to the image-recognition problem, so let us regard $D$ as a categorical variable having the set $\{ 0, 1, \ldots, 9 \}$ as possible values.  An alternative representation is the set of indicator vectors $e_0 = (1, 0, \ldots, 0)$ through $e_9 = (0, 0, \ldots, 1)$, called "one-hot encoding" in machine learning. [^one-hot]

[^one-hot]: See @wiki_one-hot and @wiki_categorical_variable.

Then the multinomial logistic regression model can be formulated as follows.

$$
\begin{align} 
  \log_e{ \frac{P(D = \nu)}{P(D = 0)} } &= (1, P_1, \ldots, P_d) \times 
    \begin{pmatrix} 
      \beta_0^{(\nu)} \\ \beta_1^{(\nu)} \\ \vdots \\ \beta_d^{(\nu)}
    \end{pmatrix} & \text{ for } \nu \in \{ 1, \ldots, 9 \}
\end{align} 
$$ {#eq-log-ratio-per-image}

with 

$$
\begin{align} 
  P(D = 0) &= 1 - \sum_{\nu = 1}^9 P(D = \nu)
\end{align} 
$$ {#eq-log-ratio-per-image-2}

For a more compact notation let $X_{\bullet} = (1, P_1, \ldots, P_d)$ and let $\beta_{\bullet}^{(\nu)} = (\beta_0^{(\nu)}, \beta_1^{(\nu)}, \ldots, \beta_d^{(\nu)})$, with the inner product [^inner-product] of these two vectors denoted as $X_{\bullet} \boldsymbol\cdot \beta_{\bullet}^{(\nu)}$.  Then we have 

[^inner-product]: The inner product of vectors $x, y \in \mathcal{V}$ has alternative notations, including $x \boldsymbol\cdot y$, $\left < x, y \right >$, and $x^\top y$.

$$
\begin{align} 
  \log_e{ \frac{P(D = \nu)}{P(D = 0)} } &= X_{\bullet} \boldsymbol\cdot \beta_{\bullet}^{(\nu)} & \text{ for } \nu \in \{ 1, \ldots, 9 \}
\end{align} 
$$ {#eq-log-ratio-per-image-3}

Exponentiation of @eq-log-ratio-per-image-3 gives:

$$
\begin{align} 
  \{ P(D = \nu) \} &= \{ P(D = 0) \} \times e^{X_{\bullet} \boldsymbol\cdot \beta_{\bullet}^{(\nu)}} & \text{ for } \nu \in \{ 1, \ldots, 9 \}
\end{align} 
$$ {#eq-log-ratio-per-image-4}

Taking the sum over $\nu$ we have: 

$$
\begin{align} 
  \sum_{\nu = 1}^9 {P(D = \nu)} &= \{ P(D = 0) \} \times \sum_{\nu = 1}^9 e^{X_{\bullet} \boldsymbol\cdot \beta_{\bullet}^{(\nu)}}
\end{align} 
$$ {#eq-log-ratio-per-image-5}

Now applying @eq-log-ratio-per-image-2 we have 

$$
\begin{align} 
  \left \{ 1 - P(D = 0) \right \} &= \{ P(D = 0) \} \times \sum_{\nu = 1}^9 e^{X_{\bullet} \boldsymbol\cdot \beta_{\bullet}^{(\nu)}}
\end{align} 
$$ {#eq-log-ratio-per-image-6}

which yields: 

$$
\begin{align} 
  P(D = 0) &= \frac{1} { 1 + \sum_{\nu = 1}^9 e^{X_{\bullet} \boldsymbol\cdot \beta_{\bullet}^{(\nu)}} }
\end{align} 
$$ {#eq-log-ratio-per-image-7}

Applying @eq-log-ratio-per-image-4 gives: 

$$
\begin{align} 
  P(D = \nu) &= \frac{ e^{X_{\bullet} \boldsymbol\cdot \beta_{\bullet}^{(\nu)}} } { 1 +  \sum_{\mu = 1}^9 e^{X_{\bullet} \boldsymbol\cdot \beta_{\bullet}^{(\mu)}} } & \text{ for } \nu \in \{ 1, \ldots, 9 \}
\end{align} 
$$ {#eq-log-ratio-per-image-8}

#### Matrix Representation

@eq-log-ratio-per-image-3 pertains to the probability that a single image represents a single digit $\nu \in \{1, \ldots, 9 \}$.  Therefore, in a data set of $n$ images, with $i$ denoting the index of a particular image, we have:

$$
\begin{align} 
  \log_e{ \frac{P(D_i = \nu)}{P(D_i = 0)} } &= X_{i, \bullet} \boldsymbol\cdot \beta_{\bullet}^{(\nu)} 
\end{align} 
$$ {#eq-log-ratio-per-image-9}

Expanding the last equation to matrix notation, with $i$ as the row index and $\nu$ as a column index, we have

$$
\begin{align} 
&
\begin{pmatrix} 
  \log_e{ \frac{P(D_1 = 1)}{P(D_1 = 0)} }, & \ldots, & \log_e{ \frac{P(D_1 = 9)}{P(D_1 = 0)} } \\ 
  \vdots & \vdots & \vdots \\
  \log_e{ \frac{P(D_n = 1)}{P(D_n = 0)} }, & \ldots, & \log_e{ \frac{P(D_n = 9)}{P(D_n = 0)} }
\end{pmatrix}  \\ \\ 
&= 
\begin{pmatrix} 
  X_{1, \bullet} \\ 
  \vdots \\
  X_{n, \bullet}
\end{pmatrix}
\begin{pmatrix} 
  \beta_{\bullet}^{(1)}, & \ldots, & \beta_{\bullet}^{(9)}
\end{pmatrix}
\end{align} 
$$ {#eq-log-ratio-per-image-10}

The matrix on the left side of @eq-log-ratio-per-image-10 has dimensions $n \times 9$.  On the right side, the first matrix factor has dimensions $n \times 785$, and the second matrix factor has dimensions $785 \times 9$.

## Notation

The preceding section introduced example data sets along with corresponding linear regression models of the following form.

$$
\begin{align} 
  y &= X \; \beta \; + \; \epsilon
\end{align} 
$$ {#eq-generic-lm}

Each of the elements of @eq-generic-lm has alternative names, including the following. [^feature-matrix]

[^feature-matrix]: Matrix $X$ on the right side of @eq-generic-lm is called a _feature matrix_ that may contain original data columns (other than the response or labeling variable) and may also contain columns that are functions of the data or of other information.  A data matrix is model-agnostic, whereas a feature matrix is constructed to support a model of some form.

$$
\begin{align} 
  y &= \text{a } \textit{response, target,} \text{ or } \textit{labeling}  \text{ variable} \\ 
  X &= \text{feature matrix of } \textit{explanatory, predictor,} \text{ or } \textit{feature}  \text{ variables} \\ 
  \beta &= \text{a vector of model } \textit{coefficients} \text{ or } \textit{parameters} \\ 
  \epsilon &= \text{an } \textit{error} \text{ or } \textit{residual} \text{ term}
\end{align} 
$$ {#eq-generic-lm-terms}

This linear regression format follows the more general mathematical notation $y = f(x)$.  In data science and machine learning, however, the response variable $y$ and the feature matrix $X$ have known values, whereas $\beta$ and $\epsilon$ are _fit_ (determined or evaluated based on $y$ and $X$) over the course of the modeling process.

In the data examples of the preceding section, the response variable took the following form.

  - Family heights: $y =$ oldest child's height
  - Better Life Index: $y =$ the Life Satisfaction indicator
  - MNIST: $y =$ a probability vector $\{ P(D = \nu) \}_{\nu = 0}^9$ assigned to each image

The MNIST example illustrates a _vector-valued_ rather than _scalar-valued_ response variable.

If the data include a labeling or response variable, $y$, then the problem is said to be _supervised_.  In _unsupervised_ problems (that lack a $y$ variable), we may need to find patterns in the given data.  For example we may seek those feature variables (columns of the feature matrix $X$), or linear combinations of feature variables, that account for most of the variability in the entire set of feature variables.  Or we may need to find observations (rows of the feature matrix $X$) that are similar and thereby form groups (or _clusters_) of observations.  In these unsupervised situations we may model the feature matrix (or its covariance matrix) as the product of other matrices of special form (to be discussed later in this chapter).

In the remainder of this chapter we will focus on ideas and methods that help us to solve @eq-generic-lm, or rather, that help us to determine the value of $\beta$ that minimizes (in some sense) the residual term $\epsilon$.  We refer to this minimization as the _linear regression problem_, which is made precise once we specify the measure of $\epsilon$ to be minimized.

## Geometry

We now consider the geometry of the _least-squares_ solution of the linear regression problem, using the example of family heights.  We begin by defining this measure of the residual term $\epsilon$.

### Distance Measures

The sum-of-squares measure of the residual vector $\epsilon_\bullet = (\epsilon_1, \ldots, \epsilon_n)$ is simply the sum of the squares of the components $\epsilon_\nu$.

$$
\begin{align} 
  \sum_{\nu = 1}^n | \epsilon_\nu |^2
\end{align} 
$$ {#eq-eps-ss}

#### Vector norms

@eq-eps-ss defines the following _norm_ on $n-$dimensional Euclidean space.  For any vector $v_\bullet = (v_1, \ldots, v_n) \in \mathbb{R}^n$ we define $\Vert v_\bullet \rVert_2$ as follows.

$$
\begin{align} 
  \Vert v_\bullet \rVert_2 
  &= \left ( \sum_{\nu = 1}^n | v_\nu |^2 \right )^{\frac{1}{2}}
\end{align} 
$$ {#eq-l2-norm-n}

This norm can be derived from (or used to define) the _inner-product_ of a pair of vectors $v_\bullet, w_\bullet \in \mathbb{R}^n$, defined as follows with the following alternative notations.

$$
\begin{align} 
  \left <  v_\bullet, w_\bullet \right >
  &= v_\bullet \boldsymbol\cdot w_\bullet \\ 
  &= v_\bullet^\top w_\bullet \\ 
  &= \sum_{\nu = 1}^n v_\nu \; w_\nu 
\end{align} 
$$ {#eq-inner-product-R-n}

More generally, for any real number $p \ge 1$, the so-called $p-$norm (or Minkowski norm of order $p$) is defined as 

$$
\begin{align} 
  \Vert v_\bullet \rVert_p 
  &= \left ( \sum_{\nu = 1}^n | v_\nu |^p \right )^{\frac{1}{p}} & \text{ for } 1 \le p < \infty
\end{align} 
$$ {#eq-lp-norm-n}

This definition can be extended to the case $p = \infty$ as follows.

$$
\begin{align} 
  \Vert v_\bullet \rVert_\infty 
  &= \max \left \{ |v_\nu | \right \}_{\nu = 1}^n
\end{align} 
$$ {#eq-max-norm-n}

More generally, for $v, w \in \mathcal{V}$, a real-valued or complex-valued vector space, a norm $\Vert \cdot \rVert$ is defined to have the following properties.

$$
\begin{align} 
  \Vert v \rVert &\ge 0 \\ 
  \Vert v \rVert &= 0 & \text{ if and only if } v = 0 \\ 
  \Vert v + w \rVert &\le \Vert v \rVert + \Vert w \rVert \\ 
  \Vert \lambda \; v \rVert &= | \lambda | \; \Vert v \rVert & \text{ for any scalar } \lambda
\end{align} 
$$ {#eq-norm-properties}

#### Matrix norms

If $\lVert v_\bullet \rVert$ denotes some defined norm for vectors $v_\bullet \in \mathbb{R}^n$ and if $M_{\bullet, \bullet}$ is an $n \times n$ numeric matrix, then the vector norm defines a corresponding matrix norm $\lVert M_{\bullet, \bullet} \rVert$ as follows.

$$
\begin{align} 
  \lVert M_{\bullet, \bullet} \rVert &= 
  \sup_{ \lVert v_\bullet \rVert = 1 } 
  \left \{ \lVert  M_{\bullet, \bullet} \; v_\bullet \rVert \right \}
\end{align} 
$$ {#eq-m-norm-from-v-norm}

Note that some matrix norms are defined otherwise.  For example, the Frobenius norm, $\lVert M_{\bullet, \bullet} \rVert_F$, is defined as $\lVert vec(M_{\bullet, \bullet}) \rVert_2$, where $vec(M_{\bullet, \bullet})$ converts an $n \times n$ matrix into a vector of length $n^2$ by concatenating matrix columns.

#### Metric spaces

A vector norm $\Vert \cdot \rVert$ defines a corresponding distance measure $\delta_{\Vert \cdot \rVert}$

$$
\begin{align} 
  \delta_{\Vert \cdot \rVert} (v_\bullet, w_\bullet) 
  &= \lVert v_\bullet - w_\bullet \rVert
\end{align} 
$$ {#eq-norm-metric}

A general distance measure or metric, $\delta (\cdot, \cdot)$, together with the set of points $\mathcal{M}$ over which it is defined constitutes a _metric space_ with the following properties, for any $m_1, m_2, m_3 \in \mathcal{M}$.

$$
\begin{align} 
  \delta (m_1, m_1) &= 0 \\ 
  \delta (m_1, m_2) &> 0 & \text{ whenever } m_1 \ne m_2 \\ 
  \delta (m_1, m_2) &= \delta (m_2, m_1) \\ 
  \delta (m_1, m_3) &\le \delta (m_1, m_2) \; + \; \delta (m_2, m_3) 
\end{align} 
$$ {#eq-metric-properties}

Hamming distance and Levenshtein distance are important examples of metrics used in natural language processing (NLP) that are not based on a vector norm. [^NLP-metrics]

[^NLP-metrics]: See @wiki_Hamming_distance and @wiki_Levenshtein_distance.

### Family heights

Here's the linear regression problem in matrix format.

$$
\begin{align} 
  y_\bullet &= X_{_\bullet, _\bullet} \; \beta_\bullet \; + \; \epsilon_\bullet
\end{align} 
$$ {#eq-generic-lm-mat}

where 

$$
\begin{align} 
  y_\bullet &= \text{heights of sons } = s_\bullet \\ 
  m_\bullet &= \text{heights of mothers } \\ 
  f_\bullet &= \text{heights of fathers } \\ 
  X_{_\bullet, _\bullet} &= \text{feature matrix} = (1_\bullet, m_\bullet, f_\bullet) \\ 
  \beta_\bullet &= \text{coefficient vector} = (\beta_0, \beta_1, \beta_2)^\top \\ 
  \epsilon_\bullet &= \text{residual vector}
\end{align} 
$$ {#eq-generic-lm-2}

```{r}
#| label: smf-smpl

n_smf        <- nrow(s_mf_tbl)
n_smf_smpl   <- 20

set.seed(37)
smf_smpl_idx <- 
  sample.int(n = n_smf, size = n_smf_smpl) |> 
  sort()
smf_smpl_tbl <- 
  s_mf_tbl [smf_smpl_idx, ] |> 
  dplyr::select(- s_hat)

```

```{r}
#| label: smf-smpl-lm

# fit linear model to sampled data
smf_smpl_lm <- lm(
  data    = smf_smpl_tbl, 
  formula = son ~ mother + father
)
```

```{r}
#| label: smf-smpl-append

# append columns: (fitted values, residuals)
smf_smpl_tbl <- smf_smpl_tbl |> 
  dplyr::mutate(
    s_hat = smf_smpl_lm$ fitted.values, 
    resid = smf_smpl_lm$ residuals
  )
```

Consider the least-squares estimate $\hat{\beta}_\bullet$ and the consequent predicted height $\hat{y}_\bullet = X_{_\bullet, _\bullet} \hat{\beta}_\bullet$ of the son.  @fig-smf-smpl is based on a random sample of `r n_smf_smpl` families and shows the heights of sons on the vertical axis, along with their vertical displacement (residual) from the predicted value lying on the _regression plane_. [^son_resid]

[^son_resid]: The residual son's height is the error term in the regression formula.  In the figure, residuals are color-coded according to their sign: black if positive and red otherwise.

```{r}
#| label: fig-smf-smpl
#| fig-cap: "Sampled son heights: residual = observed - predicted"

g_smf_smpl <- scatterplot3d(
  x = smf_smpl_tbl$ mother, xlab = "mother", 
  y = smf_smpl_tbl$ father, ylab = "father", 
  z = smf_smpl_tbl$ son,    zlab = "son", 
  pch = 16, 
  color = "steelblue",
  angle = 55, 
  main = "Sampled family heights: mother, father, son", 
  sub  = "son residual = observed - predicted")

# Add the regression plane
s_mf_lm |> g_smf_smpl$ plane3d(
  lty          = "solid",
  lty.box      = "solid",
  draw_polygon = TRUE,
  col          = "grey")

# Add line segments from predicted to actual values
# The key is using g_smf_smpl$xyz.convert() to get the 2D coordinates
for(i in 1:nrow(smf_smpl_tbl)) {
    m     <- smf_smpl_tbl$ mother [i]
    f     <- smf_smpl_tbl$ father [i]
    s     <- smf_smpl_tbl$ son    [i]
    s_hat <- smf_smpl_tbl$ s_hat  [i]
    resid <- smf_smpl_tbl$ resid  [i]
    r_clr <- ifelse(resid > 0, "black", "red")
    
  g_smf_smpl$ points3d(
    c(m, m),
    c(f, f),
    c(s_hat, s),
    type = "l", col = r_clr, lwd = 1
  )
}

```

@fig-smf-smpl represents individual rows of data, $\{ (m_i, f_i, s_i) \}_{i = 1}^n$ along with model predictions $(\hat{s} = \hat{\beta}_0 + \hat{\beta}_1 m + \hat{\beta}_2 f)$ and residuals $(\hat{\epsilon} = s - \hat{s})$ in three-dimensional $(m, f, s)$ space.

To gain more insight into linear regression we'll first reduce the regression problem to the simple case in which the response variable and the predictor variables have all been coerced to have an average value of zero, a process called _centering_.  This will eliminate the need for the intercept coefficient, $\beta_0$, and consequently eliminate the need to include the constant vector $1_\bullet$ in the feature matrix $X_{_\bullet, _\bullet}$.

### Centering Data Vectors

The regression plane that we glimpse in @fig-smf-smpl actually spans all the $(m, f)$ combinations that are mathematically possible.  If we imagine infinitesimally short parents with $(m, f) = (0, 0)$, the predicted height of their son would be $\hat{\beta}_0$, which is not zero.  That is, the plane does not pass through the origin $(0, 0, 0)$ and therefore does not qualify as a subspace of $(m, f, s)$ space. [^lin-manifold] But the regression plane determines a parallel subspace (that _does_ pass through the origin).

[^lin-manifold]:  The regression plane qualifies as a linear manifold, mathematically speaking.

The concept of a subspace is central to linear algebra.  Therefore determining the subspace parallel to the regression plane will enable us to apply linear algebra methods to better understand linear regression.

One way to generate this subspace is to center each of the $(m_i, f_i, s_i)$ data values, that is, to replace data value $v_i$ with its centered version $\dot{v}_i = v_i - \bar{v}$, where $\bar{v}$ denotes the average value (arithmetic mean) of vector $v_\bullet$.

In vector-matrix notation we have 

$$
\begin{align} 
  \bar{v} &= \frac{1}{n} \sum_{\nu = 1}^n v_\nu \\ 
  &= \frac{1}{n} \;  1_\bullet^\top \; v_\bullet \\ \\ 
  & \text{so that} \\ \\ 
  \dot{v}_\bullet &= v_\bullet - ( \bar{v} \; 1_\bullet ) \\ 
  &= v_\bullet - \frac{1}{n} \;  1_\bullet \;  1_\bullet^\top \; v_\bullet \\ 
  &= \left ( I - \frac{1}{n} \;  1_\bullet \;  1_\bullet^\top \right ) \; v_\bullet
\end{align} 
$$ {#eq-avg-via-inner-product}

Let $C_{\bullet, \bullet}$ denote the matrix factor on the right side of the last equality, and define vector $\tilde{1}_\bullet$ as follows.

$$
\begin{align} 
  \tilde{1}_\bullet &= \frac{1}{\sqrt{n}} \;  1_\bullet \\ \\ 
  & \text{so that} \\ \\ 
  \lVert \tilde{1}_\bullet \rVert &= 1 \\ \\ 
  & \text{and} \\ \\ 
  C_{\bullet, \bullet} &= I \; - \; \tilde{1}_\bullet \; \tilde{1}_\bullet^\top
\end{align} 
$$ {#eq-centering-matrix}

Then we have 

$$
\begin{align} 
  \tilde{1}_\bullet \; \tilde{1}_\bullet^\top \;  v_\bullet 
  &= \bar{v} \; 1_\bullet \\ \\ 
  C_{\bullet, \bullet} \; v_\bullet &= v_\bullet \; - \; \bar{v} \; 1_\bullet \\ 
  &= \dot{v}_\bullet 
\end{align} 
$$ {#eq-centering-projection}

Setting $v_\bullet = 1_\bullet$ gives 

$$
\begin{align} 
  \left ( \tilde{1}_\bullet \; \tilde{1}_\bullet^\top \right ) \;  1_\bullet 
  &= 1_\bullet \\ \\ 
  C_{\bullet, \bullet} \; 1_\bullet &= 0_\bullet 
\end{align} 
$$ {#eq-centering-eigenvector}

We now multiply both sides of @eq-generic-lm-mat, the regression equation, by matrix $C_{\bullet, \bullet}$ to obtain 

$$
\begin{align} 
  C_{\bullet, \bullet} \; y_\bullet 
  &= C_{\bullet, \bullet} \; X_{\bullet, \bullet} \; \beta_\bullet 
  \; + \; C_{\bullet, \bullet} \; \epsilon_\bullet 
\end{align} 
$$ {#eq-centering-generic-lm}

Now 

$$
\begin{align} 
  C_{\bullet, \bullet} \; X_{\bullet, \bullet} \; \beta_\bullet 
  &= C_{\bullet, \bullet} \; (1_\bullet, m_\bullet, f_\bullet) \; 
  \begin{pmatrix}
    \beta_0 \\ 
    \beta_1 \\ 
    \beta_2
  \end{pmatrix} \\ 
  &= (0_\bullet, \dot{m}_\bullet, \dot{f}_\bullet) \; 
  \begin{pmatrix}
    \beta_0 \\ 
    \beta_1 \\ 
    \beta_2
  \end{pmatrix} \\ 
  &= \beta_1 \; \dot{m}_\bullet \; + \; \beta_2 \; \dot{f}_\bullet \\ 
  &= (\dot{m}_\bullet, \dot{f}_\bullet) \; 
  \begin{pmatrix}
    \beta_1 \\ 
    \beta_2
  \end{pmatrix} 
\end{align} 
$$ {#eq-centering-sfm-lm}

Then the centered version of @eq-generic-lm-mat is 

$$
\begin{align} 
  \dot{y}_\bullet &= \dot{X}_{_\bullet, _\bullet} \; \beta_\bullet \; + \; \dot{\epsilon}_\bullet
\end{align} 
$$ {#eq-generic-lm-ctr}

where 

$$
\begin{align} 
  \dot{y}_\bullet &= \text{centered heights of sons } = \dot{s}_\bullet \\ 
  \dot{m}_\bullet &= \text{centered heights of mothers } \\ 
  \dot{f}_\bullet &= \text{centered heights of fathers } \\ 
  \dot{X}_{_\bullet, _\bullet} &= \text{centered feature matrix} = (\dot{m}_\bullet, \dot{f}_\bullet) \\ 
  \beta_\bullet &= \text{coefficient vector} = (\beta_1, \beta_2)^\top \\ 
  \dot{\epsilon}_\bullet &= \text{centered residual vector}
\end{align} 
$$ {#eq-generic-lm-ctr-2}

That is, we can eliminate the intercept coefficient from the centered linear model, and we can also eliminate the constant vector $1_\bullet$ from the feature matrix $X_{_\bullet, _\bullet}$.  @fig-smf-smpl-ctr is a version of @fig-smf-smpl corresponding to @eq-generic-lm-ctr.  Geometrically it's the same figure, the difference being that each of the three axes has been shifted, now with 0 as the central value.

```{r}
#| label: smf-smpl-ctr

# center data variables using base::scale()
smf_smpl_ctr <- smf_smpl_tbl |> 
  dplyr::select(- c(s_hat, resid)) |> 
  dplyr::mutate(dplyr::across(
    .cols = c(mother, father, son), 
    .fns  = scale, scale = FALSE
  )) |> 
  # change scale() output from matrix to vector
  dplyr::mutate(dplyr::across(
    .cols = c(mother, father, son), 
    .fns  = as.vector
  ))
```

```{r}
#| label: smf-smpl-ctr-lm

# fit linear model to centered variables
smf_smpl_ctr_lm <- lm(
  data    = smf_smpl_ctr, 
  formula = son ~ 0 + mother + father)
```

```{r}
#| label: smf-smpl-ctr-append

# append (fitted values, residuals)
smf_smpl_ctr <- smf_smpl_ctr |> 
  dplyr::mutate(
    s_hat = smf_smpl_ctr_lm$ fitted.values, 
    resid = smf_smpl_ctr_lm$ residuals
  )
```

```{r}
#| label: fig-smf-smpl-ctr
#| fig-cap: "Centered family heights"

g_smf_smpl_ctr <- scatterplot3d(
  x = smf_smpl_ctr$ mother, xlab = "mother", 
  y = smf_smpl_ctr$ father, ylab = "father", 
  z = smf_smpl_ctr$ son,    zlab = "son", 
  pch = 16, 
  color = "steelblue",
  angle = 55, 
  main = "Centered family heights: mother, father, son", 
  sub  = "son residual = observed - predicted")

# Add the regression plane
smf_smpl_ctr_lm |> g_smf_smpl_ctr$ plane3d(
  lty          = "solid",
  lty.box      = "solid",
  draw_polygon = TRUE,
  col          = "grey")

# Add line segments from predicted to actual values
# The key is using g_smf_smpl_ctr$xyz.convert() to get the 2D coordinates
for(i in 1:nrow(smf_smpl_ctr)) {
    m     <- smf_smpl_ctr$ mother [i]
    f     <- smf_smpl_ctr$ father [i]
    s     <- smf_smpl_ctr$ son    [i]
    s_hat <- smf_smpl_ctr$ s_hat  [i]
    resid <- smf_smpl_ctr$ resid  [i]
    r_clr <- ifelse(resid > 0, "black", "red")
    
  g_smf_smpl_ctr$ points3d(
    c(m, m),
    c(f, f),
    c(s_hat, s),
    type = "l", col = r_clr, lwd = 1
  )
}

```

### Least Squares Solutions

Having centered the data, let's discuss least-squares linear regression, which determines coefficient values $(\hat{\beta}_0, \hat{\beta}_1)$ that minimize the sum of squared residuals.

$$
\begin{align} 
  \sum_{i = 1}^n \epsilon_i^2 &= \lVert \epsilon_\bullet \rVert^2 \\ 
  &= \epsilon_\bullet^\top\epsilon_\bullet \\ 
  &= (\dot{y}_\bullet - \dot{X}_{\bullet, \bullet} \beta_\bullet)^\top (\dot{y}_\bullet - \dot{X}_{\bullet, \bullet} \beta_\bullet)
\end{align} 
$$ {#eq-resid-ss}

To find coefficient values that minimize this sum of squares, one can take derivatives of the above expression with respect to $\beta_\bullet$ and set that result to zero, which yields the following _normal equations_: 

$$
\begin{align} 
  \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet} \hat{\beta}_\bullet 
  &= \dot{X}_{\bullet, \bullet}^\top \dot{y}_\bullet
\end{align} 
$$ {#eq-lin-reg-nrml}

On the left side of the normal equations we have the matrix factor $\left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )$, which in our case is a multiple of the (mother, father) covariance matrix, a $2 \times 2$ non-negative-definite matrix. In fact the matrix is positive-definite, and thus invertible, provided only that parental heights are not perfectly correlated (which they are not).  Then we can invert this matrix to solve for $\hat{\beta}_\bullet$. 

$$
\begin{align} 
  \hat{\beta}_\bullet 
  &= \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top \dot{y}_\bullet
\end{align} 
$$ {#eq-lin-reg-beta-hat}

The predicted vector $\hat{\dot{y}}_\bullet$ is thus: 

$$
\begin{align} 
  \hat{\dot{y}}_\bullet 
  &= \dot{X}_{\bullet, \bullet} \hat{\beta}_\bullet \\ 
  &= \dot{X}_{\bullet, \bullet} \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top \dot{y}_\bullet
\end{align} 
$$ {#eq-lin-reg-y-hat}

### Orthogonal Projections

For any value of the coefficient vector $\beta_\bullet$, the mapping $\beta_\bullet \mapsto \dot{X}_{\bullet, \bullet} \beta_\bullet$ sends $\beta_\bullet$ to the linear combination $\beta_1 \; \dot{m}_\bullet \; + \; \beta_2 \; \dot{f}_\bullet$, which belongs to a 2-dimensional subspace of $n-$space [^n-space], where $n$ denotes the number of data cases, i.e., the row dimension of $\dot{X}_{\bullet, \bullet}$.  This 2-dimensional subspace (let's say the "parental" subspace) is spanned by the two $n-$vectors $(\dot{m}_\bullet, \dot{f}_\bullet)$, the centered vectors of (mother, father) heights.

[^n-space]: The terms $n-$dimensional space and $n-$space are shorthand for "an $n-$dimensional vector space $\mathcal{V}$ over the field $\mathbb{R}$ of real numbers".  Similarly, the term $n-$vector means a vector $v \in \mathcal{V}$.  Once a set of basis vectors is identified in $\mathcal{V}$, any $v \in \mathcal{V}$ can be identified by its coordinates with respect to the given basis.  An identified basis of $n$ vectors gives an isomorphism from $\mathcal{V}$ to $\mathbb{R}^n$.  Data are often collected and presented in the form of a data matrix, where a column of numeric values in $n$ successive rows provides an automatic representation as a vector in $\mathbb{R}^n$.  In the data set of family heights, for example, each row of data corresponds to a distinct family, and a column vector, the mother's height for example, is represented as $m_\bullet = (m_1, \ldots, m_n) \in \mathbb{R}^n$, where $m_i$ refers to the height in inches of the mother in family $i$.

The particular coefficient vector $\hat{\beta}_\bullet$ obtained by least squares linear regression produces the linear mapping of @eq-lin-reg-y-hat:

$$
\begin{align} 
  \hat{\dot{y}}_\bullet &= P \; \dot{y}_\bullet
\end{align} 
$$ {#eq-lin-reg-projection}

This mapping sends the centered heights of sons $\dot{y}_\bullet$ to their corresponding linear prediction $\hat{\dot{y}}_\bullet = P \; \dot{y}_\bullet$, where $P$ is the following matrix.

$$
\begin{align} 
  P
  &= \dot{X}_{\bullet, \bullet} \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top
\end{align} 
$$ {#eq-lin-reg-projection-mat}

Matrix $P$ is idempotent and symmetric, that is, both the square $P^2$ and the transpose $P^\top$ equal $P$.

$$
\begin{align} 
  P^2
  &= \left \{ \dot{X}_{\bullet, \bullet} \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top \right \} 
  \left \{ \dot{X}_{\bullet, \bullet} \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top \right \} \\ 
  &= \dot{X}_{\bullet, \bullet} \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top \\ 
  &= P \\ \\ 
  P^\top
  &= \left \{ \dot{X}_{\bullet, \bullet} \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top \right \}^\top \\ 
  &=  \dot{X}_{\bullet, \bullet} \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top \\ 
  &= P 
\end{align} 
$$ {#eq-projection-properties}

If matrix $M$ is idempotent, that is, if $M^2 = M$, then $M$ represents a _projection_.  Repeated applications of $M$ to vector $v$ return the initial application, i.e., $M^k v = Mv$ for any positive integer $k$.

If in addition matrix $M$ is symmetric, that is, if $M^\top = M$, then $M$ represents an _orthogonal projection_.  In this case the complement of $M$, $I - M$, also qualifies as an orthogonal projection and the product of the two matrices is the zero matrix.

Consequently, any vector $v$ can be expressed as the sum of two vectors $v = x + y$, with  $x = M v$ and $y = (I-M) v$.  Vector $x$ belongs to the subspace generated by $M$, which is called the _range_ of $M$, denoted as $\mathcal{R} (M)$.  Similarly $y \in \mathcal{R} (I - M)$.  Moreover, these two vectors are orthogonal: $(x^\top y = y^\top x = 0)$.  That is, subspace $\mathcal{R} (I - M)$ is the orthogonal complement of subspace $\mathcal{R} (M)$.

Let's apply these ideas to matrix $P$.  First, we have shown that matrix $P$ represents an orthogonal projection.  On closer inspection, we can show that the subspace generated by $P$, $\mathcal{R} (P)$, is the parental subspace.  That is, for any vector $v_\bullet$ we have:

$$
\begin{align} 
  P \; v_\bullet &= \dot{X}_{\bullet, \bullet} \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top \; v_\bullet \\ 
  &= \dot{X}_{\bullet, \bullet} \left \{ \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top \; v_\bullet \right \} \\ 
  &= \dot{X}_{\bullet, \bullet} \left \{ \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} 
  \begin{pmatrix}
    \dot{m}_{\bullet}^\top \; v_\bullet \\ 
    \dot{f}_{\bullet}^\top \; v_\bullet
  \end{pmatrix}
  \right \} \\ 
  &= \dot{X}_{\bullet, \bullet} \; \gamma_\bullet (v_\bullet)
\end{align} 
$$ {#eq-parental-projection}

That is, for any vector $v_\bullet$ in $n-$space, $P$ sends $v_\bullet$ to an $n-$vector of the form $\dot{X}_{\bullet, \bullet} \; \gamma_\bullet$, which is a linear combination of $(\dot{m}_\bullet, \dot{f}_\bullet)$ and thus belongs to the parental subspace.

In our example, this means that the respective vectors of predicted centered heights $\hat{\dot{y}_\bullet}$ and their residuals $\hat{\epsilon}_\bullet$ are mutually orthogonal.

$$
\begin{align} 
  \hat{\dot{y}}_\bullet &= P \; \dot{y}_\bullet \\ \\ 
  \hat{\epsilon}_\bullet &= \dot{y}_\bullet - \hat{\dot{y}}_\bullet \\ 
  &= (I - P) \; \dot{y}_\bullet \\ \\ 
  \hat{\epsilon}_\bullet^\top \; \dot{y}_\bullet &= \dot{y}_\bullet^\top \; (I - P)^\top P \; \dot{y}_\bullet \\ 
  &= \dot{y}_\bullet^\top \; (I - P) \; P \; \dot{y}_\bullet \\ 
  &= \dot{y}_\bullet^\top \; (P - P^2) \; \dot{y}_\bullet \\ 
  &= \dot{y}_\bullet^\top \; 0_{\bullet, \bullet} \; \dot{y}_\bullet \\ 
  &= 0
\end{align} 
$$ {#eq-resid-predicted-ortho}

Now let $p_\bullet$ be any vector in the parental space.  Then $p_\bullet$ is some linear combination of the parental vectors $(\dot{m}_\bullet, \dot{f}_\bullet)$ and therefore can be represented as $p_\bullet = \dot{X}_{\bullet, \bullet} \gamma_\bullet$ for some coefficient vector $\gamma_\bullet = (\gamma_1, \gamma_2)^\top$.  It now follows the $P \; p_\bullet = p_\bullet$: 

$$
\begin{align} 
  P \; p_\bullet &= P \; (\dot{X}_{\bullet, \bullet} \gamma_\bullet) \\ 
  &= \dot{X}_{\bullet, \bullet} \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top \;  (\dot{X}_{\bullet, \bullet} \gamma_\bullet) \\ 
  &= \dot{X}_{\bullet, \bullet} \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \left ( \dot{X}_{\bullet, \bullet}^\top \;  \dot{X}_{\bullet, \bullet} \right ) \gamma_\bullet \\ 
  &= \dot{X}_{\bullet, \bullet} \gamma_\bullet \\ 
  &= p_\bullet
\end{align} 
$$ {#eq-subspace-invariance}

Consequently, the residual vector $\hat{\epsilon}_\bullet$ is orthogonal to any vector $p_\bullet = \dot{X}_{\bullet, \bullet} \gamma_\bullet$ in the parental subspace: 

$$
\begin{align} 
  \hat{\epsilon}_\bullet^\top \; p_\bullet &= \dot{y}_\bullet^\top \; (I - P)^\top p_\bullet \\ 
  &= \dot{y}_\bullet^\top \; (I - P) \; p_\bullet \\ 
  &= \dot{y}_\bullet^\top \; 0_\bullet \\ 
  &= 0
\end{align} 
$$ {#eq-resid-mf-ortho}

It now follows that of all vectors $p_\bullet = \dot{X}_{\bullet, \bullet} \gamma_\bullet$ in the parental subspace, the predicted vector $\hat{\dot{y}}_\bullet$ is closest to the given vector $\dot{y}_\bullet$:

$$
\begin{align} 
  \lVert \dot{y_\bullet} - p_\bullet \rVert^2 
  &= \lVert (\dot{y_\bullet} - \hat{\dot{y}}_\bullet) + (\hat{\dot{y}}_\bullet - p_\bullet) \rVert^2  \\ 
  &= \lVert \hat{\epsilon}_\bullet + (\hat{\dot{y}}_\bullet - p_\bullet) \rVert^2 \\ 
  &= \left ( \hat{\epsilon}_\bullet + (\hat{\dot{y}}_\bullet - p_\bullet) \right )^\top 
  \left ( \hat{\epsilon}_\bullet + (\hat{\dot{y}}_\bullet - p_\bullet) \right ) \\
  &= \hat{\epsilon}_\bullet^\top \hat{\epsilon}_\bullet \; + \; 0 \; + \; 0 \; + \; (\hat{\dot{y}}_\bullet - p_\bullet)^\top (\hat{\dot{y}}_\bullet - p_\bullet) \\ 
  &= \lVert \hat{\epsilon}_\bullet \rVert^2 \; + \; \lVert \hat{\dot{y}}_\bullet - p_\bullet \rVert^2 \\
  &\ge \lVert \hat{\epsilon}_\bullet \rVert^2 \\ 
  &= \lVert \dot{y_\bullet} - \hat{\dot{y}}_\bullet \rVert^2
\end{align} 
$$ {#eq-lin-reg-min-distance}

There is one more point worth noting here.  Suppose $\dot{X}_{\bullet, \bullet}$ consisted of just a single column, say $\dot{m}_\bullet$.

$$
\begin{align} 
  \dot{X}_{\bullet, \bullet} &= \dot{m}_\bullet \\ \\ 
  \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet} 
  &= \dot{m}_\bullet^\top \dot{m}_\bullet \\ 
  &= \lVert \dot{m}_\bullet \rVert^2
\end{align} 
$$ {#eq-1D-cov-mat}

Then we would have: 

$$
\begin{align} 
  P
  &= \dot{X}_{\bullet, \bullet} \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top \\ 
  &= \dot{m}_\bullet 
    \frac{1}{\lVert \dot{m}_\bullet \rVert^2} \; \dot{m}_\bullet^\top \\ 
  &= \left ( \frac{\dot{m}_\bullet}{\lVert \dot{m}_\bullet \rVert} \right ) 
     \left ( \frac{\dot{m}_\bullet}{\lVert \dot{m}_\bullet \rVert} \right )^\top \\ 
  &= u_\bullet \; u_\bullet^\top \\ \\ 
  \text{where} \\ \\
  u_\bullet &= \frac{\dot{m}_\bullet}{\lVert \dot{m}_\bullet \rVert} \\ \\ 
  \text{so that} \\ \\
  \lVert u_\bullet \rVert &= 1
\end{align} 
$$ {#eq-1D-projection-mat}

To summarize, 

  - The parental centered heights $(\dot{m}_\bullet, \dot{f}_\bullet)$ constitute a basis of the 2-dimensional subspace (the "parental" subspace) that they span within $n-$space.
  - Matrix $P$ sends the vector $\dot{y}_\bullet$ of sons' centered heights to prediction vector $\hat{\dot{y}_\bullet}$.
  - $P$ is the orthogonal projection of $n-$space onto the parental subspace.
  - Moreover, $(\hat{\beta}_1, \hat{\beta}_2)$ are the coordinates of the prediction vector in this subspace with respect to the parental basis.
  - The formula for $P$ generalizes the 1-dimensional projection $u_\bullet \; u_\bullet^\top$ where $u_\bullet$ is a _unit vector_, that is where $\lVert u_\bullet \rVert = 1$
  - Of all the vectors in the parental subspace, the prediction vector $\hat{\dot{y}_\bullet} = P \; \dot{y}_\bullet$ is the closest (in Euclidean distance) to the given vector $\dot{y}_\bullet$.

@fig-smf-smpl-ctr above shows the result of projecting the centered sons' heights to their predicted values in the parental plane.  Each point in that figure represents an individual family, which corresponds to a single row of the centered feature matrix $\dot{X}_{\bullet, \bullet}$.  In the next section we introduce a different perspective on linear regression, namely a column-based view.

## Column versus Row Visualization

Continuing with the example of centered heights, let's now take a step back from two explanatory variables to just one, namely the mother's centered height $\dot{m}_\bullet$ as a predictor of the son's centered height $\dot{s}_\bullet$.  From @eq-1D-projection-mat we have 

$$
\begin{align} 
  P
  &= \left ( \frac{\dot{m}_\bullet}{\lVert \dot{m}_\bullet \rVert} \right ) 
     \left ( \frac{\dot{m}_\bullet}{\lVert \dot{m}_\bullet \rVert} \right )^\top \\ \\ 
  \hat{\beta}_\bullet &= \left ( \dot{X}_{\bullet, \bullet}^\top \dot{X}_{\bullet, \bullet}  \right )^{-1} \dot{X}_{\bullet, \bullet}^\top \; \dot{y}_\bullet \\
  &= \frac{\dot{m}_\bullet^\top \; \dot{y}}{\lVert \dot{m}_\bullet \rVert^2} \\ 
  &= \hat{\beta}_1 \\ \\ 
  \text{so that} \\ \\ 
  \hat{\dot{y}} &= P \; \dot{y} \\ 
  &= \left ( \frac{\dot{m}_\bullet}{\lVert \dot{m}_\bullet \rVert} \right ) 
     \left ( \frac{\dot{m}_\bullet}{\lVert \dot{m}_\bullet \rVert} \right )^\top \; \dot{y} \\ 
  &= \frac{\dot{m}_\bullet^\top \; \dot{y}}{\lVert \dot{m}_\bullet \rVert^2} \; \dot{m}_\bullet \\ 
  &= \hat{\beta}_1 \; \dot{m}_\bullet 
\end{align} 
$$ {#eq-1D-ctr-lin-reg-soln}

@fig-gg-sm-smpl-ctr shows this projection from $\dot{y}_\bullet$ (that is, $\dot{s}_\bullet$) to the one-dimensional space spanned by $\dot{m}_\bullet$.

```{r}
#| label: smf-smpl-cor

# correlation matrix
smf_smpl_cor <- smf_smpl_ctr |> 
  dplyr::select(mother, father, son) |> 
  stats::cor()

```

```{r}
#| label: theta-ms

cor_ms   <- smf_smpl_cor [["mother", "son"]]
theta_ms <- acos(cor_ms)

```

```{r}
#| label: sm-grob-prep

# define data and parameters to create next figure

grob_prep_lst <- ob_2d_grob_prep(
  u_theta  = 0,    # <dbl> u_base = (cos, sin)(u_theta)
  v_theta  = theta_ms, # <dbl> v_base = (cos, sin)(v_theta)
  n_pts    = 5L,   # <int> generates ((1:n_pts) - 1) / (n_pts - 1)
  u_name = "m",    # <chr> desired name for vector u
  v_name = "s"     # <chr> desired name for vector v
)

# basis vectors
m_base <- grob_prep_lst$ uv_base_lst$ u_base
s_base <- grob_prep_lst$ uv_base_lst$ v_base

# segments defining oblique grid
seg_tbl <- grob_prep_lst$ seg_tbl

# plotting limits
x_min_plt <- grob_prep_lst$ xy_lim_lst$ x_min_plt
x_max_plt <- grob_prep_lst$ xy_lim_lst$ x_max_plt

y_min_plt <- grob_prep_lst$ xy_lim_lst$ y_min_plt
y_max_plt <- grob_prep_lst$ xy_lim_lst$ y_max_plt

# color choices
grid_color <- "gray25"
m_color    <- "purple"
s_color    <- "steelblue"

# annotation settings
# position axis label at (1 +- eps) * basis vector
eps_axis_label <- 0.05

```

```{r}
#| label: fig-gg-sm-smpl-ctr
#| fig-cap: "Centered heights: project s-vector (son) to m-axis (mother)"

## 
#  initialize plot
## 

gg_sm_smpl_ctr <- seg_tbl |> 
  # (x, y) coordinates along (m_axis, s_axis)
  ggplot2::ggplot(mapping = aes(x = x_0, y = y_0)) + 
  # bounding box
  coord_cartesian(
    xlim = c( x_min_plt, x_max_plt ), 
    ylim = c( y_min_plt, y_max_plt )
  ) + 
  # remove standard ggplot() grid-lines, etc.
  theme_void() +
  # retain aspect ratio
  coord_fixed() +
  # main title
  labs(title = "Centered heights: project s (son) to m-axis (mother)")

## 
#  oblique grid-lines
## 

gg_sm_smpl_ctr <- gg_sm_smpl_ctr + 
  geom_segment(mapping = aes(
    x = x_0, xend = x_1, 
    y = y_0, yend = y_1
  ), 
  color = "gray25", linewidth = 0.3
  )

## 
#  oblique axes
## 

# m_axis
gg_sm_smpl_ctr <- gg_sm_smpl_ctr + 
  geom_segment(
    mapping = aes(
      x = 0, xend = m_base [["x"]], 
      y = 0, yend = m_base [["y"]]
    ), 
    arrow = arrow(length = unit(0.3, "cm"), ends = "last"), 
    linewidth = 0.8, color = "purple"
  ) +
  annotate(
    geom = "text", 
    x = (1 - eps_axis_label) * m_base [["x"]], 
    y = - eps_axis_label, 
    label = "m-axis", 
    hjust = 0, color = "purple", 
    fontface = "bold", size = 3.5)

# s_axis
gg_sm_smpl_ctr <- gg_sm_smpl_ctr + 
  geom_segment(
    mapping = aes(
      x = 0, xend = s_base [["x"]], 
      y = 0, yend = s_base [["y"]]
    ), 
    arrow = arrow(length = unit(0.3, "cm"), ends = "last"), 
    linewidth = 0.8, color = "steelblue"
  ) +
  annotate(
    geom = "text", 
    x = (1 + eps_axis_label) * s_base [["x"]], 
    y = (1 + eps_axis_label) * s_base [["y"]], 
    label = "s-axis", 
    hjust = 0, color = "steelblue", 
    fontface = "bold", size = 3.5)

## 
#  project s to m_axis
## 

gg_sm_smpl_ctr <- gg_sm_smpl_ctr +
  geom_segment(
    mapping = aes(
      x = s_base [["x"]], xend = s_base [["x"]],
      y = s_base [["y"]], yend = 0
    ),
    arrow = arrow(
      length = unit(0.3, "cm"), 
      ends = "last", type = "closed"
    ),
    linetype = 3, linewidth = 0.8, color = "purple"
  ) +
  annotate(
    geom = "text",
    x = s_base [["x"]] / 2,
    y = - eps_axis_label,
    label = "paste('(', beta[1], ', 0)')", 
    parse = TRUE, 
    hjust = 0, color = "purple",
    fontface = "bold", size = 3)

gg_sm_smpl_ctr

```

```{r}
#| label: sm-grob-cleanup

# garbage collection: 
# move temporary objects from most recent figure
# from the Global Environment to the following list

sm_grob_prep_lst <- list(
  grob_prep_lst, 
  m_base, s_base, 
  seg_tbl, 
  x_min_plt, x_max_plt, 
  y_min_plt, y_max_plt, 
  grid_color, m_color, s_color, 
  eps_axis_label
)

rm(grob_prep_lst, 
  m_base, s_base, 
  seg_tbl, 
  x_min_plt, x_max_plt, 
  y_min_plt, y_max_plt, 
  grid_color, m_color, s_color, 
  eps_axis_label)

```

The coordinate system of this figure refers to the pair of basis vectors $( \dot{m}_\bullet, \dot{s}_\bullet )$.  If $v$ is a vector in this two-dimensional space then $v$ is some linear combination of the basis vectors.

$$
\begin{align} 
  v &= \sigma \; \dot{s}_\bullet \; + \; \mu \; \dot{m}_\bullet
\end{align} 
$$ {#eq-v-in-ms-ctr-space}

Then $v$ has coordinates $(\sigma, \mu)$ with respect to the $( \dot{m}_\bullet, \dot{s}_\bullet )$ basis.

Consequently the coordinates of vectors $\dot{m}_\bullet$, $\dot{s}_\bullet$, and $\hat{\dot{y}}_\bullet$ are respectively $(1, 0)$, $(0, 1)$, and $(\hat{\beta}_1, 0)$.

Note that the $\dot{s}_\bullet$ axis is not quite perpendicular to the $\dot{m}_\bullet$ axis.  That is because the two vectors are not orthogonal:

$$
\begin{align} 
  \left < \dot{m}_\bullet, \; \dot{s}_\bullet \right > &= \dot{m}_\bullet^\top \; \dot{s}_\bullet \\ 
  &\ne 0
\end{align} 
$$ {#eq-m-s-inner-product}

Instead we have the following non-zero correlation coefficient, denoted here as $r_{m, s}$. [^r-ms-smpl]

[^r-ms-smpl]: The cited value of the correlation coefficient $r_{m, s}$ pertains to the random sample of size 20 created to facilitate a detailed view of regression residuals.  Among all 179 families whose oldest child was a son, we have $r_{m, s} \approx 0.3$.

$$
\begin{align} 
  r_{m, s} &= \left ( \frac{\dot{m}_\bullet}{\lVert \dot{m}_\bullet \rVert} \right )^\top \; 
  \left ( \frac{\dot{s}_\bullet}{\lVert \dot{s}_\bullet \rVert} \right ) \\ 
  &\approx 0.1
\end{align} 
$$ {#eq-m-s-cor}

In linear algebra the expression for $r_{m, s}$ is defined to be the cosine of the angle between vectors $\dot{m}_\bullet$ and $\dot{s}_\bullet$.

Therefore the two axes are shown with the angle, say $\theta_{m, s}$, between the two _drawn_ axes equal to the angle between the two _actual_ vectors, $\dot{m}_\bullet$ and $\dot{s}_\bullet$.  Thus $\cos(\theta_{m, s}) = r_{m, s}$.  Since the correlation coefficient is positive rather than zero, the cosine is also positive, which implies that $\theta_{m, s} < \pi / 2$.

@fig-gg-sm-smpl-ctr is a column-based view of the linear regression of the centered heights of the sons $(\dot{s}_\bullet)$ on the centered heights of their mothers $(\dot{m}_\bullet)$.  The figure illustrates the simplicity of linear least-squares regression as, in essence, an orthogonal projection.

```{r}
#| label: sm-smpl-ctr-lm

sm_smpl_ctr_lm <- lm(
  data = smf_smpl_ctr, 
  formula = son ~ 0 + mother)
```

```{r}
#| label: sm-smpl-ctr

sm_smpl_ctr <- smf_smpl_ctr |> 
  dplyr::select(- c(s_hat, resid)) |> 
  dplyr::mutate(
    s_hat = sm_smpl_ctr_lm$ fitted.values, 
    resid = sm_smpl_ctr_lm$ residuals
  )
```

@fig-sm-smpl-ctr below gives the more commonly used (row-based) illustration of the same linear regression.

```{r}
#| label: fig-sm-smpl-ctr
#| fig-cap: "Centered heights: son ~ mother regression line"

g_sm_smpl_ctr <- sm_smpl_ctr |> 
  ggplot2::ggplot(mapping = aes(
    x = mother, y = son
  )) + 
  geom_point() + 
  geom_smooth(method=lm , color="purple", se=FALSE) + 
  labs(
    title = "Centered heights: son ~ mother regression line")

g_sm_smpl_ctr

```

The two perspectives on linear regression are complementary.  @fig-sm-smpl-ctr portrays individual rows of data (families here).  This is the most common means of visualizing data in two dimensions, and with good reason: it can be very thought-provoking and thus useful for refining models.  This view of the regression problem shows model _results_.  On the other hand, the column-based perspective, illustrated by @fig-gg-sm-smpl-ctr, can help one to understand the model-fitting _process_.

```{r}
#| label: smf-smpl-ctr-basis

## 
# basis unit vectors
## 

# (m, f, s) correlation matrix
#   extract non-diagonal elements
cor_mf <- smf_smpl_cor [["mother", "father"]]
cor_ms <- smf_smpl_cor [["mother", "son"]]
cor_fs <- smf_smpl_cor [["father", "son"]]

# 3-by-6 tibble with rows = (u, v, w)
smf_smpl_uvw_tbl <- cor_xyz_sc(
  cor_12 = cor_mf, 
  cor_13 = cor_ms, 
  cor_23 = cor_fs)

# (x, y, z) columns
#   as matrix: (u, v, w) by (x, y, z)
smf_smpl_xyz_mat <- smf_smpl_uvw_tbl |> 
  dplyr::select(x, y, z) |> 
  as.matrix()
rownames(smf_smpl_xyz_mat) <- smf_smpl_uvw_tbl$ vec
#   vec_xyz from (u, v, w) rows
m_xyz <- smf_smpl_xyz_mat ["u", ]
f_xyz <- smf_smpl_xyz_mat ["v", ]
s_xyz <- smf_smpl_xyz_mat ["w", ]

# (theta, phi) columns
#   as matrix: (u, v, w) by (theta, phi)
smf_smpl_sc_mat <- smf_smpl_uvw_tbl |> 
  dplyr::select(theta, phi) |> 
  as.matrix()
rownames(smf_smpl_sc_mat) <- smf_smpl_uvw_tbl$ vec
#   vec_sc from (u, v, w) rows
m_sc <- smf_smpl_sc_mat ["u", ]
f_sc <- smf_smpl_sc_mat ["v", ]
s_sc <- smf_smpl_sc_mat ["w", ]

## 
# s_proj
## 

# project basis vector s onto (m, f) = (x,y) plane
s_proj_xyz <- s_xyz
s_proj_xyz [["z"]] <- 0

s_proj_sc <- s_sc
s_proj_sc [["phi"]] <- pi/2

```

```{r}
#| label: smf-smpl-ctr-grob-prep

## 
# grob prep
## 
smf_gprep_lst <- ob_3d_grob_prep(
  u_theta  = m_sc [["theta"]], 
  u_phi    = m_sc [["phi"]], 
  
  v_theta  = f_sc [["theta"]], 
  v_phi    = f_sc [["phi"]], 
  
  w_theta  = s_sc [["theta"]], 
  w_phi    = s_sc [["phi"]], 
  
  n_pts    = 5L, 
  u_name = "m", 
  v_name = "f", 
  w_name = "s")

## 
# segments defining oblique grid
## 
seg_tbl <- smf_gprep_lst$ seg_tbl

# extract segments within (m, f) plane

#   parallel to m_xyz along f (v_idx)
m_f_0_seg_tbl <- seg_tbl |> 
  dplyr::filter(p_nm == "m", w_idx == 1L) |> 
  dplyr::select(v_idx, p_bin, x, y, z)
m_f_0_lst <- list()
for (i in 1:(nrow(m_f_0_seg_tbl) %/% 2L)) {
  m_f_0_lst [[i]] <- m_f_0_seg_tbl |> 
    dplyr::filter(v_idx == i) |> 
    dplyr::select(x, y, z) |> 
    as.matrix()
}

#   parallel to f_xyz along m (u_idx)
f_m_0_seg_tbl <- seg_tbl |> 
  dplyr::filter(p_nm == "f", w_idx == 1L) |> 
  dplyr::select(u_idx, p_bin, x, y, z)
f_m_0_lst <- list()
for (i in 1:(nrow(f_m_0_seg_tbl) %/% 2L)) {
  f_m_0_lst [[i]] <- f_m_0_seg_tbl |> 
    dplyr::filter(u_idx == i) |> 
    dplyr::select(x, y, z) |> 
    as.matrix()
}

```

```{r}
#| label: smf-smpl-ctr-plot-limits

# segment min-max
x_seg_min <- smf_gprep_lst$ xyz_minmax_vec [["x_seg_min"]] 
x_seg_max <- smf_gprep_lst$ xyz_minmax_vec [["x_seg_max"]] 

y_seg_min <- smf_gprep_lst$ xyz_minmax_vec [["y_seg_min"]] 
y_seg_max <- smf_gprep_lst$ xyz_minmax_vec [["y_seg_max"]] 

z_seg_min <- smf_gprep_lst$ xyz_minmax_vec [["z_seg_min"]] 
z_seg_max <- smf_gprep_lst$ xyz_minmax_vec [["z_seg_max"]] 

# xyz_range
eps_plot_limit <- 0.1

x_range <- c(x_seg_min, x_seg_max) * (1 + eps_plot_limit)
y_range <- c(y_seg_min, y_seg_max) * (1 + eps_plot_limit)
z_range <- c(z_seg_min, z_seg_max) * (1 + eps_plot_limit)

```

```{r}
#| label: smf-smpl-ctr-plot-annotation

## 
# annotation settings
## 

# color choices
grid_color <- "gray25"
m_color    <- "purple"
f_color    <- "darkorange"
s_color    <- "steelblue"

```

Now let's see how these ideas carry over from 2D to 3D: we now regress $\dot{s}$ on $(\dot{m}, \dot{f})$.  @fig-s3d-2025-10-28a below shows this regression as an orthogonal projection of the $\dot{s}_\bullet$ basis vector to the plane defined by the $(\dot{m}_\bullet, \dot{f}_\bullet)$ basis vectors.  In this $(\dot{m}, \dot{f}, \dot{s})$ coordinate system, the coordinates of the predicted (that is, projected) vector $\hat{\dot{s}}_\bullet$ are $(\hat{\beta}_1, \hat{\beta}_2, 0)$.  The vector of residuals, $\dot{s}_\bullet - \hat{\dot{s}}_\bullet$, is represented by the dotted line orthogonal to the $(\dot{m}, \dot{f})$ plane.

Each of these vectors represents the 20 families in the sample, but those 20 vector elements are not visible from this column-based perspective.  The details of those 20 families, or more generally of individual data cases, are shown in row-based perspectives, like @fig-smf-smpl-ctr.  Such details are important and of interest, of course.  But the column-based perspective illustrates the geometry of the model-fitting process, and also merits our attention. 

```{r}
#| label: fig-s3d-2025-10-28a
#| fig-cap: "Projection of centered heights: son to (mother, father) plane"

## 
# Initialize plot
## 
s3d <- scatterplot3d(
  main = "Projected heights: son to (mother, father) plane",
  x = 0, y = 0, z = 0,
  xlim = x_range, ylim = y_range, zlim = z_range,
  type = "n",
  grid = FALSE,
  box = FALSE,
  angle = 55, 
  axis = FALSE,
  pch = "",
  xlab = "", ylab = "", zlab = ""
)

## 
# Draw grid segments
## 

#  parallel to m along f within (m, f) plane
for (seg in m_f_0_lst) {
  s3d$points3d(
    x = seg[, 1], y = seg[, 2], z = seg[, 3],
    type = "l", col = "gray25", lwd = 0.5)
}

#  parallel to f along m within (m, f) plane
for (seg in f_m_0_lst) {
  s3d$points3d(
    x = seg[, 1], y = seg[, 2], z = seg[, 3],
    type = "l", col = "gray25", lwd = 0.5)
}

## 
# draw axes (basis vectors)
## 

# m_axis (purple)
s3d$points3d(
  x = c(0, m_xyz [["x"]] ), 
  y = c(0, m_xyz [["y"]] ), 
  z = c(0, m_xyz [["z"]] ),
  type = "l", col = "purple", lwd = 2)

# f_axis (darkorange)
s3d$points3d(
  x = c(0, f_xyz [["x"]] ), 
  y = c(0, f_xyz [["y"]] ), 
  z = c(0, f_xyz [["z"]] ),
  type = "l", col = "darkorange", lwd = 2)

# s_axis (steelblue)
s3d$points3d(
  x = c(0, s_xyz [["x"]] ), 
  y = c(0, s_xyz [["y"]] ), 
  z = c(0, s_xyz [["z"]] ),
  type = "l", col = "steelblue", lwd = 2)

## 
#  axis labels
## 

# position axis label at (1 +- eps) * basis vector
eps_axis_label <- 0.15

m_lbl_xyz <- m_xyz * (1 + eps_axis_label)
f_lbl_xyz <- f_xyz * (1 + eps_axis_label)
s_lbl_xyz <- s_xyz * (1 + eps_axis_label)

# "m" label
s3d$points3d(
  x = m_lbl_xyz[["x"]], 
  y = m_lbl_xyz[["y"]], 
  z = m_lbl_xyz[["z"]],
  type = "p", col = "purple", 
  pch = "m", cex = 1.5)

# "f" label
s3d$points3d(
  x = f_lbl_xyz[["x"]], 
  y = f_lbl_xyz[["y"]], 
  z = f_lbl_xyz[["z"]],
  type = "p", col = "darkorange", 
  pch = "f", cex = 1.5)

# "s" label
s3d$points3d(
  x = s_lbl_xyz[["x"]], 
  y = s_lbl_xyz[["y"]], 
  z = s_lbl_xyz[["z"]],
  type = "p", col = "steelblue", 
  pch = "s", cex = 1.5)

## 
# Project basis vector s onto the (m, f) plane
## 

# s_proj, projected point in the (m, f) plane
s3d$points3d(
  x = s_proj_xyz[["x"]], 
  y = s_proj_xyz[["y"]], 
  z = s_proj_xyz[["z"]],
  type = "h", col = "black", 
  pch = 16, cex = 1.2)

# s_proj coefficients label

s3d$points3d(
  x = s_proj_xyz[["x"]] * 1.2, 
  y = s_proj_xyz[["y"]] * 1.2, 
  z = s_proj_xyz[["z"]] * 1.2,
  type = "p", col = "black", pch = "", cex = 0)
text(
  s3d$xyz.convert(
    s_proj_xyz[["x"]] * 1.2, 
    s_proj_xyz[["y"]] * 1.2, 
    s_proj_xyz[["z"]] * 1.2),
  labels = expression(paste(
    "(", beta[1], ", ", beta[2], ")"
  )),
  cex = 1.2, pos = 4)

# residual vector orthogonal to (m, f) plane

s3d$points3d(
  x = c(s_proj_xyz [["x"]], s_xyz [["x"]] ), 
  y = c(s_proj_xyz [["y"]], s_xyz [["y"]] ), 
  z = c(s_proj_xyz [["z"]], s_xyz [["z"]] ),
  type = "l", col = "gray10", lwd = 1.5, lty = 3
)

```

## Re-Centering

Recall that we centered the family heights data in order to present least-squares linear regression as an orthogonal projection $P$ of the vector $\dot{s}_\bullet$ of sons' centered heights onto the subspace spanned by $(\dot{m}_\bullet, \dot{f}_\bullet)$, the vectors of (mother, father) centered heights.  Here are some points to keep in mind:

  1.  Orthogonal projection applies to the original uncentered data, so long as the feature matrix $X_{\bullet, \bullet}$ includes the constant vector $1_\bullet$.

  2.  The coefficient values $(\hat{\beta}_1, \hat{\beta}_2)$ fitted to the centered model $\dot{s} \approx \beta_1 \dot{m} \; + \; \beta_2 \dot{f}$ are equal to the fitted coefficients appearing in the uncentered model $s \approx \beta_0 \; + \; \beta_1 m \; + \; \beta_1 f$.

To elaborate, consider the original data along with a linear regression model that includes the constant coefficient $\beta_0$.

$$
\begin{align} 
  s & \approx \mathcal{l}_{\beta_\bullet} (m, f)  \\ 
  &= \beta_0 \; + \; \beta_1 m \; + \; \beta_2 f 
\end{align} 
$$ {#eq-s-per-mf-lr-model-abstract}

Each set of coefficient values $\beta_\bullet = (\beta_0, \beta_1, \beta_2)$ generates a 2D regression plane, say $\mathcal{L}_{\beta_\bullet}$, within the 3D subspace spanned by the three $n-$vectors $(m_\bullet, f_\bullet, s_\bullet)$.

$$
\begin{align} 
  \mathcal{L}_{\beta_\bullet} &= \mathcal{L}_{\beta_\bullet} (m, f, s) \\ 
  &= \left \{ (m, f, s): s = \mathcal{l}_{\beta_\bullet} (m, f) \right \} 
\end{align} 
$$ {#eq-mf-regresion-plane}

Excluding coefficient $\beta_0$ from the model is equivalent to the constraint, $\beta_0 = 0$.  This constraint restricts the regression plane to pass through the origin.  The role of $\beta_0$ is to account for differences in the average heights of mothers, fathers, and sons, respectively, which we denote as $(\bar{m}, \bar{f}, \bar{s})$.  That is, the inclusion of $\beta_0$ enables us to set 

$$
\begin{align} 
  \tilde{\beta}_0 &= \bar{s} - (\beta_1 \bar{m} \; + \; \beta_2 \bar{f}) \\ \\ 
  &\text{to ensure} \\ \\
  \mathcal{l}_{\beta_\bullet} (\bar{m}, \bar{f}) &= \tilde{\beta}_0 \; + \; \beta_1 \bar{m} \; + \; \beta_2 \bar{f} \\ 
  &= \bar{s}
\end{align} 
$$ {#eq-smf-tilde-beta-0}

That is, this formula for $\tilde{\beta}_0$ ensures that the model maps parents of average height to a son of average height.  A model of this form can be re-expressed as a model of centered heights.

$$
\begin{align} 
  & \text{If} & \tilde{\beta}_0 &= \bar{s} - (\beta_1 \bar{m} \; + \; \beta_2 \bar{f}) \\ 
  & \text{and} & s & \approx \tilde{\beta}_0 \; + \; \beta_1 m \; + \; \beta_2 f  \\ 
  & \text{then} & s - \bar{s} & \approx \beta_1 (m - \bar{m}) \; + \; \beta_2 (f - \bar{f})
\end{align} 
$$ {#eq-s-per-mf-lr-centered}

Now let's return to the representation of the model using the feature matrix $X_{\bullet, \bullet}$, the coefficient vector $\beta_\bullet$, and the vector of recorded heights of mothers, fathers, and sons $(m_\bullet, f_\bullet, s_\bullet)$.  Then we have 

$$
\begin{align} 
  s_\bullet & \approx \mathcal{l}_{\beta_\bullet} (m_\bullet, f_\bullet)  \\ 
  &= \beta_0 1_\bullet \; + \; \beta_1 m_\bullet \; + \; \beta_2 f_\bullet \\ 
  &= (1_\bullet, m_\bullet, f_\bullet) \; \beta_\bullet \\ 
  &= X_{\bullet, \bullet} \; \beta_\bullet
\end{align} 
$$ {#eq-s-per-mf-lr-model-vector}

## Summary

## Exercises

### Least Squares: Two Perspectives

### Data Visualization

- The 3D scatterplot view: each point is (mother_height, father_height, son_height)
- The fitted regression plane:  =  + mother + father
- This is how we visualize and interpret, but NOT where the projection occurs
- Dimension: p = 3 (number of features including intercept)

### Sample Space

- Each observation is a vector in  (n = 179)
- The response vector y   has components [y, y, ..., y]
- Each column of X is also a vector in 
- Example: the "mother_heights" column is one vector in 179-dimensional space

### Column Space

- C(X) is the 3-dimensional subspace of  spanned by the columns of X
- All possible linear combinations: X for any   
- Key insight: C(X) contains all predictions our model can make
- Dimension: rank(X)  min(n,p) = 3 (assuming full column rank)

### Orthogonal Projection

-  = X is the orthogonal projection of y onto C(X)
- The residual vector e = y -  is orthogonal to C(X)
- Geometric interpretation:  is the point in C(X) closest to y (in L distance)
- The normal equations: XX  = Xy arise from orthogonality condition Xe = 0
- Include a 2D schematic diagram: y, C(X) as a plane,  as projection, e perpendicular

### Reconciling Perspectives

- The 3D scatterplot shows relationships between variables (p-dimensional)
- The projection happens in observation space (n-dimensional)
- Both are valid and useful for different purposes
- The regression coefficients  connect the two views

### Why This Matters

- Degrees of freedom: n - p (observations minus parameters)
- Overfitting: what happens when p approaches n?
- Preview: other norms (L) change the geometry but still live in 
