# MCMC: Markov Chain Monte Carlo {#sec-mcmc-appendix}

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo    = TRUE, 
  error   = FALSE, 
  message = FALSE, 
  warning = FALSE
)
```

```{r}
#| label: CRAN-libraries

library(astsa)
library(tidyverse)

```

```{r}
#| label: local-libraries

library(eda4mlr)

```

------------------------------------------------------------------------

This appendix provides a detailed treatment of Markov Chain Monte Carlo (MCMC) methods, which are fundamental to Bayesian statistics and probabilistic machine learning. The material assumes familiarity with basic probability theory and Markov chains.[^book-mcmc-peng]

[^book-mcmc-peng]: See @Peng_2022_adv_stat_comp [chap. 7] for additional background.

## Overview

Markov Chain Monte Carlo (MCMC) simulation encompasses many variations. The basic problem to be addressed is similar to that of importance sampling:

-   We would like to generate a random sample from a distribution having given density, $p(x), x \in \mathbb{R}^d$.

-   We cannot directly generate the desired random sample, although we can evaluate $p(x)$ for any given $x \in \mathbb{R}^d$.

The approach toward a solution can be framed as follows:

-   Construct a Markov chain $X: \mathbb{Z} \rightarrow \mathbb{R}^d$ (a discrete-time random process) having a stationary distribution (to which $X(\cdot)$ converges) given by density $p(\cdot)$.

-   Generate a sufficient number $n$ of samples from $X(\cdot)$ so that $\{ X(n), X(n+1), \ldots \}$ approximates a sample from the desired stationary distribution.

## Markov Chains

Recall the definition of a Markov chain, $X(\cdot)$. The conditional distribution of $X(t)$ given past values, $X(t-1), X(t-2), \ldots$, is just the conditional distribution of $X(t)$ given its preceding value, $X(t-1)$.

To illustrate, suppose that the values of $X(\cdot)$ are restricted to a finite set, say $\{ 1, 2, 3 \}$. Then we define the transition probability $p_{a, b}$ as the following conditional probability.

$$
\begin{aligned}
  p_{a, b} &= P(X(t) = b \; | \; X(t-1) = a)
\end{aligned}
$$ {#eq-discrete-transition-prob}

These transition probabilities are the elements of the following transition probability matrix $T$.

$$
\begin{aligned}
  T &= 
  \left (
  \begin{matrix}
    p_{1, 1} & p_{1,2} & p_{1,3} \\ 
    p_{2, 1} & p_{2,2} & p_{2,3} \\ 
    p_{3, 1} & p_{3,2} & p_{3,3}
  \end{matrix}
  \right )
\end{aligned}
$$ {#eq-transition-matrix-example}

If we denote the (marginal) probability distribution of $X(n)$ as $\pi_n$:

$$
\begin{aligned}
  \pi_n &= \left ( P(X(n) = 1), P(X(n) = 2), P(X(n) = 3) \right )
\end{aligned}
$$ {#eq-marginal-prob-example}

then we have:

$$
\begin{aligned}
  \pi_{n+1} &= \pi_n  \; T
\end{aligned}
$$ {#eq-marginal-prob-transition}

Consequently we have:

$$
\begin{aligned}
  \pi_n &= \pi_0  \; T^n
\end{aligned}
$$ {#eq-marginal-prob-power}

Under regularity conditions[^transition-regularity] there is a unique stationary distribution, say $\pi_\infty$, such that $\pi_n \rightarrow \pi_\infty$. Convergence is accompanied by diminishing values of $\lVert \pi_{n+1} - \pi_n \rVert$.

[^transition-regularity]: A unique, stationary distribution exists if the chain is aperiodic, and if all states communicate and are positive recurrent. For further details see the [Wikipedia article](https://en.wikipedia.org/wiki/Discrete-time_Markov_chain#Stationary_distributions).

## Time Reversal and Detailed Balance

Recall that $p(x)$ is the prescribed stationary distribution of a Markov chain $X(\cdot)$ that we must construct. To do so, we will use the fact that if $X(\cdot)$ is time-reversible, then the marginal distribution of $X(\nu)$ is stationary. This can be seen as follows.

$X(\cdot)$ is time-reversible if the following holds. For any given $n > 0$ define $Y(\nu) = X(n - \nu)$ for $0 \le \nu \le n$. Then the joint probability distribution of the sequence $( X(0), \ldots, X(n) )$ is equal to that of the sequence $( Y(0), \ldots, Y(n) ) = ( X(n), \ldots, X(0) )$.

It follows that the ordered pair of random variables $(X(0), X(n))$ has the same distribution as that of $(X(n), X(0))$, so that $X(0)$ has the same distribution as $X(n)$, and this holds for all $n > 0$. Thus if $X(\cdot)$ is time-reversible its marginal distribution must be stationary.

Now assume for the moment that the values of $X(t)$ form a discrete set, so that:

$$
\begin{aligned}
  P\left ( X(0) = a, X(1) = b \right ) &= \pi_\infty(a) \times p_{a, b} \\ 
  P\left ( X(1) = a, X(0) = b \right ) &= \pi_\infty(b) \times p_{b, a}
\end{aligned}
$$ {#eq-balance-eqns-prep}

If $X(\cdot)$ is time-reversible, it follows that:

$$
\begin{aligned}
  \pi_\infty(a) \times p_{a, b} &= \pi_\infty(b) \times p_{b, a}
\end{aligned}
$$ {#eq-balance-eqns}

Equations of this form are called _detailed balance equations_ (or _local balance equations_). If they hold then marginal distribution $\pi_\infty$ along with transition probabilities $p_{a, b}$ define a stationary Markov chain.

## Metropolis-Hastings Algorithm[^art-MH-algo]

[^art-MH-algo]: See @wiki_MH_algo.

Suppose now that $q(Y \; | \; X )$ is a transition density from which we can generate a random sequence $X(0), X(1), \ldots$, and let $p(x)$ denote the prescribed stationary distribution of $X(\cdot)$. Also suppose that the current state of the generated sequence is $X(t) = x$.

### Outline

The Metropolis-Hastings procedure is an iterative algorithm for generating the next value of the sequence as follows.

1.  Generate a candidate value $y \sim q(Y \; | \; x )$.
2.  Calculate _acceptance ratio_ $\alpha(y \; | \; x)$ shown below.
3.  Generate $u \sim \text{Unif}(0, 1)$. Set the next value to $y$ if $u \le \alpha(y \; | \; x)$. Otherwise set the next value to the current value, $x$.

$$
\begin{aligned}
  \alpha(y \; | \; x) &= \min \left \{1, \frac{ p(y) \; q(x \; | \; y) }{ p(x) \; q(y \; | \; x ) } \right \}
\end{aligned}
$$ {#eq-accept-ratio}

### Metropolis Algorithm

The original algorithm, known as the Metropolis algorithm, is a simpler procedure that generates a random walk. Let $g(\cdot)$ denote a probability density function that is symmetric with respect to $x = 0$. Given the current state $X(t) = x$, generate the candidate value $y$ as:

$$
\begin{aligned}
  y &= x + \epsilon \\ \\ 
  &\text{where } \\ \\ 
  \epsilon &\sim g(\cdot)
\end{aligned}
$$ {#eq-Metro-next-y}

Then the density $q(y \; | \; x)$ of the transition from $x$ to $y = x + \epsilon$ is:

$$
\begin{aligned}
  q(y \; | \; x) &= g(y - x)
\end{aligned}
$$ {#eq-Metro-transition-dens}

Since $g(\cdot)$ is symmetric about zero, $q(y \; | \; x )$ is symmetric in its two arguments.

$$
\begin{aligned}
  q(x \; | \; y) &= g(x - y) \\ 
  &= g(y - x) \\ 
  &= q(y \; | \; x)
\end{aligned}
$$ {#eq-Metro-transition-symm}

Consequently the acceptance ratio simplifies as follows.

$$
\begin{aligned}
  \alpha(y \; | \; x) &= \min \left \{1, \frac{ p(y) \; q(x \; | \; y) }{ p(x) \; q(y \; | \; x ) } \right \} \\ 
  &= \min \left \{1, \frac{ p(y) }{ p(x) } \right \}
\end{aligned}
$$ {#eq-Metro-accept-ratio}

## Example: Sampling from the Standard Normal

For some fixed $\delta > 0$ let $g(\cdot)$ be the uniform density on the interval $(-\delta, \delta)$. Let the specified stationary distribution be given by the standard normal density function $\phi(\cdot)$. Then the Metropolis algorithm outlined above will generate a random walk in discrete time having jump-magnitudes governed by $\delta$.

The steps to generate the next value $y$ given the current value $x$ are a simplification of the steps previously outlined.

1.  Generate candidate value $y = x + \epsilon$, where $\epsilon \sim \text{Unif}(-\delta, \delta)$.
2.  Calculate _acceptance ratio_ $\alpha(y \; | \; x)$ shown below.
3.  Generate $u \sim \text{Unif}(0, 1)$. Set the next value to $y$ if $u \le \alpha(y \; | \; x)$. Otherwise set the next value to the current value, $x$.

$$
\begin{aligned}
  \alpha(y \; | \; x) &= \min \left \{1, \frac{ \phi(y) }{ \phi(x) } \right \}
\end{aligned}
$$ {#eq-Metro-accept-ratio-normal}

### Implementation

Here is R code implementing the Metropolis algorithm for sampling from a standard normal distribution:

```{r}
#| label: alpha-nrm-ln

# Acceptance log-ratio for standard normal density
alpha_nrm_ln <- function(
    y, # <dbl> candidate for next value
    x  # <dbl> current value
) {
  min(dnorm(y, log = TRUE) - dnorm(x, log = TRUE), 0)
}
```

```{r}
#| label: Met-algo-nrm

# Generate samples from standard normal via Metropolis algorithm
gen_nrw <- function(
  delta = 0.25,  # <dbl> max jump-size  
  n_pts = 1024L, # <int> sequence length  
  x_0   = 0      # <dbl> starting value
) {
  x <- vector(mode = "numeric", length = n_pts)
  x[[1]] <- x_0
  eps  <- runif(n_pts, -delta, delta)
  u_ln <- runif(n_pts) |> log()
  
  for (k in 2:n_pts) {
    y <- x[[k - 1L]] + eps[[k]]
    if (u_ln[[k]] <= alpha_nrm_ln(y, x[[k - 1L]])) {
      x[[k]] <- y
    } else {
      x[[k]] <- x[[k - 1L]]
    }
  }
  return(x)
}
```

### Effect of Step Size

Here are three simulations, each starting at $X(0) = 0$, but with $\delta \in \{ \frac{1}{8}, \frac{1}{2}, 2 \}$, respectively. Recall that $| X(t+1) - X(t) | \le \delta$. Thus the trajectories are decreasingly smooth.

```{r}
#| label: normal-rw

set.seed(97)

# Try several values of delta
d_vec <- c(1, 4, 16) / 8

nrm_rw_wide <- tibble::tibble(
  x_1   = gen_nrw(delta = d_vec[[1]]), 
  x_2   = gen_nrw(delta = d_vec[[2]]), 
  x_3   = gen_nrw(delta = d_vec[[3]]), 
  t_idx = 1:length(x_1)
) |> 
  dplyr::select(t_idx, x_1:x_3)

nrm_rw_long <- nrm_rw_wide |> 
  tidyr::pivot_longer(
    cols = x_1:x_3, 
    names_to = "d_idx", 
    names_prefix = "x_", 
    values_to = "z"
  ) |> 
  dplyr::mutate(
    delta_label = dplyr::case_when(
      d_idx == "1" ~ "δ = 1/8",
      d_idx == "2" ~ "δ = 1/2",
      d_idx == "3" ~ "δ = 2"
    )
  )
```

```{r}
#| label: fig-nrw-trajectory
#| fig-cap: "Metropolis random walks with different step sizes"

nrm_rw_long |> 
  dplyr::rename(time = t_idx) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(
    x = time, y = z, color = delta_label
  )) + 
  ggplot2::geom_line(show.legend = FALSE) + 
  ggplot2::facet_grid(
    rows = ggplot2::vars(delta_label)
  ) + 
  ggplot2::labs(
    title = "Metropolis algorithm: effect of step size",
    x = "Time",
    y = "X(t)"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

Next we see histograms of the generated values. For the first and smoothest series more time would have been needed to traverse a greater portion of the normal distribution values.

```{r}
#| label: fig-nrw-hist
#| fig-cap: "Histograms of Metropolis samples with different step sizes"

nrm_rw_long |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = z, fill = delta_label)) + 
  ggplot2::geom_histogram(bins = 30, show.legend = FALSE) + 
  ggplot2::facet_grid(rows = ggplot2::vars(delta_label)) + 
  ggplot2::labs(
    title = "Distribution of Metropolis samples",
    x = "Value",
    y = "Count"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

## Effective Sample Size

Because MCMC samples are correlated (each sample depends on the previous one), the effective amount of information is less than the nominal sample size. The **effective sample size (ESS)** quantifies this.

```{r}
#| label: tbl-nrw-smy
#| tbl-cap: "Summary statistics for Metropolis random walks"

nrm_rw_smy <- nrm_rw_long |> 
  dplyr::group_by(delta_label) |> 
  dplyr::summarise(
    min = min(z), 
    q_1 = stats::quantile(z, probs = 0.25), 
    mean = mean(z), 
    q_3 = stats::quantile(z, probs = 0.75), 
    max = max(z), 
    sd  = sd(z), 
    n   = n(), 
    ESS = round(astsa::ESS(z))
  )

nrm_rw_smy |> 
  knitr::kable(digits = 2, col.names = c("Step size", "Min", "Q1", "Mean", "Q3", "Max", "SD", "n", "ESS"))
```

Comparing the sets of summary statistics above, we see that the first random walk ($\delta = 1/8$) was confined to a narrower range of values and has a much lower effective sample size. The final column is the effective sample size estimated by `astsa::ESS()`.

::: {.callout-note}
## Tuning the Step Size

The choice of $\delta$ involves a trade-off:

- **Too small:** The chain moves slowly through the state space, leading to high autocorrelation and low ESS.
- **Too large:** Many proposals are rejected (because they land in low-probability regions), also leading to slow exploration.

A common rule of thumb is to tune the step size to achieve an acceptance rate between 20% and 50%.
:::

## Diagnostics

When using MCMC in practice, several diagnostics help assess convergence:

1. **Trace plots:** Visual inspection of the chain's trajectory over time. Look for stationarity (no trends) and good mixing (rapid exploration).

2. **Autocorrelation:** High autocorrelation indicates slow mixing. The ESS quantifies the effective number of independent samples.

3. **Multiple chains:** Run several chains from different starting points. They should converge to the same distribution.

4. **Burn-in:** Discard early samples before the chain has reached equilibrium.

## Applications in Machine Learning

MCMC methods are fundamental to:

- **Bayesian inference:** Sampling from posterior distributions over model parameters
- **Probabilistic graphical models:** Inference in complex dependency structures
- **Latent variable models:** Topic models, mixture models
- **Uncertainty quantification:** Generating samples to characterize uncertainty in predictions

Modern variants include Hamiltonian Monte Carlo (HMC) and the No-U-Turn Sampler (NUTS), which use gradient information to make more efficient proposals.

## Resources

[Advanced Statistical Computing](https://bookdown.org/rdpeng/advstatcomp/) by Roger Peng (Chapter 7)

[Metropolis-Hastings algorithm - Wikipedia](https://en.wikipedia.org/wiki/Metropolis%E2%80%93Hastings_algorithm)

[MCMC Interactive Gallery](https://chi-feng.github.io/mcmc-demo/)

[Stan](https://mc-stan.org/) - Modern probabilistic programming with HMC/NUTS
