---
title: "Clustering"
subtitle: "EDA in Higher Dimensions"
author: "Chapter 3"
format:
  revealjs:
    theme: [dark, eda4ml-slides.scss]
    slide-number: true
    incremental: false
    toc: false
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(knitr)
library(tidyverse)
library(ISLR2)

# Load college data
college <- ISLR2::College |> 
  tibble::as_tibble(rownames = "college_name")

# Create z-scores for numeric variables
college_numeric <- college |> 

  dplyr::select(where(is.numeric))

college_z <- college_numeric |> 
  dplyr::mutate(across(everything(), ~ (. - mean(.)) / sd(.)))

# Add college names back for reference
college_z <- college_z |> 
  dplyr::mutate(college_name = college$college_name, .before = 1)
```

# Clustering as EDA

## The Central Question

When we have many variables, how do we simplify?

. . .

**One approach:** Group observations with similar profiles.

. . .

This is **clustering**—EDA extended to higher dimensions.

## Unsupervised Learning

In **supervised learning**, we have a response variable $Y$ to predict from features $X$.

. . .

In **unsupervised learning**, we have only $X$—no labels, no "right answer."

. . .

**Goal:** Discover structure in the data itself.

## The EDA Spirit

> "We are looking for unanticipated patterns in the data."

. . .

Clustering shares EDA's exploratory mindset:

- No single "correct" answer
- Results depend on choices we make
- Interpretation requires domain knowledge

# The College Data

## US Colleges (1995)

```{r}
#| label: tbl-college-vars
#| tbl-cap: "Selected variables from ISLR2::College"

tibble::tibble(
  Variable = c("Private", "Apps", "Accept", "Enroll", "Top10perc", 
               "Expend", "Grad.Rate"),
  Description = c(
    "Private or public institution",
    "Applications received",
    "Applications accepted", 
    "New students enrolled",
    "Pct. from top 10% of H.S. class",
    "Instructional expenditure per student",
    "Graduation rate"
  )
) |> knitr::kable()
```

## Exploring the Data

```{r}
#| label: fig-college-scatter
#| fig-cap: "Expenditure vs. Top 10% students"

college |> 
  ggplot2::ggplot(ggplot2::aes(x = Top10perc, y = Expend, color = Private)) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::labs(
    x = "% from top 10% of H.S. class",
    y = "Expenditure per student ($)"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## Class Exercise

Before running any algorithm:

. . .

1. How many private vs. public schools?
2. What is the range of expenditure?
3. Which variables are correlated?
4. What groupings would you expect to find?

. . .

→ Build intuition before automation.

# Grouping Variables

## Manual Grouping

We can create groups from the data ourselves:

. . .

```{r}
#| label: manual-groups

college_grouped <- college |> 
  dplyr::mutate(
    top10_group = if_else(Top10perc < median(Top10perc), "lower", "upper"),
    expend_group = if_else(Expend < median(Expend), "lower", "upper"),
    group_label = paste0(
      if_else(top10_group == "lower", "0", "1"),
      if_else(expend_group == "lower", "0", "1")
    )
  )
```

Split each variable at its median → four groups.

. . .

| Group | Top10perc | Expend |
|-------|-----------|--------|
| g_00 | below median | below median |
| g_01 | below median | above median |
| g_10 | above median | below median |
| g_11 | above median | above median |

## The Four Groups

```{r}
#| label: fig-four-groups
#| fig-cap: "Manual grouping by median splits"

college_grouped |> 
  ggplot2::ggplot(ggplot2::aes(
    x = Top10perc, y = Expend, 
    color = group_label, shape = group_label
  )) +
  ggplot2::geom_point(alpha = 0.6) +
  ggplot2::geom_vline(xintercept = median(college$Top10perc), linetype = "dashed") +
  ggplot2::geom_hline(yintercept = median(college$Expend), linetype = "dashed") +
  ggplot2::labs(
    x = "% from top 10% of H.S. class",
    y = "Expenditure per student ($)",
    color = "Group", shape = "Group"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## The Limitation

Manual grouping works when we have 2-3 variables.

. . .

But the college data has **17 numeric variables**.

. . .

We need an algorithm that can find groups in high-dimensional space.

# Distance and Standardization

## Euclidean Distance

To measure similarity, we use **distance**:

$$d(a, b) = \sqrt{\sum_{j=1}^{p} (a_j - b_j)^2}$$

. . .

Observations that are "close" in this sense have similar profiles.

## The Scale Problem

```{r}
#| label: tbl-variable-ranges
#| tbl-cap: "Variable ranges differ dramatically"

college_numeric |> 
  dplyr::summarise(across(everything(), list(
    min = ~ min(., na.rm = TRUE),
    max = ~ max(., na.rm = TRUE)
  ))) |> 
  tidyr::pivot_longer(everything(), names_to = c("variable", ".value"), 
                      names_pattern = "(.*)_(.*)") |> 
  dplyr::mutate(range = max - min) |> 
  dplyr::arrange(desc(range)) |> 
  dplyr::slice_head(n = 5) |> 
  knitr::kable(digits = 0, col.names = c("Variable", "Min", "Max", "Range"))
```

. . .

Variables with larger ranges dominate the distance calculation.

## The Solution: Z-Scores

Standardize each variable:

$$z_j = \frac{x_j - \bar{x}_j}{s_j}$$

. . .

After transformation:

- All variables have mean 0 and SD 1
- A 1-unit difference represents one standard deviation
- Variables contribute equally to distance

# K-Means Clustering

## The Core Idea

Partition $n$ observations into $K$ clusters such that each observation belongs to the cluster with the **nearest centroid**.

. . .

**Objective:** Minimize the total within-cluster sum of squares (WCSS):

$$\min \sum_{k=1}^{K} \sum_{i \in C_k} \|x_i - \mu_k\|^2$$

## The Algorithm

1. **Initialize:** Randomly select $K$ observations as initial centroids

. . .

2. **Assign:** Assign each observation to the nearest centroid

. . .

3. **Update:** Recalculate centroids as the mean of assigned observations

. . .

4. **Repeat:** Go to step 2 until assignments stop changing

## K-Means on College Data

```{r}
#| label: kmeans-fit

set.seed(42)

# Use z-scores of numeric variables
college_z_numeric <- college_z |> 
  dplyr::select(-college_name)

kmeans_fit <- stats::kmeans(college_z_numeric, centers = 4, nstart = 25)
```

```{r}
#| label: fig-kmeans-result
#| fig-cap: "K-means clusters (K=4) projected onto two variables"

college |> 
  dplyr::mutate(cluster = factor(kmeans_fit$cluster)) |> 
  ggplot2::ggplot(ggplot2::aes(
    x = Top10perc, y = Expend, 
    color = cluster, shape = cluster
  )) +
  ggplot2::geom_point(alpha = 0.6) +
  ggplot2::scale_shape_manual(values = c(16, 17, 15, 18)) +
  ggplot2::labs(
    x = "% from top 10% of H.S. class",
    y = "Expenditure per student ($)",
    color = "Cluster", shape = "Cluster"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## Practical Note: nstart

K-means depends on random initialization.

. . .

Different starting points → different solutions.

. . .

**Solution:** Run multiple times, keep the best result.

```r
kmeans(X, centers = 4, nstart = 25)
```

. . .

Rule of thumb: `nstart` between 20 and 50.

# Evaluating Clusters

## Within-Cluster Sum of Squares

$$\text{WCSS} = \sum_{k=1}^{K} \sum_{i \in C_k} \|x_i - \mu_k\|^2$$

. . .

**Lower WCSS** = tighter clusters.

. . .

But WCSS always decreases as $K$ increases—even with meaningless clusters.

## Choosing K: The Elbow Method

```{r}
#| label: compute-wcss

set.seed(42)

wcss_values <- tibble::tibble(k = 2:12) |> 
  dplyr::mutate(
    wcss = purrr::map_dbl(k, ~ {
      kmeans(college_z_numeric, centers = .x, nstart = 20)$tot.withinss
    })
  )
```

```{r}
#| label: fig-elbow
#| fig-cap: "WCSS decreases as K increases—look for the 'elbow'"

wcss_values |> 
  ggplot2::ggplot(ggplot2::aes(x = k, y = wcss)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_x_continuous(breaks = 2:12) +
  ggplot2::labs(
    x = "Number of clusters (K)",
    y = "Within-cluster sum of squares"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## Reading the Elbow Plot

**Steep decline:** Adding clusters captures real structure.

. . .

**Gradual decline:** Diminishing returns—overfitting.

. . .

The "elbow" is where the rate of decrease sharply levels off.

. . .

**Caveat:** The elbow is subjective. Domain knowledge matters.

# Comparing Clusterings

## Jaccard Similarity

How do we compare two different clusterings?

. . .

The **Jaccard index** measures agreement by looking at pairs:

$$J = \frac{|A \cap B|}{|A \cup B|}$$

. . .

- $J = 1$: Perfect agreement
- $J \approx 0$: No agreement

. . .

**Use cases:** Compare K-means runs for stability; compare to known groupings.

## Manual Grouping versus K-means

How well does K-means recover our manual grouping?

```{r}
#| label: tbl-cluster-vs-group
#| tbl-cap: "K-means clusters vs. manual groups"

college_grouped |> 
  dplyr::mutate(cluster = factor(kmeans_fit$cluster)) |> 
  dplyr::count(group_label, cluster) |> 
  tidyr::pivot_wider(names_from = cluster, values_from = n, values_fill = 0) |> 
  knitr::kable()
```

. . .

Some clusters align well with manual groups; others split across them.

# Interpretation

## What Do Clusters Mean?

The algorithm found groups—now ask what characterizes each one.

. . .

**Profile each cluster:**

- Compute cluster means for each variable
- Compare to overall means
- Look at distributions within clusters

## Cluster Profiles

```{r}
#| label: tbl-cluster-profiles
#| tbl-cap: "Cluster means for selected variables"

college |> 
  dplyr::mutate(cluster = factor(kmeans_fit$cluster)) |> 
  dplyr::group_by(cluster) |> 
  dplyr::summarise(
    n = n(),
    Top10perc = mean(Top10perc),
    Expend = mean(Expend),
    Grad.Rate = mean(Grad.Rate),
    .groups = "drop"
  ) |> 
  knitr::kable(digits = 0)
```

. . .

What story do these profiles tell?

## Example Interpretation

Review of additional variable profiles suggests:

. . .

- **Cluster 1:** High selectivity, high expenditure → "Elite institutions"
- **Cluster 2:** Lower selectivity, moderate expenditure → "Regional schools"
- **Cluster 3:** Large enrollment, lower cost → "Large state universities"
- **Cluster 4:** Small, moderate selectivity → "Small privates"

. . .

Does this match domain knowledge?

# Connections to ML

## Clustering in the ML Pipeline

Clustering connects to supervised learning:

. . .

**Preprocessing:** Standardization is essential for many ML algorithms, not just clustering.

. . .

**Feature engineering:** Cluster membership can become a derived feature.

. . .

**Model selection:** The elbow heuristic → cross-validation, regularization.

## Other Clustering Methods

| Type | Examples |
|------|----------|
| Centroid-based | K-means, K-medoids |
| Density-based | DBSCAN, HDBSCAN |
| Hierarchical | Agglomerative, Divisive |
| Soft clustering | Gaussian mixture models |

. . .

K-means assumes spherical clusters of similar size. Other methods relax these assumptions.

# Summary

## Chapter 3: Key Takeaways

1. **Clustering is EDA** for high-dimensional data
2. **Standardize first**—z-scores equalize variable contributions
3. **K-means minimizes WCSS** via iterative assignment and update
4. **The elbow method** guides (but doesn't determine) K selection
5. **Interpretation completes** the analysis—profile your clusters

## Key Formulas

| Concept | Formula |
|---------|---------|
| Z-score | $z_j = \frac{x_j - \bar{x}_j}{s_j}$ |
| Euclidean distance | $d(a,b) = \sqrt{\sum (a_j - b_j)^2}$ |
| WCSS | $\sum_k \sum_{i \in C_k} \|x_i - \mu_k\|^2$ |
| Jaccard index | $J = \frac{|A \cap B|}{|A \cup B|}$ |

## Team Exercises

1. Apply K-means to `datasets::iris` measurements
2. Compare clusters to the `Species` variable using `xtabs()`
3. Experiment with different values of K

## Resources

- ISLR2: *An Introduction to Statistical Learning* (Ch. 12)
- tidymodels: [K-means with tidy data principles](https://www.tidymodels.org/learn/statistics/k-means/)
- Wikipedia: [Cluster analysis](https://en.wikipedia.org/wiki/Cluster_analysis)
- CRAN: [dbscan package](https://cran.r-project.org/web/packages/dbscan/)
