---
title: "Conditional Distributions"
subtitle: "EDA for Machine Learning"
author: "Chapter 2"
format:
  revealjs:
    theme: [dark, eda4ml-slides.scss]
    slide-number: true
    incremental: false
    toc: false
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(knitr)
library(tidyverse)
library(UsingR)

father_son_ht <- UsingR::father.son |> 

  tibble::as_tibble() |> 
  dplyr::rename(father = fheight, son = sheight) |>
  dplyr::mutate(
    f_ivl = cut(father, seq(58, 76, 2)), 
    f_mpt = (2 * ceiling(father/2)) - 1
  )

fs_moments <- father_son_ht |> 
  dplyr::summarise(
    f_avg = mean(father, na.rm = TRUE), 
    s_avg = mean(son, na.rm = TRUE), 
    f_sd  = sd(father, na.rm = TRUE), 
    s_sd  = sd(son, na.rm = TRUE), 
    r     = cor(father, son)
  )
```

# Conditional Distributions

## The Central Question

Given information about one variable, what can we say about another?

. . .

**Example:** If we know a father's height, what can we predict about his son's height?

. . .

This leads to the concept of **conditional distributions**.

## Father-Son Heights: Review

```{r}
#| label: fig-scatter-review
#| fig-cap: "Heights of father-son pairs"

father_son_ht |> 
  ggplot2::ggplot(ggplot2::aes(x = father, y = son)) + 
  ggplot2::geom_point(alpha = 0.4) + 
  ggplot2::labs(
    x = "Father's height (inches)",
    y = "Son's height (inches)"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## Conditional Distribution of Son's Height

```{r}
#| label: fig-conditional-boxplot
#| fig-cap: "Son's height distribution for each father height interval"

father_son_ht |> 
  dplyr::filter(!is.na(f_mpt)) |> 
  ggplot2::ggplot(ggplot2::aes(x = forcats::as_factor(f_mpt), y = son)) + 
  ggplot2::geom_boxplot() + 
  ggplot2::labs(
    x = "Father's height (nearest odd inch)",
    y = "Son's height (inches)"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## Conditional Expectation

The **conditional expectation** $E(Y | X)$ is the average value of $Y$ given $X$.

. . .

```{r}
#| label: tbl-conditional-means
#| tbl-cap: "Average son's height per father's height interval"

father_son_ht |>    
  dplyr::filter(!is.na(f_mpt)) |>
  dplyr::group_by(f_mpt) |>    
  dplyr::summarise(     
    n     = dplyr::n(),
    s_avg = mean(son, na.rm = TRUE)
  ) |> 
  knitr::kable(
    digits = 1,
    col.names = c("Father height", "n", "Mean son height")
  )
```

## The Graph of Averages

```{r}
#| label: fig-graph-of-averages
#| fig-cap: "Conditional mean of son's height given father's height"

s_stats_per_f_mpt <- father_son_ht |>    
  dplyr::filter(!is.na(f_mpt)) |>
  dplyr::group_by(f_mpt) |>    
  dplyr::summarise(     
    s_count = dplyr::n(),
    s_avg   = mean(son, na.rm = TRUE)
  ) |>
  dplyr::ungroup()

s_stats_per_f_mpt |> 
  ggplot2::ggplot(ggplot2::aes(x = f_mpt, y = s_avg)) + 
  ggplot2::geom_line(linewidth = 1) + 
  ggplot2::geom_point(size = sqrt(s_stats_per_f_mpt$s_count), shape = 18) + 
  ggplot2::labs(
    x = "Father's height (inches)",
    y = "Average son's height (inches)"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## Key Observation

The graph of averages is approximately **linear**.

. . .

This suggests we can approximate the conditional expectation with a straight line.

. . .

→ The **regression line**

# Z-Scores and Standard Units

## Standardizing Variables

To compare variables on different scales, convert to **standard units** (z-scores):

$$Z_x = \frac{X - \mu_x}{\sigma_x}$$

. . .

**Interpretation:** Number of standard deviations above or below the mean.

## Sample Z-Scores

In practice, we use sample estimates:

$$\hat{Z}_x(x_k) = \frac{x_k - \bar{x}}{s_x}$$

. . .

where $\bar{x}$ is the sample mean and $s_x$ is the sample standard deviation.

# Two Lines Through the Data

## The SD Line

The **SD line** passes through the point of averages with slope $\frac{s_y}{s_x}$:

$$y = \bar{y} + \frac{s_y}{s_x}(x - \bar{x})$$

. . .

Equivalently: $\hat{Z}_y = \hat{Z}_x$

. . .

**Property:** Minimizes sum of squared *perpendicular* distances to points.

## The Regression Line

The **regression line** has slope $r \cdot \frac{s_y}{s_x}$:

$$y = \bar{y} + r \frac{s_y}{s_x}(x - \bar{x})$$

. . .

Equivalently: $\hat{Z}_y = r \cdot \hat{Z}_x$

. . .

**Property:** Minimizes sum of squared *vertical* distances to points.

## Comparing the Two Lines

```{r}
#| label: fig-two-lines
#| fig-cap: "Regression line (blue) vs SD line (red)"

SD_slope <- fs_moments$s_sd / fs_moments$f_sd
R_slope  <- fs_moments$r * SD_slope

father_son_ht |> 
  ggplot2::ggplot(ggplot2::aes(x = father, y = son)) + 
  ggplot2::geom_point(alpha = 0.3) + 
  ggplot2::geom_abline(
    slope = SD_slope, 
    intercept = fs_moments$s_avg - SD_slope * fs_moments$f_avg, 
    linetype = "dashed", linewidth = 1.5, color = "red"
  ) + 
  ggplot2::geom_abline(
    slope = R_slope, 
    intercept = fs_moments$s_avg - R_slope * fs_moments$f_avg, 
    linetype = "solid", linewidth = 1.5, color = "steelblue"
  ) + 
  ggplot2::labs(
    x = "Father's height (inches)",
    y = "Son's height (inches)"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## Why the Difference?

Both lines pass through $(\bar{x}, \bar{y})$.

. . .

The regression line is **less steep** because $|r| \le 1$.

. . .

This is the mathematical basis of **regression to the mean**.

## The Correlation Coefficient

$$r = \frac{1}{n-1} \sum_{k=1}^{n} \hat{Z}_x(x_k) \cdot \hat{Z}_y(y_k)$$

. . .

**Properties:**

- $-1 \le r \le 1$
- $r = \pm 1$ only if points fall exactly on a line
- $r = 0$ means no *linear* association

## Father-Son Correlation

```{r}
#| label: tbl-fs-moments

tibble::tibble(
  Statistic = c("Father mean", "Son mean", "Father SD", "Son SD", "Correlation r"),
  Value = c(
    round(fs_moments$f_avg, 1),
    round(fs_moments$s_avg, 1),
    round(fs_moments$f_sd, 2),
    round(fs_moments$s_sd, 2),
    round(fs_moments$r, 2)
  )
) |> knitr::kable()
```

. . .

With $r \approx 0.5$, the regression line has about half the slope of the SD line.

# Regression Residuals

## Fitted Values and Residuals

For each observation:

- **Fitted value:** $\hat{y}_k = \bar{y} + r \frac{s_y}{s_x}(x_k - \bar{x})$
- **Residual:** $e_k = y_k - \hat{y}_k$

. . .

Residuals measure how far each point falls from the regression line.

## Distribution of Residuals

```{r}
#| label: fig-residuals
#| fig-cap: "Histogram of regression residuals"

lm_fs <- stats::lm(son ~ father, data = father_son_ht)

father_son_ht |> 
  dplyr::mutate(residual = stats::residuals(lm_fs)) |>
  ggplot2::ggplot(ggplot2::aes(x = residual)) + 
  ggplot2::geom_histogram(binwidth = 0.5, fill = "steelblue", color = "white") + 
  ggplot2::labs(
    x = "Residual (inches)",
    y = "Count"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## Residuals Should Be...

- Centered around zero ✓
- Roughly symmetric ✓
- No pattern when plotted against $x$ or $\hat{y}$

. . .

Patterns in residuals suggest the model is missing something.

# The Bivariate Normal

## A Special Case

If $(X, Y)$ follows a **bivariate normal** distribution:

. . .

- The conditional distribution $Y | X$ is normal
- The conditional mean $E(Y|X)$ is exactly the regression line
- The conditional SD is $\sigma_y \sqrt{1 - r^2}$

. . .

Father-son heights are well approximated by a bivariate normal.

## Variance Reduction

Conditioning on $X$ reduces the variance of $Y$:

$$\text{Var}(Y|X) = \sigma_y^2 (1 - r^2)$$

. . .

For father-son data with $r \approx 0.5$:

$$\sqrt{1 - r^2} \approx 0.87$$

. . .

Knowing father's height reduces son's height SD by about 13%.

# Cautionary Tales

## Robust Statistics

The mean and SD are sensitive to outliers.

. . .

**Alternatives:**

| Sensitive | Robust |
|-----------|--------|
| Mean | Median |
| Standard deviation | IQR |

. . .

The regression line inherits this sensitivity.

## Anscombe's Quartet

Four data sets with **identical** summary statistics:

- Same means, SDs, and correlation
- Same regression line

. . .

Yet the data look completely different!

## Anscombe's Quartet: The Data

```{r}
#| label: fig-anscombe
#| fig-cap: "Four data sets with identical regression statistics"

anscombe_long <- datasets::anscombe |>
  tibble::as_tibble() |>
  dplyr::mutate(idx = dplyr::row_number()) |>
  tidyr::pivot_longer(
    cols = -idx,
    names_to = c(".value", "group"),
    names_pattern = "(.)(.)"
  )

anscombe_long |>
  ggplot2::ggplot(ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point(size = 2) +
  ggplot2::geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  ggplot2::facet_wrap(~group, ncol = 4) +
  ggplot2::theme_minimal(base_size = 14)
```

## The Lesson

**Always visualize your data!**

. . .

Summary statistics can hide:

- Nonlinear relationships
- Outliers
- Clusters
- Data errors

# Statistical Independence

## Definition

Random variables $(X, Y)$ are **independent** if:

$$P(X \in A, Y \in B) = P(X \in A) \cdot P(Y \in B)$$

for all sets $A$ and $B$.

. . .

**Implication:** Knowing $X$ tells you nothing about $Y$.

## Independence and Correlation

If $(X, Y)$ are independent, then $r = 0$.

. . .

**But the converse is false!**

. . .

$r = 0$ only means no *linear* association.

Variables can be dependent but uncorrelated.

## Testing Independence: Categorical Variables

For categorical variables, use the **chi-squared test**.

. . .

**Idea:** Compare observed counts to expected counts under independence.

$$\chi^2 = \sum_{j,k} \frac{(O_{jk} - E_{jk})^2}{E_{jk}}$$

## Example: Handedness and Sex

```{r}
#| label: tbl-handedness

hs_tbl <- tibble::tribble(
    ~handedness, ~male, ~female,
    "right",  934L, 1070L,
    "left",   113L,   92L,
    "ambi",    20L,    8L
  )

hs_tbl |> knitr::kable()
```

. . .

Is handedness independent of sex?

## Chi-Squared Test Result

```{r}
#| label: chi-sq-test

hs_matrix <- hs_tbl |> 
  dplyr::select(male, female) |> 
  as.matrix()

hs_chi_sq <- stats::chisq.test(hs_matrix)
```

$\chi^2 =$ `r round(hs_chi_sq$statistic, 1)`, df = `r hs_chi_sq$parameter`, p-value = `r round(hs_chi_sq$p.value, 4)`

. . .

**Conclusion:** Strong evidence against independence.

Males are more likely to be left-handed or ambidextrous.

# Simpson's Paradox

## UC Berkeley Admissions (1973)

```{r}
#| label: tbl-ucb-overall

tibble::tibble(
  Sex = c("Male", "Female"),
  `Admission Rate` = c("44.5%", "30.4%")
) |> knitr::kable()
```

. . .

This looks like clear evidence of bias against women.

## But Wait...

```{r}
#| label: tbl-ucb-by-dept

tibble::tibble(
  Dept = c("A", "B", "C", "D", "E", "F"),
  `Male %` = c(62, 63, 37, 33, 28, 6),
  `Female %` = c(82, 68, 34, 35, 24, 7)
) |> knitr::kable()
```

. . .

Four of six departments admitted women at **higher** rates!

## What Happened?

Women applied disproportionately to departments with **low overall admission rates**.

. . .

Department is a **confounding variable**.

. . .

The aggregate pattern reverses when we condition on department.

## Simpson's Paradox: The Lesson

A pattern in aggregated data can **reverse** when data are disaggregated by a relevant variable.

. . .

Always ask: Is there a confounding variable I'm missing?

# Measures of Association

## What We've Covered

| Variables | Measure |
|-----------|---------|
| Both continuous | Correlation $r$ |
| Both categorical | Chi-squared $\chi^2$ |

. . .

Both measure departure from independence.

## Looking Ahead

**Information theory** provides an alternative framework:

- Entropy
- Mutual information
- KL divergence

. . .

These capture **nonlinear** relationships that correlation might miss.

→ Chapter 6

# Summary

## Chapter 2: Key Takeaways

1. **Conditional distributions** show how one variable varies given another
2. The **graph of averages** can be approximated by the **regression line**
3. **Correlation** measures linear association; $r = 0$ doesn't mean independence
4. **Always visualize** — summary statistics can deceive (Anscombe)
5. **Simpson's paradox** reminds us to look for confounders

## Key Formulas

| Concept | Formula |
|---------|---------|
| Z-score | $\hat{Z}_x = \frac{x - \bar{x}}{s_x}$ |
| Regression line | $\hat{Z}_y = r \cdot \hat{Z}_x$ |
| Correlation | $r = \frac{1}{n-1}\sum \hat{Z}_x \hat{Z}_y$ |
| Chi-squared | $\chi^2 = \sum \frac{(O - E)^2}{E}$ |

## Team Exercises

1. Simulate bivariate normal variables
2. Find another example of Simpson's paradox
3. Construct dependent variables with $r = 0$

## Resources

- Freedman, Pisani, Purves: *Statistics* (4e)
- Bickel et al.: "Sex Bias in Graduate Admissions" (1975)
- Wikipedia: Simpson's paradox
