---
title: "Exploratory Data Analysis"
subtitle: "EDA for Machine Learning"
author: "Chapter 1"
format:
  revealjs:
    theme: [dark, eda4ml-slides.scss]
    slide-number: true
    incremental: false
    toc: true
    toc-depth: 1
    toc-title: "Chapter 1"
    preview-links: auto
    progress: true
    hash: true
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(knitr)
library(tidyverse)
library(UsingR)

father_son_ht <- UsingR::father.son |> 
  as_tibble() |> 
  rename(father = fheight, son = sheight)
```

# What is EDA?

## Exploratory Data Analysis

**John Tukey (1977):** Statisticians had become too focused on formal inference at the expense of simply *looking at data*.

. . .

EDA is the corrective:

- Visualization
- Summary
- Iterative development of questions

## EDA vs. Confirmatory Analysis

| Confirmatory | Exploratory |
|--------------|-------------|
| Begins with a hypothesis | Begins with data |
| Asks: do data support or refute? | Asks: what patterns are present? |
| Tests pre-specified claims | Generates new hypotheses |

. . .

A complete analysis typically involves **both**.

## Why EDA Matters for ML

1. **Understanding data before modeling**
   - Variable types, distributions, missing values, outliers

2. **Generating hypotheses**
   - Patterns become candidates for formal modeling

3. **Diagnosing models**
   - Residual plots reveal what models miss

4. **Communicating findings**
   - Visualizations convey results to stakeholders

## EDA is a Mindset

> A willingness to look at data from multiple angles, to transform variables, to compare subgroups, and above all to **ask questions**.

# Foundational Concepts

## Data as Sample from Population

:::: {.columns}

::: {.column width="50%"}
**Population**

The complete set of cases we care about

- All possible customers
- All manufacturing runs
- All images of a certain type
:::

::: {.column width="50%"}
**Sample**

The subset we actually observe

- Our goal: learn about the population
- EDA helps assess whether the sample supports generalization
:::

::::

## Central Value: Two Approaches

:::: {.columns}

::: {.column width="50%"}
**Mean (Average)**

$$\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i$$

- Minimizes sum of squared deviations
- Sensitive to outliers
:::

::: {.column width="50%"}
**Median**

Middle value when sorted

- Half above, half below
- Robust to outliers
- Often better for skewed data
:::

::::

## Dispersion: Two Approaches

:::: {.columns}

::: {.column width="50%"}
**Standard Deviation**

$$s = \sqrt{\frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^2}$$

- Typical distance from mean
- Sensitive to outliers
:::

::: {.column width="50%"}
**Interquartile Range (IQR)**

$$\text{IQR} = Q_3 - Q_1$$

- Spread of middle 50%
- Robust to outliers
:::

::::

## Choosing Among Summaries

**Key insight:** Compute *both* and compare.

. . .

Large discrepancy between mean and median signals:

- Skewness, or
- Outliers

. . .

→ Warrants further investigation

## Questions Drive the Analysis

**Standard starting questions:**

- What is the distribution of each variable?
- Are there outliers or unusual values?
- Are there missing values?
- How are pairs of variables related?

. . .

**Domain-specific questions** arise from context and purpose.

## EDA Throughout Modeling

| Stage | EDA Role |
|-------|----------|
| **Before** | Understand distributions, detect quality issues |
| **During** | Compare models, identify difficult cases |
| **After** | Examine residuals, check assumptions |

# Example: Father-Son Heights

## Historical Context

**Francis Galton (1885):** Investigated whether height is hereditary

**Karl Pearson:** Founded first statistics department (UCL, 1911); collected father-son height data

. . .

This data set introduced:

- Correlation
- Regression
- "Regression to the mean"

## The Data

```{r}
#| label: data-summary

tibble::tibble(
  Statistic = c("Number of pairs", "Father mean height", "Son mean height"),
  Value = c(
    nrow(father_son_ht),
    paste0(round(mean(father_son_ht$father), 1), " inches"),
    paste0(round(mean(father_son_ht$son), 1), " inches")
  )
) |> knitr::kable()
```

## First Look: Scatter Plot

```{r}
#| label: fig-scatter
#| fig-cap: "Heights of father-son pairs"

father_son_ht |> 
  ggplot2::ggplot(aes(x = father, y = son)) + 
  ggplot2::geom_point(alpha = 0.4) + 
  ggplot2::labs(
    x = "Father's height (inches)",
    y = "Son's height (inches)"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## What the Scatter Plot Reveals

- **Positive association:** Taller fathers → taller sons
- **Roughly linear** relationship
- **Substantial scatter** around the trend
- **Density in middle** makes individual points hard to see

. . .

One visualization → one question answered, new questions raised

## Conditional Distributions

```{r}
#| label: fig-boxplot
#| fig-cap: "Son's height by father's height interval"

father_son_ht |> 
  dplyr::mutate(f_mpt = (2 * ceiling(father/2)) - 1) |>
  dplyr::filter(!is.na(f_mpt)) |> 
  ggplot2::ggplot(aes(x = as_factor(f_mpt), y = son)) + 
  ggplot2::geom_boxplot() + 
  ggplot2::labs(
    x = "Father's height (nearest odd inch)",
    y = "Son's height (inches)"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## What the Box Plots Reveal

- **Median increases** with father's height
- **Spread (IQR) similar** across groups
- A few **outliers** among sons of shorter fathers

. . .

This shows the **conditional distribution** of son's height given father's height.

→ Central concept in Chapter 2

## Are Sons Taller Than Fathers?

```{r}
#| label: fig-compare
#| fig-cap: "Comparing height distributions"

height_per_fs <- father_son_ht |> 
  dplyr::mutate(idx = row_number()) |>
  dplyr::select(idx, father, son) |> 
  tidyr::pivot_longer(
    cols = c(father, son), 
    names_to = "generation", 
    values_to = "height"
  )

height_per_fs |> 
  ggplot2::ggplot(aes(x = generation, y = height)) + 
  ggplot2::geom_boxplot() + 
  ggplot2::labs(x = "", y = "Height (inches)") +
  ggplot2::theme_minimal(base_size = 14)
```

## Summary: Fathers vs. Sons

```{r}
#| label: tbl-compare

height_per_fs |> 
  dplyr::group_by(generation) |> 
  dplyr::summarise(
    n = n(),
    mean = mean(height),
    median = median(height),
    sd = sd(height)
  ) |>
  knitr::kable(
    digits = 1, 
    col.names = c("Generation", "n", "Mean", "Median", "SD")
  )
```

. . .

Sons are about **one inch taller** on average.

## Paired Differences

```{r}
#| label: fig-diff
#| fig-cap: "Distribution of son's height minus father's height"

father_son_ht |> 
  dplyr::mutate(diff = son - father) |>
  ggplot2::ggplot(aes(x = diff)) + 
  ggplot2::geom_histogram(binwidth = 1, fill = "steelblue", color = "white") + 
  ggplot2::geom_vline(
    xintercept = 0, linetype = "dashed", 
    color = "red", linewidth = 1
  ) + 
  ggplot2::labs(x = "Height difference (inches)", y = "Count") +
  ggplot2::theme_minimal(base_size = 14)
```

## Regression to the Mean

**Galton's observation:**

> Extremely tall fathers tend to have sons who are tall, but *not quite as extreme*.

. . .

This is a **statistical phenomenon**, not a genetic one.

It arises whenever two variables are correlated but not perfectly so.

. . .

→ Explored mathematically in Chapter 2

# The Practice of EDA

## Showing Your Work

Throughout this book, code is generally hidden.

. . .

But seeing *how* visualizations are constructed is part of learning EDA:

```{r}
#| label: ggplot-demo-code
#| echo: true
#| eval: false

father_son_ht |> 
  ggplot2::ggplot(aes(x = father, y = son)) + 
  ggplot2::geom_point(alpha = 0.4) + 
  ggplot2::labs(
    x = "Father's height (inches)",
    y = "Son's height (inches)"
  )
```

## The Grammar of Graphics

1. Start with **data**
2. Specify **aesthetic mappings** (variables → visual properties)
3. Add **geometric objects** (points, lines, bars)
4. Customize **labels** and appearance

. . .

This framework (ggplot2) provides flexibility for constructing visualizations.

## Class Exercises

**Diamond Data** (ggplot2::diamonds)

- ~54,000 diamonds with price and attributes
- Explore distributions, relationships, predictors of price

. . .

**General Social Survey** (forcats::gss_cat)

- US survey tracking social attitudes
- Practice with categorical variables
- Explore changes over time

## Guiding Questions for Any Data Set

1. How many observations? Variables?
2. What are the variable types?
3. What are the distributions?
4. Are there missing values? Outliers?
5. What relationships exist between variables?
6. What would I need to know to make a decision?

# Looking Ahead

## Part 1: Foundations of EDA

| Chapter | Topic |
|---------|-------|
| 2 | Conditional Distributions |
| 3 | Clustering |
| 4 | Statistical Simulation |
| 5 | Sampling and Study Design |
| 6 | Information Theory |

## Key Themes

- **Conditional distributions** underlie regression and classification
- **Clustering** exemplifies EDA in higher dimensions
- **Simulation** builds intuition about variability
- **Study design** determines what conclusions are valid
- **Information theory** provides vocabulary for ML

## The EDA Mindset

> Look at the data, ask questions, and let the data guide your understanding.

. . .

This mindset remains central throughout the book.

# Summary

## Chapter 1: Key Takeaways

1. **EDA is question-driven** exploration of data
2. **Central value and dispersion** have multiple implementations
3. **Data = sample from population** frames our inferences
4. **Multiple views** of the same data reveal different patterns
5. **EDA continues throughout** the modeling process

## Resources

- Tukey (1977): *Exploratory Data Analysis*
- Wickham et al.: *R for Data Science* (r4ds.hadley.nz)
- Wickham: *ggplot2: Elegant Graphics for Data Analysis*

## Team Exercises

1. Response vs. predictor variables
2. Regression to the mean: evidence in the data?
3. Diamond data: explore one relationship with two visualizations
4. Survey data: investigate change over time
5. Find/construct an example of misleading summaries
