---
title: "Statistical Simulation"
subtitle: "Exploring What-If Scenarios"
author: "Chapter 4"
format:
  revealjs:
    theme: [dark, eda4ml-slides.scss]
    slide-number: true
    incremental: false
    toc: true
    toc-depth: 1
    toc-title: "Chapter 4"
    preview-links: auto
    progress: true
    hash: true
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(knitr)
library(tidyverse)

set.seed(42)
```

# Why Simulate?

## The Central Question

How might our plans, methods, model $\ldots$ fail?

. . .

**Simulation** lets us explore scenarios we can't observe directly.

. . .

Generate data from known processes → see how methods behave.

## Uses of Simulation

1. **Compare methods:** Which estimator performs better?

. . .

2. **Assess robustness:** What happens with outliers? Small samples?

. . .

3. **Understand systems:** How does randomness propagate?

. . .

4. **Quantify uncertainty:** How precise are our estimates?

## The EDA Connection

Simulation extends EDA beyond the data we have.

. . .

> "What would happen if $\dots$?"

. . .

- $\dots$ the sample size were larger?
- $\dots$ the distribution had heavier tails?
- $\dots$ there were outliers?

. . .

Simulation lets us **explore failure modes** before they occur in practice.

# Random Number Generation

## R's Distribution Functions

For each distribution, R provides four functions:

. . .

| Prefix | Function | Example |
|--------|----------|---------|
| `d` | density/mass | `dnorm(0)` → 0.399 |
| `p` | cumulative probability | `pnorm(1.96)` → 0.975 |
| `q` | quantile (inverse CDF) | `qnorm(0.975)` → 1.96 |
| `r` | random generation | `rnorm(10)` → 10 values |

## Continuous Distributions (selected)

```{r}
#| label: tbl-continuous-dist

tibble::tribble(
  ~Function, ~Distribution,
  "[d,p,q,r]norm", "Normal",
  "[d,p,q,r]unif", "Uniform",
  "[d,p,q,r]exp", "Exponential",
  "[d,p,q,r]gamma", "Gamma",
  "[d,p,q,r]beta", "Beta",
  "[d,p,q,r]t", "Student t",
  "[d,p,q,r]chisq", "Chi-squared"
) |> knitr::kable()
```

See `?Distributions` for a complete list.

## Discrete Distributions (selected)

```{r}
#| label: tbl-discrete-dist

tibble::tribble(
  ~Function, ~Distribution,
  "[d,p,q,r]binom", "Binomial",
  "[d,p,q,r]pois", "Poisson",
  "[d,p,q,r]geom", "Geometric",
  "[d,p,q,r]nbinom", "Negative Binomial"
) |> knitr::kable()
```

. . .

The Poisson is the "law of rare events"—the limit of Binomial$(n, p)$ as $n \to \infty$ and $p \to 0$ with $np \to \lambda$.

## Generating Random Samples

```{r}
#| label: rng-demo
#| echo: true
#| eval: false

# 100 standard normal values
x <- rnorm(n = 100, mean = 0, sd = 1)

# 100 uniform values on [0, 1]
u <- runif(n = 100, min = 0, max = 1)

# 100 Poisson values with mean 5
counts <- rpois(n = 100, lambda = 5)
```

. . .

**Key insight:** We specify the distribution; R generates samples.

# Comparing Estimators

## Mean vs. Median

Both estimate the center of a symmetric distribution.

. . .

**Mean:** $\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$

. . .

**Median:** $\hat{m}$ = middle value (or average of two middle values)

. . .

Which is better? **It depends on the distribution.**

## The Efficiency Question

For normal data, theory tells us:

$$\frac{\text{Var}(\hat{m})}{\text{Var}(\hat{\mu})} \approx \frac{\pi}{2} \approx 1.57$$

. . .

The median has **57% higher variance** than the mean.

. . .

Can we verify this by simulation?

## Simulation Design

```{r}
#| label: sim-design
#| echo: true
#| eval: false

# Parameters
n <- 30      # sample size
R <- 10000   # number of replications

# Storage
means <- numeric(R)
medians <- numeric(R)

# Simulation loop
for (r in 1:R) {
  x <- rnorm(n)
  means[r] <- mean(x)
  medians[r] <- median(x)
}

# Compare variances
var(medians) / var(means)
```

## Simulation Results

```{r}
#| label: mean-median-sim

n <- 30
R <- 10000

sim_results <- tibble::tibble(
  rep = 1:R,
  sample = purrr::map(rep, ~ rnorm(n)),
  mean = purrr::map_dbl(sample, mean),
  median = purrr::map_dbl(sample, median)
)

var_ratio <- var(sim_results$median) / var(sim_results$mean)
```

```{r}
#| label: fig-mean-median
#| fig-cap: "Sampling distributions of mean and median (n=30, R=10,000)"

sim_results |> 
  tidyr::pivot_longer(cols = c(mean, median), names_to = "estimator", values_to = "value") |> 
  ggplot2::ggplot(ggplot2::aes(x = value, fill = estimator)) +
  ggplot2::geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  ggplot2::labs(
    x = "Estimate",
    y = "Count",
    fill = "Estimator"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## Verifying Theory

```{r}
#| label: tbl-mean-median-var

tibble::tibble(
  Estimator = c("Mean", "Median"),
  Variance = c(var(sim_results$mean), var(sim_results$median)),
  `Std Error` = c(sd(sim_results$mean), sd(sim_results$median))
) |> 
  knitr::kable(digits = 4)
```

. . .

**Variance ratio:** `r round(var_ratio, 3)` (theory: $\pi/2 \approx$ `r round(pi/2, 3)`)

. . .

Simulation confirms the theoretical result.

## Why This Matters

We verified a known result. But simulation shines when:

. . .

- Theory is intractable
- Assumptions are violated
- We need to compare many methods

. . .

**EDA application:** "How would my estimator behave if the data were slightly different?"

# Monte Carlo Methods

## Estimating π

Classic example: estimate $\pi$ by random sampling.

. . .

**Idea:** A quarter circle of radius 1 has area $\pi/4$.

. . .

1. Generate random points $(X, Y)$ uniform on $[0,1]^2$
2. Count points inside the quarter circle: $X^2 + Y^2 < 1$
3. Proportion inside $\approx \pi/4$

## Visual Intuition

```{r}
#| label: fig-pi-estimate
#| fig-cap: "Monte Carlo estimation of π"

n_points <- 5000
pi_data <- tibble::tibble(
  x = runif(n_points),
  y = runif(n_points),
  inside = (x^2 + y^2) < 1
)

pi_estimate <- 4 * mean(pi_data$inside)

pi_data |> 
  ggplot2::ggplot(ggplot2::aes(x = x, y = y, color = inside, shape = inside)) +
  ggplot2::geom_point(alpha = 0.6, size = 1.0) +
  ggplot2::stat_function(fun = ~ sqrt(1 - .x^2), color = "white", linewidth = 1) +
  ggplot2::coord_fixed() +
  ggplot2::scale_color_manual(values = c("FALSE" = "#E74C3C", "TRUE" = "#2ECC71")) +
  ggplot2::scale_shape_manual(values = c("FALSE" = 4, "TRUE" = 16)) +
  ggplot2::labs(
    title = paste0("Estimate: ", round(pi_estimate, 4), " (true: ", round(pi, 4), ")"),
    color = "Inside circle",
    shape = "Inside circle"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

## The Monte Carlo Principle

To estimate $E[g(X)]$:

$$\hat{\theta} = \frac{1}{n}\sum_{i=1}^n g(X_i)$$

. . .

where $X_1, \ldots, X_n$ are random samples from the distribution of $X$.

. . .

**Law of Large Numbers:** $\hat{\theta} \to E[g(X)]$ as $n \to \infty$

## Convergence Rate

```{r}
#| label: fig-pi-convergence
#| fig-cap: "Monte Carlo estimate converges to true value"

# Cumulative estimate of pi
cumulative_pi <- tibble::tibble(
  n = 1:n_points,
  cumsum_inside = cumsum(pi_data$inside),
  estimate = 4 * cumsum_inside / n
)

cumulative_pi |> 
  ggplot2::ggplot(ggplot2::aes(x = n, y = estimate)) +
  ggplot2::geom_line(color = "steelblue") +
  ggplot2::geom_hline(yintercept = pi, linetype = "dashed", color = "red") +
  ggplot2::scale_x_log10() +
  ggplot2::labs(
    x = "Number of samples (log scale)",
    y = "Estimate of π"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

. . .

Standard error decreases as $1/\sqrt{n}$.

# Rare Events

## The Problem

What if we want to estimate $P(Z > 6)$ where $Z \sim N(0,1)$?

. . .

```{r}
#| label: rare-event-prob
#| echo: true

# True probability
pnorm(6, lower.tail = FALSE)
```

. . .

About 1 in a billion. Naive simulation won't work.

## Naive Approach Fails

```{r}
#| label: naive-sim
#| echo: true

# Generate 1 million standard normals
z <- rnorm(1e6)

# Count how many exceed 6
sum(z > 6)
```

. . .

We'd need billions of samples to see even one event.

## Importance Sampling

**Key idea:** Sample from a different distribution, then reweight.

. . .

Instead of sampling from $p(x)$, sample from $q(x)$ where rare events are common.

. . .

$$E_p[g(X)] = E_q\left[g(X) \cdot \frac{p(X)}{q(X)}\right]$$

. . .

The ratio $w(x) = p(x)/q(x)$ is the **importance weight**.

## Importance Sampling for $P(Z > 6)$

**Strategy:** Shift the normal distribution to center it near 6.

. . .

Let $q(x) = \phi(x - 6)$ (normal centered at 6).

. . .

```{r}
#| label: importance-sampling
#| echo: true

# Sample from shifted normal
n <- 10000
x <- rnorm(n, mean = 6, sd = 1)

# Importance weights: p(x) / q(x)
log_weights <- dnorm(x, log = TRUE) - dnorm(x, mean = 6, log = TRUE)
weights <- exp(log_weights)

# Estimate P(Z > 6)
estimate <- mean((x > 6) * weights)
estimate
```

## Comparing Estimates

```{r}
#| label: tbl-importance-sampling

tibble::tibble(
  Method = c("True value", "Importance sampling", "Naive (n = 10^6)"),
  Estimate = c(
    format(pnorm(6, lower.tail = FALSE), scientific = TRUE, digits = 3), 
    format(estimate, scientific = TRUE, digits = 3), 
    "0"
  )
) |> knitr::kable()
```

. . .

Importance sampling makes the impossible feasible.

# Bootstrap

## Quantifying Uncertainty

How precise is our estimate?

. . .

Classical approach: derive standard error formula.

. . .

**Bootstrap approach:** Resample from the data itself.

## The Bootstrap Idea

1. Treat the sample as a proxy for the population
2. Draw **with replacement** from the sample
3. Compute statistic on each resample
4. Use variability across resamples to estimate uncertainty

## Bootstrap in Action

```{r}
#| label: bootstrap-demo
#| echo: true
#| eval: false

# Original sample
x <- c(12, 15, 18, 22, 25, 28, 31, 35, 42, 55)

# Bootstrap
B <- 10000
boot_means <- numeric(B)

for (b in 1:B) {
  x_star <- sample(x, replace = TRUE)
  boot_means[b] <- mean(x_star)
}

# 95% confidence interval
quantile(boot_means, c(0.025, 0.975))
```

## Bootstrap Example

```{r}
#| label: bootstrap-example

x <- c(12, 15, 18, 22, 25, 28, 31, 35, 42, 55)
B <- 10000

boot_means <- purrr::map_dbl(1:B, ~ mean(sample(x, replace = TRUE)))
boot_ci <- quantile(boot_means, c(0.025, 0.975))

# Normal theory CI for comparison
n_x <- length(x)
se_x <- sd(x) / sqrt(n_x)
normal_ci <- mean(x) + c(-1, 1) * qt(0.975, df = n_x - 1) * se_x
```

```{r}
#| label: fig-bootstrap
#| fig-cap: "Bootstrap distribution of the sample mean"

tibble::tibble(mean = boot_means) |> 
  ggplot2::ggplot(ggplot2::aes(x = mean)) +
  ggplot2::geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  ggplot2::geom_vline(xintercept = mean(x), color = "red", linewidth = 1) +
  ggplot2::geom_vline(xintercept = boot_ci, color = "red", linetype = "dashed") +
  ggplot2::labs(
    x = "Bootstrap sample mean",
    y = "Count",
    title = paste0("Bootstrap 95% CI: [", round(boot_ci[1], 1), ", ", round(boot_ci[2], 1), "]")
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

. . .

Normal theory 95% CI: [`r round(normal_ci[1], 1)`, `r round(normal_ci[2], 1)`] — similar result.

## Bootstrap for Any Statistic

The bootstrap works for almost any statistic:

. . .

- Correlation coefficients
- Regression coefficients
- Median, trimmed mean
- Ratios, differences

. . .

**EDA application:** "How much would this pattern change with different data?"

# Connections to ML

## Simulation in the ML Pipeline

Simulation connects to machine learning:

. . .

**Cross-validation:** Repeatedly resample to estimate generalization error.

. . .

**Stochastic gradient descent:** Random sampling of training batches.

. . .

**Bayesian ML:** MCMC and variational inference sample from posteriors.

## Simulation as Model Checking

After fitting a model:

. . .

1. Simulate data from the fitted model
2. Compare simulated data to real data
3. Discrepancies reveal model failures

. . .

This is the **posterior predictive check**—simulation for model criticism.

# Summary

## Chapter 4: Key Takeaways

1. **Simulation explores what-if scenarios** before they occur

2. **R provides d/p/q/r functions** for common distributions

3. **Monte Carlo methods** estimate quantities via random sampling

4. **Importance sampling** handles rare events

5. **Bootstrap** quantifies uncertainty without formulas

## Key Formulas

| Concept | Formula |
|---------|---------|
| Monte Carlo estimate | $\hat{\theta} = \frac{1}{n}\sum_{i=1}^n g(X_i)$ |
| Importance weight | $w(x) = p(x) / q(x)$ |
| Variance ratio (mean vs. median) | $\pi/2 \approx 1.57$ |

## The Simulation Workflow

1. **Specify** the data-generating process
2. **Generate** random samples
3. **Compute** statistics of interest
4. **Repeat** many times
5. **Summarize** the distribution of results

## Team Exercises

1. Modify the mean vs. median simulation to use Cauchy data (`rcauchy()`). What happens to the variance ratio?

2. Use bootstrap to estimate a 95% CI for the median of `mtcars$mpg`.

3. Estimate $P(Z > 4)$ using importance sampling. Compare to `pnorm(4, lower.tail = FALSE)`.

## Resources

- [Advanced Statistical Computing](https://bookdown.org/rdpeng/advstatcomp/) by Roger Peng
- [Monte Carlo method - Wikipedia](https://en.wikipedia.org/wiki/Monte_Carlo_method)
- [Bootstrap Methods](https://en.wikipedia.org/wiki/Bootstrapping_(statistics))
- R documentation: `?Distributions`
