<!-- 
  ADDITION 3: eda.qmd
  Insert after line 198 (after "...What is the typical deviation from the trend?).")
  Insert before line 200 ("### Conditional Distributions: Box Plots...")
  
  This placement follows the scatter plot introduction, exploring the 
  marginal distribution of father's heights before moving to conditionals.
-->

### Tabular and Graphical Summaries

R provides several functions for computing and displaying these summaries. The `summary()` function produces a quick overview of a numeric vector:

```{r}
#| label: summary-father
#| echo: true

summary(father_son_ht$father)
```

This output displays the five-number summary (minimum, Q1, median, Q3, maximum) plus the mean—sufficient for a first assessment of center, spread, and potential skewness. When mean and median are close, the distribution is roughly symmetric; when they diverge, skewness is present.

For graphical summaries, histograms and density plots serve complementary purposes. A histogram partitions the data into bins and counts observations in each bin:

```{r}
#| label: fig-father-histogram
#| fig-cap: "Distribution of father's heights"

father_son_ht |> 
  ggplot2::ggplot(mapping = aes(x = father)) + 
  ggplot2::geom_histogram(
    binwidth = 1, 
    fill = "steelblue", 
    color = "white") + 
  ggplot2::labs(
    x = "Father's height (inches)",
    y = "Count"
  )
```

A density plot smooths the histogram into a continuous curve, making the overall shape easier to perceive:

```{r}
#| label: fig-father-density
#| fig-cap: "Kernel density estimate of father's heights"

father_son_ht |> 
  ggplot2::ggplot(mapping = aes(x = father)) + 
  ggplot2::geom_density(fill = "steelblue", alpha = 0.5) + 
  ggplot2::labs(
    x = "Father's height (inches)",
    y = "Density"
  )
```

The density curve is produced by *kernel density estimation* (KDE), a smoothing technique that places a small "bump" (typically Gaussian) at each observation and sums them. The base R function `stats::density()` computes KDE; `ggplot2::geom_density()` wraps this for plotting. Smoothing is a recurring theme in data analysis—we will encounter it again when fitting curves to scatter plots in @sec-conditioning.

The choice between histogram and density plot is partly a matter of taste. Histograms reveal the actual counts and make gaps or ties in the data visible. Density plots emphasize the continuous shape and facilitate comparison across groups (by overlaying multiple curves). In practice, examining both is often worthwhile.

