# Clustering: EDA in Higher Dimensions {#sec-clustering}

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo    = FALSE, 
  error   = FALSE, 
  message = FALSE, 
  warning = FALSE
)
```

```{r}
#| label: CRAN-libraries

library(assertthat)
library(dbscan)
library(GGally)
library(here)
library(ISLR2)
library(latex2exp)
library(tidyverse)
library(tinytex)
library(UsingR)

```

```{r}
#| label: conflicts-prefer

# resolve conflicts with plotly
# conflicted::conflicts_prefer(plotly::layout)

# resolve conflicts with tidymodels packages
# tidymodels::tidymodels_prefer()

```

```{r}
#| label: local-libraries

library(eda4mldata)

```

```{r}
#| label: local-source

# stabilize simulated data
source(here("code", "eda4ml_set_seed.R"))

source(here("code", "rmse_per_grp.R"))
source(here("code", "xtabs_to_jaccard.R"))

```

------------------------------------------------------------------------

## Introduction

When we encounter a new data set with many variables, our first challenge is simplification. With two or three variables, we can make scatterplots. But what about ten variables? Fifty? We cannot visualize high-dimensional data directly, so we need other strategies.

**Clustering** offers one such strategy: group observations that have similar profiles across many variables, then explore these groups. This approach is firmly in the spirit of exploratory data analysis (EDA). We are not testing hypotheses or optimizing a loss function against known labels—we are looking for unanticipated patterns in the data.

In machine learning terminology, clustering is an example of **unsupervised learning**: algorithms that seek structure in data without the benefit of a response variable or observation labels. Most unsupervised methods are designed for high-dimensional settings where direct visualization fails.

::: {.callout-note}
## The EDA mindset

Clustering shares EDA's exploratory character:

- There is no single "correct" answer
- Results depend on choices we make (distance metric, number of clusters, algorithm)
- Interpretation requires domain knowledge
- The goal is insight, not prediction
:::

In this chapter we'll discuss how to group observations having similar profiles. We begin with manual grouping to build intuition, then introduce K-means clustering as an algorithmic approach, and conclude with a brief look at alternative methods.

## US Colleges

```{r}
#| label: college

college <- ISLR2::College |> 
  as_tibble(rownames = "college_name")
```

```{r}
#| label: college-vars-tbl

college_vars_tbl <- tibble::tibble(
  var_name = names(college),
) |> 
  mutate(
    dscr = case_when(
      var_name == "college_name" ~ 
        "Name of the college or university", 
      var_name == "Private" ~ 
        "No or Yes indicating private or public", 
      var_name == "Apps" ~ 
        "Number of applications received", 
      var_name == "Accept" ~ 
        "Number of applications accepted", 
      var_name == "Enroll" ~ 
        "Number of new students enrolled", 
      var_name == "Top10perc" ~ 
        "Pct. new students from top 10% of H.S. class", 
      var_name == "Top25perc" ~ 
        "Pct. new students from top 25% of H.S. class", 
      var_name == "F.Undergrad" ~ 
        "Number of fulltime undergraduates", 
      var_name == "P.Undergrad" ~ 
        "Number of parttime undergraduates", 
      var_name == "Outstate" ~ 
        "Out-of-state tuition", 
      var_name == "Room.Board" ~ 
        "Room and board costs", 
      var_name == "Books" ~ 
        "Estimated book costs", 
      var_name == "Personal" ~ 
        "Estimated personal spending", 
      var_name == "PhD" ~ 
        "Pct. of faculty with PhD's", 
      var_name == "Terminal" ~ 
        "Pct. of faculty with terminal degree", 
      var_name == "S.F.Ratio" ~ 
        "Student/faculty ratio", 
      var_name == "perc.alumni" ~ 
        "Pct. alumni who donate", 
      var_name == "Expend" ~ 
        "Instructional expenditure per student", 
      var_name == "Grad.Rate" ~ 
        "Graduation rate"
    )
  )
```

To illustrate ideas we'll use data about US universities and colleges from 1995, described in the book, *An Introduction to Statistical Learning with applications in R* ([ISLR](https://www.statlearning.com)). The data are recorded within the `R` package `ISLR2` as `ISLR2::College`. From the `R` command `help("College")` we see that the data consist of 777 observations (data rows), and obtain the following description of the variables (data columns).

```{r}
#| label: tbl-college-vars-tbl
#| tbl-cap: "Variables in the ISLR2::College data set"

college_vars_tbl |> knitr::kable(
  caption = "US college variables (1995 issue of USNWR)", 
  col.names = c("variable", "description")
)
```

## Grouping Variables

Clustering algorithms construct groups of observations based solely on statistical information obtained from a given data set, with no reliance on the meaning or importance of the data variables.

A complementary way to form groups is to use so-called "grouping variables" that we select or construct from the data. Often, the grouping variables are evident from the questions we're trying to answer, e.g., to compare spending patterns among different socio-economic groups.

When we want to construct groups (clusters) of observations as a means of exploring the data, we should consider what would make one grouping (clustering) more or less effective than another for this purpose. We'll return to this point.

### Example: college data grouping variables

To help us consider the goals and criteria of clustering (grouping), let's return to the US college data and choose a few variables with which to group the data.

Consider the distinction between more expensive versus less expensive schools as indicated by variable `Expend` (instructional expenditure per student). We might ask, "Are the more expensive schools able to attract more of the top students?" So let's also use the variable `Top10perc` (percent of new students coming from the top 10% of their high school class). We'll simply split each variable at its median value, creating two binary variables, which we'll use as our grouping variables.

Before proceeding, let's examine the distribution of these variables via density estimates and a scatter plot.

```{r}
#| label: college-factors

# split selected variables at their median value
college_factors <- college |> 
  mutate(
    top_10_fct = Top10perc  < median(Top10perc),
    expend_fct = Expend     < median(Expend)
  ) |> 
  # for binary factors above, re-code (T, F) to ("lwr", "upr")
  mutate(across(
    .cols = c(top_10_fct, expend_fct), 
    .fns  = ~ if_else(.x, "lwr", "upr")
  )) |> 
  # label each group with a bitcode
  mutate(
    top_10_lbl = if_else(top_10_fct == "lwr", 0L, 1L), 
    expend_lbl = if_else(expend_fct == "lwr", 0L, 1L), 
    grp_lbl    = paste0("g_", top_10_lbl, expend_lbl)
  ) |> 
  # remove redundant labeling vars
  dplyr::select(- top_10_lbl, - expend_lbl)

# record the median values
top_10_mid  <- median(college_factors$ Top10perc)
expend_mid  <- median(college_factors$ Expend)
```

```{r}
#| label: fig-grp-var-matrix
#| fig-cap: "Distribution of variables `Top10perc` and `Expend`"

# Form a matrix of figures, showing each 
# distinct pair of variables in a scatter diagram.
g_grp_var_matrix <- college_factors |> 
  GGally::ggpairs(
    columns = c("Top10perc", "Expend"), 
    title   = "Variables (Top10perc, Expend)"
  )
g_grp_var_matrix
```

We see that `Top10perc` and `Expend` are positively correlated, and have relatively few departures from a monotonic relationship. Also note that `Expend` is measured (in US dollars) on a very different scale from `Top10perc`.

Now let's look at the two variables more closely, one at a time.

```{r}
#| label: fig-top-10
#| fig-cap: "Pct of new students per college from top 10% in their HS"

g_top_10 <- college_factors |> 
  ggplot(mapping = aes(x = Top10perc)) + 
  geom_histogram() + 
  geom_vline(
    xintercept = top_10_mid, linewidth = 2, colour = "red"
  ) + 
  labs(
    title = "% of new students from top 10% in their HS", 
    subtitle = "(vertical line at median)"
  )
g_top_10
```

```{r}
#| label: fig-expend
#| fig-cap: "Instructional expenditure per student"

g_Expend <- college_factors |> 
  ggplot2::ggplot(mapping = aes(x = Expend)) + 
  ggplot2::geom_histogram() + 
  ggplot2::geom_vline(
    xintercept = expend_mid, linewidth = 2, colour = "red"
  ) + 
  ggplot2::labs(
    title = "Instructional expenditure per student", 
    subtitle = "(vertical line at median)"
  )
g_Expend
```

The figures above show decent variation on either side of the median value. Therefore cutting each variable at its median value seems reasonable, as long as we bear in mind that each distribution has a right tail (a quite long tail for `Expend`).

For each of the variables shown we create a binary variable having two levels, "lwr" and "upr", meaning an observation is either below the median value, or else is no less than the median value. We then form 4 groups, each a combination of levels of the 2 binary variables, which we label $g_{00}, g_{01}, g_{10}, g_{11}$ to denote the binary factors based on `Top10perc`, and `Expend` (in that order), with {"lwr", "upr"} coded as {0, 1}.

To check this grouping of the data let's first select the variables from which we defined the 4 groups and examine the scatter diagram, now using the color and shape of each data point to distinguish the different groups.

```{r}
#| label: fig-top-10-expend
#| fig-cap: "(Top10perc, Expend): binary grouping"

g_top_10_Expend <- college_factors |> 
  ggplot2::ggplot(mapping = aes(
    x = Top10perc, y = Expend, 
    colour = grp_lbl, shape = grp_lbl
  )) + 
  ggplot2::scale_shape_manual(values = 0:3) +
  ggplot2::geom_point() + 
  ggplot2::labs(
    title = "(Top10perc, Expend)", 
    subtitle = "grouped by binary versions of the two variables"
  )
g_top_10_Expend
```

The figure above confirms that we have constructed the binary grouping variables as we intended.

The table below shows the mean per group of the underlying grouping variables.

Note: we've calculated the mean per group of *all* the numeric variables. We'll return to those calculations in the next section.

```{r}
#| label: means-per-grp

means_per_grp <- college_factors |> 
  # remove original non-numeric variables
  dplyr::select(- college_name, - Private) |> 
  # reduce grouping variables to just grp_lbl
  dplyr::select(- top_10_fct, - expend_fct) |> 
  # vector of mean values per group
  dplyr::summarise(dplyr::across(
    .cols = tidyselect::everything(), 
    .fns  = ~ mean(.x, na.rm = TRUE)
  ), 
  # grouping variable
  .by = "grp_lbl", 
  # group size: number of original data rows per group
  count = n()
  ) |> 
  dplyr::arrange(grp_lbl)
```

```{r}
#| label: tbl-means-per-grp
#| tbl-cap: "(`Top10perc`, `Expend`): mean values per group"

means_per_grp |> 
  dplyr::select(
    grp_lbl, count, Top10perc, Expend
  ) |> 
  knitr::kable(
    caption = "Mean values per group of underlying grouping variables", 
    col.names = c("group", "# obs", "Top10perc", "Expend"), 
    digits = 1
  )
  
```

### Measures of clustering effectiveness

In the context of clustering algorithms, groups are not given but are rather to be calculated from the data. But under what criteria? Broadly speaking we want the points (each an observation of several numeric variables) to be close together within each cluster (group), and we want the clusters to be well separated from one another.

So how well do the groups constructed above perform against these criteria? We want points within each cluster (group) to be close to one another, but how should we measure distance between points? Euclidean distance springs to mind, but there is a catch: the numeric variables are recorded on differing scales.

### Standardizing variables

The following table shows the range (minimum, maximum) of each numeric variable in the college data.

```{r}
#| label: college-ranges

# calculate range of each numeric variable
college_ranges <- college_factors |> 
  # remove original non-numeric variables
  dplyr::select(- college_name, - Private) |> 
  # remove grouping variables
  dplyr::select(- top_10_fct, - expend_fct, - grp_lbl) |> 
  dplyr::summarise(dplyr::across(
    .cols = tidyselect::everything(), 
    .fns  = list(
      min = ~ min(.x, na.rm = TRUE),
      max = ~ max(.x, na.rm = TRUE)
    )
  )) |> 
  tidyr::pivot_longer(
    cols = everything(),
    names_to = c("var", ".value"),
    names_pattern = "(.*)_(.*)"
  ) |> 
  dplyr::mutate(range = max - min) |> 
  dplyr::arrange(desc(range))
```

```{r}
#| label: tbl-college-ranges
#| tbl-cap: "College variables: range (max - min)"

college_ranges |> 
  dplyr::slice_head(n = 8) |>
  knitr::kable(
    caption = "Variables with largest ranges", 
    col.names = c("variable", "min", "max", "range"), 
    digits = 0
  )
```

The financial variables and enrollment counts range into the tens of thousands, whereas the percentages merely range from 0 to 100. If we compute Euclidean distances using raw values, the high-range variables will dominate. A difference of 1,000 in `Apps` would overwhelm a difference of 10 in `Grad.Rate`, even though both might represent meaningful distinctions.

This issue can be addressed by **standardizing** each numeric variable: subtract the overall mean and divide by the standard deviation. The result is a **z-score**:

$$z_j = \frac{x_j - \bar{x}_j}{s_j}$$

After standardization, all variables have mean 0 and standard deviation 1. A 1-unit difference in any standardized variable represents one standard deviation—a comparable "statistical distance" across all variables.

```{r}
#| label: college-z-scores

# convert numeric data into standard units (z-scores)
college_z_scores <- college_factors |> 
  # remove original non-numeric variables
  dplyr::select(- college_name, - Private) |> 
  # grouping variables: use only grp_lbl
  dplyr::select(- top_10_fct, - expend_fct) |> 
  # standardize numeric variables
  dplyr::mutate(across(
    .cols = - grp_lbl, 
    .fns  = ~ (.x - mean(.x))/sd(.x)
  ))

# count numeric variables (exclude grp_lbl)
n_z_score_vars <- ncol(college_z_scores) - 1L
```

The code below demonstrates the z-score transformation:
  
```{r}
#| label: z-score-demo
#| echo: true
#| eval: false

# Standardize numeric variables to z-scores
college_z_scores <- college_numeric |> 
  dplyr::mutate(across(
    .cols = everything(), 
    .fns  = ~ (.x - mean(.x)) / sd(.x)
  ))
```

### Within-group distances in standard units

Having standardized the college numeric variables we now calculate Euclidean distances. That is, within each group we calculate the distance of each point to the group mean vector. This vector of point-minus-mean is referred to as "error" (more precisely, "mean deviation") in statistical parlance. We calculate the squared Euclidean length of this vector, and sum these squared distances across all points in the group. The result is called the (within-group) sum of squared errors (SSE). Here are the calculated SSE values (and related statistics) per group.

```{r}
#| label: coll-z-stats-per-grp

# compute z-score means and std-devs per group

# number of observations per group as a data frame
coll_counts_per_grp <- college_z_scores |> 
  dplyr::summarise(
    .by = grp_lbl, 
    count = n()
  ) |> 
  dplyr::arrange(grp_lbl) |> 
  tibble::column_to_rownames("grp_lbl")

coll_z_means_per_grp <- college_z_scores |> 
  dplyr::summarise(
    .by = grp_lbl, 
    dplyr::across(
      .cols = tidyselect::everything(), 
      .fns  = ~ mean(.x, na.rm = TRUE)
    )
  ) |> 
  dplyr::arrange(grp_lbl)

coll_z_sds_per_grp <- college_z_scores |> 
  dplyr::summarise(
    .by = grp_lbl, 
    dplyr::across(
      .cols = tidyselect::everything(), 
      .fns  = ~ sd(.x, na.rm = TRUE)
    )
  ) |> 
  dplyr::arrange(grp_lbl)

# transpose coll_z_sds_per_grp
grp_z_sds_per_var <- coll_z_sds_per_grp |> 
  tibble::column_to_rownames("grp_lbl") |> 
  t() |> 
  tibble::as_tibble(rownames = "var")

grp_z_ss <- grp_z_sds_per_var |> 
  # (RMSE per var) to (MSE per var)
  dplyr::mutate( dplyr::across(
    .cols = - var, 
    .fns  = ~ .x^2
  )) |> 
  # (MSE per var) to (SSE per var)
  dplyr::mutate(
    g_00 = g_00 * (-1 + coll_counts_per_grp[["g_00", "count"]]), 
    g_01 = g_01 * (-1 + coll_counts_per_grp[["g_01", "count"]]), 
    g_10 = g_10 * (-1 + coll_counts_per_grp[["g_10", "count"]]), 
    g_11 = g_11 * (-1 + coll_counts_per_grp[["g_11", "count"]])
  ) |> 
  # sum (SSE per var) across vars
  dplyr::summarise(dplyr::across(
    .cols = - var, 
    .fns  = ~ sum(.x, na.rm = TRUE)
  )) |> 
  # transpose to facilitate summing across groups
  tidyr::pivot_longer(
    cols = everything(), 
    names_to = "group", 
    values_to = "SSE"
  ) |> 
  # calculate (count, df, MSE, RMSE) per group
  dplyr::mutate(
    count = coll_counts_per_grp [, "count"], 
    df    = count - n_z_score_vars, 
    MSE   = SSE / df, 
    RMSE  = sqrt(MSE)
  ) |> 
  dplyr::select(
    group, count, df, RMSE, MSE, SSE
  )

# within sum of squares (across groups)
grp_z_ss_all <- sum(grp_z_ss$SSE, na.rm = TRUE)
```

```{r}
#| label: tbl-grp-z-ss
#| tbl-cap: "Point-center distances per group"

grp_z_ss |> knitr::kable(
  caption = "Point-center distances per group", 
  digits = 1
)
```

The columns of the table are as follows.

-   group: the label of each group
-   count: the number of observations (data rows)
-   df: count minus the number of numeric variables (`r n_z_score_vars`)
-   RMSE: $\surd{\text{MSE}}$
-   MSE: SSE / df
-   SSE: sum of squared point-center distances

The RMSE (root mean squared error) is a measure of the typical length of the point-center distances within each group

The sum of SSE across groups (the "total within-group SSE") turns out to be `r as.integer(round(grp_z_ss_all))`. This value serves as a reference point for other groupings (clusterings) of observations.

## K-means Clustering

We'll start our discussion of clustering using a function in the R `stats` package, namely `stats::kmeans()`. The function requires us to specify either the desired number $K$ of clusters (groups) of observations to be formed, or else to provide an initial set of $K$ cluster centers. If we merely provide the desired number $K$ of clusters, `kmeans()` randomly selects $K$ data rows (observations) as the initial set of cluster centers.

Given a set of cluster centers the algorithm iteratively searches for a better set, meaning a set having a smaller "within-cluster sum of squares". In broad terms, this is done in two steps.

1.  **Assign:** the algorithm assigns each data point to the nearest center (in Euclidean distance) thereby partitioning the data points into $K$ clusters.
2.  **Update:** For each cluster of data points, the algorithm calculates a new cluster center, namely, the mean vector of the data points within the cluster. This yields a new set of cluster centers.

**Convergence criteria:** For each cluster the algorithm adds up the (point, center) squared distances. Those $K$ sums of squares per cluster are summed across clusters to form the "total within-cluster sum of squares (WCSS)" (previously referred to as the "total within-group SSE"). In addition the algorithm records the grand mean vector, the average across the entire data set and forms a weighted sum of the squared (cluster-center, grand-mean) Euclidean distances, which is called the "between-cluster sum of squares (BCSS)".[^BCSS-WCSS] The algorithm concludes the search when cluster-membership no longer changes, or when the decrease in WCSS is sufficiently small.

[^BCSS-WCSS]: The sum of squared (point, grand-mean) distances does not depend on any grouping (clustering) of the observations, and is called the "total sum of squares". The between-cluster sum of squares (BCSS) equals the total sum of squares minus the within-cluster sum of squares. BCSS is also equal to a weighted sum of (cluster-center, grand-mean) squared distances.

### Results for specified initial cluster centers

We now provide the `kmeans()` function with cluster centers in the form of the group means calculated above.[^k-means]

[^k-means]: `kmeans()` expects the cluster centers to be expressed as a matrix, in which each *row* represents a single cluster center and each column represents a "coordinate" of the cluster centers corresponding to one of the numeric variables (columns) of the data matrix.

The resulting clusters are shown in the scatter diagrams below, where for purposes of comparison we have projected the clusters onto the grouping variables we previously adopted.

```{r}
#| label: m-kmeans-ctrs

# to ensure reproducibility, set random seed
set.seed(42) # nod to Hitchhiker's Guide to the Galaxy

# use college_z_means_per_grp as initial cluster centers
m_kmeans_ctrs <- college_z_scores |>
  # remove grp_lbl
  dplyr::select(- grp_lbl) |>
  stats::kmeans(
    centers = coll_z_means_per_grp |>
      # remove grp_lbl
      dplyr::select(- grp_lbl)
  )

# record total within-cluster sum of squares
km_wcss <- m_kmeans_ctrs$ tot.withinss
```

```{r}
#| label: college-km-ctrs

# # assign observations to km clusters
college_km_ctrs <- college_z_scores |>
  dplyr::mutate(cluster = m_kmeans_ctrs$ cluster)
```

```{r}
#| label: fig-top-10-expend-kmc
#| fig-cap: "(`Top10perc`, `Expend`): clusters identified by kmeans()"

g_top_10_Expend_kmc <- college_km_ctrs |>
  mutate(cluster = as_factor(cluster)) |>
  ggplot(mapping = aes(
    x = Top10perc, y = Expend,
    colour = cluster, shape = cluster
  )) +
  scale_shape_manual(values = 0:3) +
  geom_point() +
  labs(
    title = "(Top10perc, Expend)",
    subtitle = "clusters identified by kmeans()"
  )
g_top_10_Expend_kmc
```

### Manual grouping versus K-means

The figure above shows that the `kmeans()` clusters are distinct from but similar to the groups formed above whose mean values served as initial cluster centers. Here are the number of observations belonging to each combination of $g_{x,y}$ group and `kmeans()` cluster.

```{r}
#| label: tbl-xtabs-cluster-grp
#| tbl-cap: "Overlap: binary groups v kmeans clusters"

xtabs_cluster_grp <- xtabs(
  data    = college_km_ctrs, 
  formula = ~ cluster + grp_lbl
)
xtabs_cluster_grp
```

Some clusters align well with manual groups (e.g., clusters largely contained within a single group), while others split across groups. This comparison helps us understand whether the algorithm is finding structure similar to what we might construct manually.

### Jaccard similarity

The table above provides in more detail the similarities and differences between the initial (`grp_lbl`) and final (`cluster`) clusters from `kmeans()`. The table can be represented and summarized by [Jaccard's measure](https://en.wikipedia.org/wiki/Jaccard_index) of the similarity of two finite sets, $(A, B)$, namely:

$$
\begin{align}
  J(A, B) &= \frac{\left| A \cap B \right|}{\left| A \cup B \right|}
\end{align}
$$ {#eq-Jaccard-similarity-measure}

Letting $(A, B)$ denote the respective (row, column) of each cell in the above table, we obtain the following table of Jaccard similarity measures of each (`cluster`, `grp_lbl`) combination.

```{r}
#| label: jaccard-cluster-grp-lst

jaccard_cluster_grp_lst <- xtabs_cluster_grp |> 
  xtabs_to_jaccard()
```

```{r}
#| label: jaccard-cluster-grp-wide

jaccard_cluster_grp_wide <- 
  jaccard_cluster_grp_lst [["xt_sim_long"]] |> 
  dplyr::select(cluster, grp_lbl, jaccard) |> 
  tidyr::pivot_wider(
    names_from = "grp_lbl", 
    values_from = "jaccard"
  )
```

```{r}
#| label: tbl-jaccard-cluster-grp-wide
#| tbl-cap: "Jaccard similarity: binary groups v kmeans clusters"

jaccard_cluster_grp_wide |> 
  knitr::kable(
    caption = "Jaccard similarity measures", 
    digits = 2
  )
```

```{r}
#| label: jaccard-cluster-grp-avg

jaccard_cluster_grp_avg <- jaccard_cluster_grp_lst [[2]]
```

The weighted average (weighted by cell count) of these similarity values is `r round(jaccard_cluster_grp_avg, 2)`.

The Jaccard index is useful for comparing clusterings:

- **Compare K-means runs:** Are different random starts converging to similar solutions?
- **Compare to known groupings:** Does the algorithm recover expected structure?
- **Assess stability:** How much does the clustering change with different parameter choices?

### Distribution of point-center distances

The two figures below show the range of point-center distances for the original grouping and for the `kmeans()` clustering. In the latter case, points are closer to the mean of the cluster, as can be seen by comparing the scale of the two figures. The total within-cluster sum of squares is `r as.integer(round(km_wcss))` for `kmeans()` compared to `r as.integer(round(grp_z_ss_all))` for the original grouping.

```{r}
#| label: group-norms-lst

group_norms_lst <- college_km_ctrs |> 
  dplyr::select(- cluster) |> 
  pt_ctr_dist(grp_name = "grp_lbl")
```

```{r}
#| label: cluster-norms-lst

cluster_norms_lst <- college_km_ctrs |> 
  dplyr::select(- grp_lbl) |> 
  pt_ctr_dist(grp_name = "cluster")
```

```{r}
#| label: fig-grp-norms
#| fig-cap: "Initial binary grouping: point-center distances"

group_norms_lst$ grp_norms |> 
  ggplot(mapping = aes(y = norm)) + 
  geom_boxplot() + 
  facet_grid(cols = vars(grp_lbl)) + 
  labs(
    title = "Initial groups: point-center distances"
  )
```

```{r}
#| label: fig-cluster-norms
#| fig-cap: "K-means clusters: point-center distances"

cluster_norms_lst$ grp_norms |> 
  ggplot(mapping = aes(y = norm)) + 
  geom_boxplot() + 
  facet_grid(cols = vars(cluster)) + 
  labs(
    title = "K-means clusters: point-center distances", 
    subtitle = "(initial centers from initial grouping)"
  )
```

### Results for random initial cluster centers

Let's run `kmeans()` again, this time specifying only the desired number $K$ of clusters. The function will then select $K$ rows of the data at random as the initial set of cluster centers. In any case, no matter how the initial cluster centers may be selected, that selection influences the subsequent iterative formation of clusters.

To make the algorithm more robust we specify parameter `nstart`, the number times to run `kmeans()`, each using a distinct random selection of data rows as initial cluster centers. For each run the algorithm computes WCSS (the total within-cluster sum of squares). Once `nstart` runs have been completed, the run having the smallest value of WCSS is reported to the user. A rule of thumb is to set `nstart` to a value between 20 and 50.

We now run `kmeans()` again, setting $K = 4$ and `nstart = 25`. For purposes of comparison with previous results, here is a projection of the resulting 4 clusters onto a (Top10perc, Expend) scatter diagram.

The key code for running K-means with multiple random starts:

```{r}
#| label: kmeans-code-demo
#| echo: true
#| eval: false

# K-means with multiple random starts for robustness
kmeans_result <- college_z_scores |>
  dplyr::select(-grp_lbl) |>
  stats::kmeans(
    centers = 4,      # number of clusters
    nstart  = 25      # number of random starts
  )

# Extract results
kmeans_result$cluster       # cluster assignments
kmeans_result$centers       # cluster centroids  
kmeans_result$tot.withinss  # total within-cluster SS
```

```{r}
#| label: m-kmeans-nstart

# to ensure reproducibility, set random seed
set.seed(42)

m_kmeans_nstart <- college_z_scores |>
  # remove grp_lbl
  dplyr::select(- grp_lbl) |>
  stats::kmeans(
    centers = 4, 
    nstart = 25
  )
```

```{r}
#| label: fig-top-10-expend-nstart
#| fig-cap: "Clusters identified from repeated random cluster centers"

g_top_10_Expend_nstart <- college_z_scores |>
  mutate(
    cluster = m_kmeans_nstart$ cluster |> as_factor()
  ) |>
  ggplot(mapping = aes(
    x = Top10perc, y = Expend,
    colour = cluster, shape = cluster
  )) +
  scale_shape_manual(values = 0:3) +
  geom_point() +
  labs(
    title = "(Top10perc, Expend)",
    subtitle = "identified by kmeans cluster (nstart = 25)"
  )
g_top_10_Expend_nstart
```

Interestingly, we again see strong separation of clusters associated with the values of `Top10perc` and `Expend`.

The table below shows the number of observations in each combination of the original grouping and the new `kmeans()` clusters.

```{r}
#| label: xtabs-nstart-grp

xtabs_nstart_grp <- xtabs(
  data    = college_z_scores |> 
  dplyr::mutate(
    cluster = m_kmeans_nstart$ cluster |> as_factor()
  ), 
  formula = ~ cluster + grp_lbl
)
xtabs_nstart_grp
```

As with the previous `kmeans()` clusters, we again see two of the clusters well aligned with groups $g_{00}$ and $g_{11}$, respectively.

The similarity of the current clusters with the previous clusters prompts us to examine the number of observations in each pair (old, new) clusters, distinguishing the two as (cl_group, cl_nstart).

```{r}
#| label: tbl-xtabs-cl-group-nstart
#| tbl-cap: "kmeans() clusters formed with and without `nstart`"

xtabs_cl_group_nstart  <- xtabs(
  data    = college_z_scores |> 
  dplyr::mutate(
    cl_group  = m_kmeans_ctrs$ cluster |> as_factor(), 
    cl_nstart = m_kmeans_nstart$ cluster |> as_factor()
  ), 
  formula = ~ cl_group + cl_nstart
)
xtabs_cl_group_nstart
```

We see that the new clusters are essentially a relabeling of the previous clusters. The figure below of new point-center distances is essentially the same as the previous figure with the first two clusters swapped. In fact the total within-cluster sum of squares is unchanged from the previous clustering.

```{r}
#| label: nstart-norms-lst

nstart_norms_lst <- college_z_scores |> 
  dplyr::select(- grp_lbl) |> 
  dplyr::mutate(
    cluster = m_kmeans_nstart$ cluster |> as_factor()
  ) |> 
  pt_ctr_dist(grp_name = "cluster")
```

```{r}
#| label: fig-nstart-norms
#| fig-cap: "Revised point-center distances (nstart = 25)"

g_nstart_norms <- nstart_norms_lst$ grp_norms |> 
  ggplot(mapping = aes(y = norm)) + 
  geom_boxplot() + 
  facet_grid(cols = vars(cluster)) + 
  labs(
    title    = "K-means clusters: point-center distances", 
    subtitle = "(nstart = 25)"
  )
g_nstart_norms
```

### Which value of $K$?

So far in our application of `kmeans()` to the college data we have set $K = 4$. How would the results change with different choices for the value of $K$? Breaking the observations into smaller but more numerous clusters should in general reduce the total within-cluster sum of squares (WCSS), but we can also expect to reach a point of diminishing returns.

For the college data, the figure below shows how WCSS decreases as $K$ increases.

The code to generate the elbow plot:

```{r}
#| label: elbow-code-demo
#| echo: true
#| eval: false

# Compute WCSS for different values of K
wcss_by_k <- tibble::tibble(k = 2:12) |> 
  dplyr::mutate(
    wcss = purrr::map_dbl(k, ~ {
      kmeans(college_z, centers = .x, nstart = 20)$tot.withinss
    })
  )

# Plot WCSS vs K (elbow plot)
ggplot(wcss_by_k, aes(x = k, y = wcss)) +
  geom_line() + geom_point() +
  labs(x = "Number of clusters (K)", y = "Within-cluster SS")
```

```{r}
#| label: mk-tbl

# run kmeans on college data for successive values of k

# to ensure reproducibility, set random seed
set.seed(42)

k_max <- 12L
nstart_per_k <- 20L
mk_tbl <- tibble::tibble()

coll_z <- college_z_scores |> 
  # use only z-scores of original numeric variables
  dplyr::select(- grp_lbl)

for (k in 2:k_max) {
  m_tmp <- coll_z |> 
    kmeans(centers = k, nstart = nstart_per_k)
  mk_tbl <- mk_tbl |> 
    dplyr::bind_rows(
      tibble::tibble(
        k    = k, 
        wcss = m_tmp$ tot.withinss)
    )
}
```

```{r}
#| label: fig-mk-tbl
#| fig-cap: "Within-cluster sum of squares versus number of clusters"

g_mk_tbl <- mk_tbl |> 
  ggplot2::ggplot(mapping = aes(
    x = k, y = wcss
  )) + 
  ggplot2::scale_x_continuous(
    name = "K", 
    breaks = seq(2, 12, 2)) + 
  ggplot2::geom_line() + 
  ggplot2::geom_point() +
  ggplot2::labs(
    title = "WCSS versus K for college z-scores", 
    subtitle = "(nstart = 20)")
g_mk_tbl
```

In general, smaller values of $K$ are more interpretable. Based on the figure above we might set $K$ to be 4 or 5. This is a judgement call that depends on the circumstances and goals of one's project.

::: {.callout-note}
## The elbow method

Look for an "elbow" in the plot—a point where the rate of decrease in WCSS sharply levels off. Before the elbow, adding clusters captures real structure. After the elbow, we see diminishing returns.

The elbow is often subjective; domain knowledge and interpretability should also guide your choice of $K$.
:::

## Interpreting the Clusters

Clustering is not complete until we interpret what the clusters mean. The algorithm found groups—now we ask what characterizes each one.

```{r}
#| label: cluster-profiles

# Add cluster assignments to original data
college_with_clusters <- college |> 
  dplyr::mutate(cluster = factor(m_kmeans_nstart$cluster))

# Compute cluster profiles
cluster_profiles <- college_with_clusters |> 
  dplyr::group_by(cluster) |> 
  dplyr::summarise(
    n = n(),
    Pct_Private = mean(Private == "Yes") * 100,
    Top10perc = mean(Top10perc),
    Expend = mean(Expend),
    Grad.Rate = mean(Grad.Rate),
    F.Undergrad = mean(F.Undergrad),
    Outstate = mean(Outstate),
    .groups = "drop"
  )
```

```{r}
#| label: tbl-cluster-profiles
#| tbl-cap: "Cluster profiles: mean values for selected variables"

cluster_profiles |> 
  knitr::kable(
    digits = 0,
    col.names = c("Cluster", "n", "% Private", "Top10%", 
                  "Expend", "Grad Rate", "Enrollment", "Outstate")
  )
```

Review of these profiles and additional variable summaries suggests the following interpretations:

::: {.callout-tip}
## Cluster interpretation (example)

The specific cluster numbers may vary depending on random seed, but typical patterns in college data include:

- **High-resource selective schools:** High `Top10perc`, high `Expend`, high `Outstate`, mostly private—likely elite private institutions
- **Large public universities:** High `F.Undergrad`, lower `Expend` per student, lower `Outstate`—flagship state universities
- **Regional/teaching-focused:** Moderate values across metrics, mix of private and public
- **Open-access institutions:** Lower `Top10perc`, lower `Expend`, lower graduation rates

Does this match your domain knowledge about the US higher education landscape?
:::

This interpretive step is essential. Without it, clustering remains a mathematical exercise. With it, clustering becomes a tool for generating insights and hypotheses.

The interpretive step exemplifies the dual aims of data analysis introduced in the book's preface. The algorithm provides decision support: it assigns observations to groups. But the assignment is useful only when accompanied by understanding: what characterizes each group, and why might these distinctions matter? Clustering without interpretation is computation; clustering with interpretation is analysis.

## Other Clustering Algorithms

Clustering methods are an active area of research. The K-means algorithm, first introduced in the 1930's, has the longest history. More recent methods can be categorized along the following lines.

-   **Soft versus Hard:** "Hard" clustering assigns each data point to a single cluster. "Soft" clustering assigns to each point a finite probability distribution of cluster membership.

-   **Agglomerative versus Divisive:** methods define the initial partition of the data into clusters in opposite ways. Agglomerative clustering initially defines each data point to be a cluster, and then merges clusters. Divisive clustering initially defines the entire data set to be a single cluster, and then breaks up this and subsequent clusters into smaller clusters.

-   **Hierarchical or not:** Instead of a partition, hierarchical methods create a hierarchy of parent (superset) and child (subset) clusters.

-   **Centroid-based versus Density-based:** K-means and other centroid-based methods identify a cluster of points with the mean, median, or other central measure of the points in the cluster. Density-based methods find areas having a high density of points.

Some methods of note are [DBSCAN](https://en.wikipedia.org/wiki/DBSCAN) (Density Based Spatial Clustering of Applications with Noise) and [HDBSCAN](https://hdbscan.readthedocs.io/en/latest/how_hdbscan_works.html) (a hierarchical version of DBSCAN).

### Distance Measures

Both K-means and density-based methods like DBSCAN depend on a distance (or dissimilarity) measure to determine which observations are "close." The default is Euclidean distance (the $L_2$ norm):

$$
d_2(\mathbf{x}, \mathbf{y}) = \sqrt{\sum_{j=1}^{p} (x_j - y_j)^2}
$$

Other common choices include Manhattan distance ($L_1$), which sums absolute differences and is less sensitive to outliers in individual dimensions:

$$
d_1(\mathbf{x}, \mathbf{y}) = \sum_{j=1}^{p} |x_j - y_j|
$$

and Chebyshev distance ($L_\infty$), which considers only the largest single-dimension difference:

$$
d_\infty(\mathbf{x}, \mathbf{y}) = \max_{j} |x_j - y_j|
$$

The choice of distance measure affects cluster shapes and which observations are grouped together. Euclidean distance implicitly assumes that features are on comparable scales—reinforcing why standardization (z-scores) is typically applied before clustering. In high-dimensional settings, all distance measures face the "curse of dimensionality," a phenomenon we examine in @sec-reduce-dim.

### Example: DBSCAN

Unlike K-means, DBSCAN does not require specifying the number of clusters in advance. Instead, it finds clusters based on density—regions where points are packed closely together. Points in sparse regions are labeled as noise.

```{r}
#| label: dbscan-demo
#| echo: true

# DBSCAN clustering on standardized college data
dbscan_result <- college_z_scores |> 
  dplyr::select(-grp_lbl) |> 
  dbscan::dbscan(eps = 2.5, minPts = 10)

# Cluster assignments (0 = noise)
table(dbscan_result$cluster)
```

```{r}
#| label: fig-dbscan
#| fig-cap: "DBSCAN clusters (0 = noise points)"

college_z_scores |> 
  dplyr::mutate(cluster = factor(dbscan_result$cluster)) |> 
  ggplot2::ggplot(ggplot2::aes(
    x = Top10perc, y = Expend, 
    color = cluster, shape = cluster
  )) +
  ggplot2::geom_point(alpha = 0.6) +
  ggplot2::labs(
    title = "DBSCAN clustering",
    subtitle = "eps = 2.5, minPts = 10"
  )
```

DBSCAN's key parameters are `eps` (the neighborhood radius) and `minPts` (minimum points to form a dense region). Points not belonging to any dense region are classified as noise (cluster 0). This can be useful for identifying outliers or when clusters have irregular shapes.

::: {.callout-note}
## When to use DBSCAN vs K-means

- **K-means:** Works well when clusters are roughly spherical and similar in size. Requires specifying $K$.
- **DBSCAN:** Works well for clusters of arbitrary shape, can identify outliers, doesn't require specifying $K$. Sensitive to the choice of `eps` and `minPts`.
:::

## Exercises {#sec-clustering-exercises}

1.  **(College Data Exploration)** Make your own copy of the `ISLR2::College` data and explore it. Consider: How many private vs. public schools are in the data? What is the range of `Expend` (expenditure per student) compared to `Grad.Rate`? Which numeric variables appear to be correlated (use `GGally::ggpairs()` on a subset)? Based on your domain knowledge, what groupings would you expect to find among colleges—elite vs. non-elite, large vs. small, research-focused vs. teaching-focused? What questions about colleges might clustering help answer?

2.  **(Iris Data)** Submit R command `help("iris")` to learn about the measurements of iris flowers available in R as `datasets::iris`. Summarise these measurements in a graph or table, for example by applying the `GGally::ggpairs()` function.

3.  **(Iris K-means)** Apply the K-means algorithm to the iris measurements. How do the computed clusters compare to the grouping by variable `Species`? Count the number of observations corresponding to each combination of cluster and `Species`, for example by using R function `stats::xtabs()`.

4.  **(Wine Quality Data, optional)** The `eda4mldata::wine_quality` dataset contains chemical measurements of wines. Apply K-means clustering to the numeric variables and interpret the resulting clusters. Do they correspond to quality ratings or wine color?

## Connections to Machine Learning

Clustering connects to supervised learning in several ways:

::: {.callout-note}
## ML connections

- **Preprocessing:** Standardization (z-scores) is essential for many ML algorithms, not just clustering.
- **Feature engineering:** Cluster membership can become a derived feature for supervised models.
- **Model selection:** The elbow heuristic parallels cross-validation and regularization—balancing fit against complexity.
- **Loss minimization:** K-means minimizes WCSS, a precursor to the loss functions used in supervised learning.
:::

## Resources

[An Introduction to Statistical Learning](https://www.statlearning.com/) by James, Witten, Hastie, & Tibshirani

[K-means clustering with tidy data principles – tidymodels](https://www.tidymodels.org/learn/statistics/k-means/)
 
[tidyclust::k_means](https://tidyclust.tidymodels.org/articles/k_means.html)

[Cluster analysis - Wikipedia](https://en.wikipedia.org/wiki/Cluster_analysis)

[HDBSCAN with the dbscan package](https://cran.r-project.org/web/packages/dbscan/vignettes/hdbscan.html)
