# Graph Theory for Machine Learning {#sec-graph-theory}

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo    = FALSE,
  error   = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r}
#| label: cran_libraries

library(here)
library(igraph)
library(knitr)
library(latex2exp)
library(tidyverse)
```

```{r}
#| label: local-libraries

library(eda4mldata)

```

```{r}
#| label: local_source

# stabilize simulated data
source(here::here("code", "eda4ml_set_seed.R"))

```

------------------------------------------------------------------------

## Introduction

In previous chapters we have seen that understanding and decision support are complementary: knowing *why* a pattern exists enables us to predict *what* will happen next. For network data this stereoscopic view of theory and application is especially useful.

Graphs arise whenever entities are connected: people linked by friendship, proteins linked by interaction, web pages linked by hyperlinks, concepts linked by semantic relationships. The mathematical abstraction is simple—*nodes* and *edges*—yet this structure appears across domains so diverse that graph theory has become a common language for network science, social science, biology, and machine learning.

The connections encoded in a graph are not merely data to be stored. They are *information*: information about influence, about similarity, about pathways and dependencies. The structure of a social network reveals who influences whom. The structure of a citation network reveals how ideas propagate. The structure of a prerequisite graph reveals what must be learned before what. To analyze a graph is to extract this information—to move from raw connectivity to insight.

### Two Perspectives, One Structure

Graph analysis serves two complementary aims:

- **Understanding**: What structure exists? What communities or clusters are present? Which nodes are central or influential? How does information flow through the network?

- **Prediction**: What will happen next? Which nodes will connect? What should we recommend? What is the shortest path from here to there?

These perspectives are not opposed. The same structural properties that reveal *why* a network behaves as it does also enable prediction of *what* it will do. Communities exist because people preferentially associate with others who are similar, a homophily that enables us to predict future connections. Nodes with high betweenness centrality control information flow and are more likely to broker new relationships.

This duality runs throughout graph analysis. We detect communities to understand social structure, and we use that structure to predict group behavior. We measure node similarity to understand relationships, and we use that similarity for recommendation. Structure is created by the forging of links between nodes, and is therefore predictive of new links.

### Three Examples

We develop graph theory through three examples of increasing complexity:

| Example | Domain | Key Concepts |
|---------|--------|--------------|
| Zachary's Karate Club | Social network | Community detection, centrality |
| MovieLens | Recommendations | Bipartite graphs, projection |
| LearningGraph | Knowledge graph | Typed nodes, DAGs, path queries |

Each example illuminates different aspects of graph structure and different analytical questions. The Karate Club shows how network topology predicts group behavior. MovieLens shows how bipartite structure enables recommendation. The LearningGraph shows how typed nodes and labeled edges enable richer reasoning about learning pathways.

### Chapter Roadmap

This chapter proceeds as follows:

**Examples first**: We begin with data—social networks, recommendation systems, knowledge graphs—examining each from both structural and predictive perspectives. These examples build intuition for what graph analysis can reveal.

**Fundamental concepts**: We define the basic vocabulary: nodes, edges, directed and undirected graphs, weighted edges, paths, connectivity. We introduce matrix representations—adjacency, degree, Laplacian—that connect graph structure to linear algebra.

**Measuring importance**: We develop centrality measures that quantify which nodes matter. Different questions lead to different measures: degree centrality counts connections; betweenness centrality measures control of information flow; PageRank propagates importance through the network.

**Finding structure**: We examine community detection—algorithms that partition a graph into densely connected subgroups. Modularity provides a quality measure; spectral methods connect community structure to the eigenvectors of the graph Laplacian.

**Graphs and machine learning**: We show how graph structure becomes features for prediction tasks. Link prediction asks which unconnected nodes will connect; recommendation asks what to suggest based on network neighbors. We briefly introduce graph neural networks, which learn representations directly from graph structure.

Throughout, we maintain the dual perspective: structural analysis and prediction are two views of the same underlying reality.


## Zachary's Karate Club

In the early 1970s, sociologist Wayne Zachary spent two years observing a university karate club.[^zachary] He recorded who interacted with whom *outside* the formal club setting—who practiced together, who socialized, who sought each other out. The result was a social network of 34 members and 78 connections.

[^zachary]: See @Zachary_1977.

During Zachary's observation period, a conflict arose between the club's instructor (Mr. Hi, node 1) and its president (John A, node 34). The dispute escalated. Eventually, the club split into two factions—some members following Mr. Hi to form a new club, others remaining with John A.

Zachary had recorded the social network *before* the split occurred. This created a natural experiment: could the network structure predict which faction each member would join?

### The Network

@fig-karate-network displays the Karate Club network. Node colors indicate the actual faction each member joined after the split—red for Mr. Hi's group, blue for John A's group. The two leaders are shown as larger nodes.

```{r}
#| label: fig-karate-network
#| fig-cap: "Zachary's Karate Club network. Colors indicate faction membership after the split; larger nodes mark the two leaders."

karate <- igraph::make_graph("Zachary")

# Set up node attributes
igraph::V(karate)$name <- 1:igraph::vcount(karate)

# Known faction membership (ground truth)
faction <- c(1,1,1,1,1,1,1,1,2,2,1,1,1,1,2,2,1,1,2,1,2,1,2,2,2,2,2,2,2,2,2,2,2,2)
igraph::V(karate)$faction <- faction
igraph::V(karate)$color <- ifelse(faction == 1, "#E41A1C", "#377EB8")

# Highlight the two leaders
igraph::V(karate)$size <- 15
igraph::V(karate)$size[1] <- 25
igraph::V(karate)$size[34] <- 25

set.seed(42)
plot(karate, 
     layout = igraph::layout_with_fr(karate),
     vertex.label = igraph::V(karate)$name,
     vertex.label.color = "white",
     vertex.label.cex = 0.8,
     edge.color = "gray70",
     main = "")
legend("bottomleft", 
       legend = c("Mr. Hi's faction", "John A's faction"),
       fill = c("#E41A1C", "#377EB8"),
       bty = "n")
```

Visual inspection already suggests structure. The network separates into two densely connected regions with relatively few edges between them. Node 1 (Mr. Hi) and node 34 (John A) are clearly hubs of their respective groups. The social topology foreshadows the coming split.

Zachary's result was striking: the network correctly predicted 33 of 34 members' choices. The single "misclassified" member (node 9) had stronger ties to Mr. Hi's group but followed a close friend to John A's faction. Even this exception proves the rule—the prediction failed because of a social tie, just not the ones the algorithm counted.

### Centrality: Who Matters?

The Karate Club raises a fundamental question: which nodes are important? Different notions of importance lead to different measures.

**Degree centrality** is the simplest: count the connections. A node with many edges is central in the sense of having many direct relationships. In the Karate Club, Mr. Hi (node 1) has degree 16—he interacts with nearly half the club members. John A (node 34) has degree 17.

But raw connection count may miss something. Consider node 3, who interacts with both leaders. This node may be important not because of how many connections it has, but because of *where* it sits in the network—bridging the two factions.

**Betweenness centrality** captures this idea. A node has high betweenness if it lies on many shortest paths between other nodes. Formally, for node $v$:

$$
\gamma(v) = \sum_{s \neq v \neq t} \frac{\sigma_{st}(v)}{\sigma_{st}}
$$

where $\sigma_{st}$ is the number of shortest paths from node $s$ to node $t$, and $\sigma_{st}(v)$ is the number of those paths that pass through $v$. A node with high betweenness is a *bridge*—removing it would fragment the network or force information to take longer routes.

@fig-karate-betweenness shows the Karate Club with nodes sized by betweenness centrality.

```{r}
#| label: fig-karate-betweenness
#| fig-cap: "Karate Club with nodes sized by betweenness centrality. Mr. Hi (node 1) has the highest betweenness—he bridges many pairs of members."

# Reset colors
igraph::V(karate)$color <- ifelse(faction == 1, "#E41A1C", "#377EB8")

# Size by betweenness
betw <- igraph::betweenness(karate, normalized = TRUE)
igraph::V(karate)$size <- 8 + 35 * betw

set.seed(42)
plot(karate,
     layout = igraph::layout_with_fr(karate),
     vertex.label = igraph::V(karate)$name,
     vertex.label.color = "white",
     vertex.label.cex = 0.7,
     edge.color = "gray70",
     main = "")
```

Mr. Hi has the highest betweenness. As the instructor, he connects to members who might not otherwise interact—beginners who come for lessons, advanced students who assist him, social members who attend club events. He is not just well-connected; he is a *connector*. John A, despite having slightly higher degree, has lower betweenness—his connections are more concentrated within his faction.

This distinction matters. When the club split, Mr. Hi's bridging position meant that his departure fragmented more relationships than John A's would have. Betweenness centrality identified this structural importance.

### Community Detection

The visual structure of @fig-karate-network suggests two groups. Can an algorithm recover this partition without knowing the ground truth?

Community detection algorithms seek to partition a graph into subgroups that are internally dense and externally sparse. The intuition is that communities represent meaningful clusters—people who preferentially interact with each other, proteins that function together, web pages on related topics.

The **Louvain algorithm** is a widely used method that greedily optimizes *modularity*—a measure comparing the actual density of edges within groups to what would be expected in a random graph. High modularity indicates genuine community structure, not just arbitrary partitioning.

@fig-karate-louvain shows the communities detected by the Louvain algorithm.

```{r}
#| label: fig-karate-louvain
#| fig-cap: "Communities detected by the Louvain algorithm, which had no knowledge of the actual split."

# Reset size
igraph::V(karate)$size <- 15
igraph::V(karate)$size[1] <- 25
igraph::V(karate)$size[34] <- 25

# Detect communities
set.seed(42)
comm <- igraph::cluster_louvain(karate)

# Color by detected community
comm_colors <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3")
igraph::V(karate)$color <- comm_colors[igraph::membership(comm)]

set.seed(42)
plot(karate,
     layout = igraph::layout_with_fr(karate),
     vertex.label = igraph::V(karate)$name,
     vertex.label.color = "white",
     vertex.label.cex = 0.8,
     edge.color = "gray70",
     main = "")
```

The algorithm, knowing nothing about the dispute or its outcome, finds a partition closely matching the actual split. The modularity score is `r round(igraph::modularity(comm), 3)`—well above the threshold of 0.3 typically indicating significant community structure.

This is not a coincidence. The algorithm detects communities because the network *has* community structure.  People who are similar tend to associate with one another. The same homophily that created the dense within-group connections also predicted which faction members would join. The algorithm did not foresee the future; it revealed structure that was already present.

### Prediction Meets Understanding

The Karate Club illustrates the dual perspective that runs through graph analysis.

From the **understanding** perspective, the network structure reveals social organization. Communities exist because of homophily—the tendency for similar people to associate. Centrality measures identify structural roles: hubs, bridges, peripheral members. The network is a map of social relationships.

From the **prediction** perspective, that same structure forecasts behavior. When forced to choose, members followed their network ties. The topology that existed before the conflict predicted the outcome of the conflict. Community detection was not just descriptive—it was predictive.

This connection is not magical. Structure is predictive because structure reflects mechanism. The network encoded the social relationships that would determine people's choices. Community detection worked because the algorithm detected the same structure that drove the actual decision-making process.

This is the power of graph analysis: **structure reveals mechanism, and mechanism enables prediction**. We will see this pattern repeated in the examples that follow.
