## Zachary's Karate Club (continued)

[... earlier content unchanged through line 225 ...]

The algorithm, knowing nothing about the dispute or its outcome, finds a partition closely matching the actual split. The modularity score is `r round(igraph::modularity(comm), 3)`—well above the threshold of 0.3 typically indicating significant community structure.

This is not a coincidence. The algorithm detects communities because the network *has* community structure. People who are similar tend to associate with one another. The same homophily that created the dense within-group connections also predicted which faction members would join. The algorithm did not foresee the future; it revealed structure that was already present.

The Karate Club is a small example, but its lesson scales. The network structure that existed before the conflict predicted the outcome of the conflict—not because algorithms are clairvoyant, but because the topology encoded the social relationships that would determine people's choices.

Our next example shifts from social networks to recommendations, and from a single graph to a bipartite structure connecting users and items.


## MovieLens: Bipartite Graphs and Recommendation

The MovieLens dataset contains millions of movie ratings from thousands of users.[^movielens] For our purposes, we use a small subset: 100 users, 20 movies, and 697 ratings on a 1–5 scale.

[^movielens]: See @Harper_Konstan_2015. The GroupLens research lab at the University of Minnesota has maintained MovieLens since 1997.

This data has a natural graph structure, but not the same kind we saw in the Karate Club. There, nodes were all of one type (people) and edges represented a single relationship (social interaction). Here, we have two distinct types of nodes—users and movies—and edges connect nodes of *different* types. A user rates a movie; movies do not rate each other; users do not rate users.

This is a **bipartite graph**: a graph whose nodes partition into two disjoint sets, with edges only between sets, never within.

### The Bipartite Structure

@fig-movielens-bipartite shows a portion of the MovieLens bipartite graph. Users appear on the left, movies on the right. An edge connects user $u$ to movie $m$ if user $u$ rated movie $m$.

```{r}
#| label: fig-movielens-bipartite
#| fig-cap: "A bipartite graph connecting users to the movies they rated. Users appear on the left, movies on the right."

# Create a small example bipartite graph
set.seed(123)

# Sample users and movies for visualization
users <- paste0("U", 1:8)
movies <- c("Toy Story", "Pulp Fiction", "Forrest Gump", "Silence of the Lambs", 
            "Star Wars", "Schindler's List")

# Create edges (user-movie ratings)
edges <- tibble::tribble(
  ~user, ~movie,
  "U1", "Toy Story",
  "U1", "Star Wars",
  "U1", "Forrest Gump",
  "U2", "Pulp Fiction",
  "U2", "Forrest Gump",
  "U2", "Silence of the Lambs",
  "U3", "Toy Story",
  "U3", "Star Wars",
  "U4", "Pulp Fiction",
  "U4", "Schindler's List",
  "U5", "Toy Story",
  "U5", "Forrest Gump",
  "U5", "Star Wars",
  "U6", "Silence of the Lambs",
  "U6", "Schindler's List",
  "U7", "Star Wars",
  "U7", "Forrest Gump",
  "U8", "Pulp Fiction",
  "U8", "Silence of the Lambs"
)

# Build bipartite graph
g_bip <- igraph::graph_from_data_frame(edges, directed = FALSE)

# Set node types
igraph::V(g_bip)$type <- igraph::V(g_bip)$name %in% movies

# Layout: users on left, movies on right
coords <- matrix(0, nrow = igraph::vcount(g_bip), ncol = 2)
user_nodes <- which(!igraph::V(g_bip)$type)
movie_nodes <- which(igraph::V(g_bip)$type)

coords[user_nodes, 1] <- 0
coords[user_nodes, 2] <- seq(length(user_nodes), 1, length.out = length(user_nodes))
coords[movie_nodes, 1] <- 2
coords[movie_nodes, 2] <- seq(length(movie_nodes), 1, length.out = length(movie_nodes))

# Colors
igraph::V(g_bip)$color <- ifelse(igraph::V(g_bip)$type, "#377EB8", "#E41A1C")

plot(g_bip,
     layout = coords,
     vertex.size = 12,
     vertex.label.cex = 0.7,
     vertex.label.color = "black",
     vertex.label.dist = ifelse(igraph::V(g_bip)$type, 2.5, -2.5),
     edge.color = "gray70",
     main = "")
legend("bottom",
       legend = c("Users", "Movies"),
       fill = c("#E41A1C", "#377EB8"),
       horiz = TRUE,
       bty = "n")
```

The bipartite structure immediately suggests a recommendation strategy. If user U1 and user U3 have both watched Toy Story and Star Wars, they may have similar tastes. If U3 has also watched a movie that U1 hasn't seen, that movie becomes a candidate recommendation for U1. The graph encodes a path: U1 → shared movie → U3 → candidate movie.

This is **collaborative filtering** expressed as graph traversal: find users similar to you (connected through shared items), then recommend items those similar users have chosen.

### Projection: From Bipartite to Similarity

The bipartite graph contains all the information, but reasoning about user similarity is easier in a different representation. We can **project** the bipartite graph onto one of its node types.

The *user projection* creates a new graph containing only user nodes. Two users are connected in the projection if they share at least one movie in the original bipartite graph. The edge weight can reflect how many movies they share—more shared movies implies stronger similarity.

@fig-user-projection illustrates this transformation.

```{r}
#| label: fig-user-projection
#| fig-cap: "User projection: the dashed edge connects users who share a movie. Users U1 and U3 both rated Star Wars (highlighted), creating a projected edge between them."

# Show the bipartite graph with highlighted shared rating and projected edge
# Focus on U1, U3, and their shared movie (Star Wars)

par(mfrow = c(1, 1))

# Rebuild for this figure
edges_highlight <- edges

g_bip2 <- igraph::graph_from_data_frame(edges_highlight, directed = FALSE)
igraph::V(g_bip2)$type <- igraph::V(g_bip2)$name %in% movies

# Identify edges to highlight: U1-Star Wars and U3-Star Wars
edge_list <- igraph::as_edgelist(g_bip2)
highlight_idx <- which(
  (edge_list[,1] == "U1" & edge_list[,2] == "Star Wars") |
  (edge_list[,1] == "U3" & edge_list[,2] == "Star Wars") |
  (edge_list[,1] == "Star Wars" & edge_list[,2] == "U1") |
  (edge_list[,1] == "Star Wars" & edge_list[,2] == "U3")
)

edge_colors <- rep("gray80", igraph::ecount(g_bip2))
edge_widths <- rep(1, igraph::ecount(g_bip2))
edge_colors[highlight_idx] <- "#E74C3C"
edge_widths[highlight_idx] <- 3

# Colors
igraph::V(g_bip2)$color <- ifelse(igraph::V(g_bip2)$type, "#377EB8", "#E41A1C")

# Layout
coords2 <- matrix(0, nrow = igraph::vcount(g_bip2), ncol = 2)
user_nodes2 <- which(!igraph::V(g_bip2)$type)
movie_nodes2 <- which(igraph::V(g_bip2)$type)

coords2[user_nodes2, 1] <- 0
coords2[user_nodes2, 2] <- seq(length(user_nodes2), 1, length.out = length(user_nodes2))
coords2[movie_nodes2, 1] <- 2
coords2[movie_nodes2, 2] <- seq(length(movie_nodes2), 1, length.out = length(movie_nodes2))

plot(g_bip2,
     layout = coords2,
     vertex.size = 12,
     vertex.label.cex = 0.7,
     vertex.label.color = "black",
     vertex.label.dist = ifelse(igraph::V(g_bip2)$type, 2.5, -2.5),
     edge.color = edge_colors,
     edge.width = edge_widths,
     main = "")

# Add projected edge as dashed line between U1 and U3
u1_idx <- which(igraph::V(g_bip2)$name == "U1")
u3_idx <- which(igraph::V(g_bip2)$name == "U3")
segments(x0 = coords2[u1_idx, 1], y0 = coords2[u1_idx, 2],
         x1 = coords2[u3_idx, 1], y1 = coords2[u3_idx, 2],
         lty = 2, lwd = 3, col = "#3498DB")

legend("bottom",
       legend = c("Users", "Movies", "Shared rating", "Projected edge"),
       col = c("#E41A1C", "#377EB8", "#E74C3C", "#3498DB"),
       pch = c(16, 16, NA, NA),
       lty = c(NA, NA, 1, 2),
       lwd = c(NA, NA, 3, 3),
       horiz = TRUE,
       bty = "n",
       cex = 0.8)
```

In the bipartite graph, U1 and U3 are not directly connected—users never connect to users. But both rated Star Wars (shown in red). The projection creates a direct edge between them (dashed blue line), representing their shared taste.

@fig-user-projection-full shows the complete user projection for our example.

```{r}
#| label: fig-user-projection-full
#| fig-cap: "The user projection graph. Edge thickness reflects the number of shared movies."

# Create user projection
g_user_proj <- igraph::bipartite_projection(g_bip, which = "false")

# Edge weights = number of shared movies (already computed by projection)
edge_weights <- igraph::E(g_user_proj)$weight

igraph::V(g_user_proj)$color <- "#E41A1C"

set.seed(42)
plot(g_user_proj,
     layout = igraph::layout_with_fr(g_user_proj),
     vertex.size = 20,
     vertex.label.cex = 0.9,
     vertex.label.color = "white",
     edge.width = edge_weights * 1.5,
     edge.color = "gray50",
     main = "")
```

The projection transforms a bipartite user-item graph into a user-user similarity network. Users who share many movies are strongly connected; users with no overlap are disconnected. This is the same structure we analyzed in the Karate Club—but derived from behavioral data rather than observed directly.

### From Structure to Recommendation

The user projection enables a simple recommendation algorithm:

1. Find the target user's neighbors in the projection (users with shared movies)
2. Weight neighbors by similarity (number of shared movies)
3. Collect movies those neighbors have rated but the target user hasn't
4. Rank candidates by weighted frequency

This is **user-based collaborative filtering** expressed as graph operations. The projection makes explicit the similarity relationships implicit in the bipartite structure.

We can equally well project onto movies rather than users. The *movie projection* connects movies that share viewers. Two movies watched by the same users are likely similar in some way—genre, tone, appeal to a particular audience. This yields **item-based collaborative filtering**: given a movie the user liked, recommend movies connected to it in the movie projection.

```{r}
#| label: fig-movie-projection
#| fig-cap: "The movie projection graph. Movies that share viewers are connected."

# Create movie projection
g_movie_proj <- igraph::bipartite_projection(g_bip, which = "true")

igraph::V(g_movie_proj)$color <- "#377EB8"

set.seed(42)
plot(g_movie_proj,
     layout = igraph::layout_with_fr(g_movie_proj),
     vertex.size = 20,
     vertex.label.cex = 0.7,
     vertex.label.color = "white",
     edge.width = igraph::E(g_movie_proj)$weight * 1.5,
     edge.color = "gray50",
     main = "")
```

### The Dual View

The MovieLens example illustrates how different graph representations serve different purposes:

| Representation | Nodes | Edges | Best For |
|----------------|-------|-------|----------|
| Bipartite | Users + Movies | Ratings | Complete information, rating prediction |
| User projection | Users | Shared movies | Finding similar users |
| Movie projection | Movies | Shared viewers | Finding similar movies |

All three representations contain the same underlying information—the bipartite structure can be reconstructed from either projection plus the original edges. But each view makes certain analyses natural. The bipartite graph shows what happened; the projections show similarity patterns that support recommendation.

This flexibility is characteristic of graph analysis. The same data can be represented in multiple ways, each illuminating different aspects of the structure.

Our next example moves from implicit structure (ratings that imply similarity) to explicit structure (typed nodes, labeled edges, declared relationships). Knowledge graphs make the semantics of connections explicit, enabling richer reasoning.
