# Time Series Data {#sec-ts-data}

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo    = FALSE,
  error   = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r}
#| label: cran_libraries

library(assertthat)
library(astsa)
library(here)
library(knitr)
library(latex2exp)
library(tidyverse)
library(tinytex)
library(tsibble)
library(tsibbledata)
library(xts)
```

```{r}
#| label: conflicts-prefer

# resolve conflicts with plotly
# conflicted::conflicts_prefer(plotly::layout)

# resolve conflicts with tidymodels packages
# tidymodels::tidymodels_prefer()

```

```{r}
#| label: local-libraries

library(eda4mldata)

```

```{r}
#| label: local_source

# stabilize simulated data
source(here("code", "eda4ml_set_seed.R"))

# gen_ds_tbl
# compile a tibble that identifies data sets
source(here("code", "gen_ds_tbl.R"))

# get_ref_materials
# encapsulate reference materials as separate modules
source(here("code", "get_ref_materials.R"))
```

------------------------------------------------------------------------

## Introduction

Throughout this book we have seen that understanding and decision support are intertwined—that knowing *why* a method works is what allows you to recognize when it applies and when it doesn't. Time series analysis makes this interplay especially vivid.

The standard assumption underlying most statistical methods is that observations are independent: knowing the value of one observation tells you nothing about another. Time series data violate this assumption. Successive observations are dependent—today's temperature is informative about tomorrow's; this quarter's earnings relate to last quarter's; the sunspot count this month connects to counts in surrounding months.

This dependence is not merely a technical nuisance requiring adjusted formulas. It is *information*: information about how systems evolve, about underlying periodic structure, about the mechanisms that generate the data we observe. The autocorrelation structure of a time series is a fingerprint of the process that produced it. To ignore this structure is not just to get wrong standard errors—it is to miss what the data are telling us.

### Two Perspectives, One Reality

The dependence structure of a time series can be characterized in two complementary ways:

- **Time domain**: How does the present relate to the past? The autocorrelation function $\rho(u)$ measures the correlation between observations separated by $u$ time units. This perspective emphasizes prediction—using past values to forecast future ones.

- **Frequency domain**: What periodic components are present? The spectral density $f(\lambda)$ measures how much variance is attributable to oscillations at frequency $\lambda$. This perspective emphasizes understanding—identifying the rhythms and cycles that structure the process.

These two descriptions contain the same information. They are Fourier transform pairs: the spectrum is the transform of the autocovariance function, and vice versa. Yet they illuminate different aspects of the data. The time-domain view asks how the past predicts the future; the frequency-domain view asks what structure generated this process. Together, they provide a complete picture.

This duality is not accidental. There is a deep mathematical reason why Fourier analysis is fundamental to time series: the complex exponential functions $e^{i\lambda t}$ are eigenfunctions of the time-shift operator. Any linear, time-invariant system—a moving average, a filter, a physical process that doesn't change its character over time—has these sinusoidal functions as its natural modes. When we decompose a signal into frequency components, we are working in the natural coordinate system for such systems. We will develop this insight further in the sections that follow.

### Decision Support and Scientific Understanding

The two perspectives on time series—time domain and frequency domain—map onto the dual aims of data analysis.

Forecasting is the archetypal decision-support application of time series analysis. Given observations up to the present, what should we expect tomorrow, next quarter, next year? Time-domain methods, particularly ARIMA models, are the workhorses of this enterprise. They exploit the autocorrelation structure to extrapolate patterns into the future.

But effective forecasting rests on understanding. A forecast that merely extrapolates a trend will fail when conditions change. The question "what should we expect?" is inseparable from the question "why does the series behave as it does?" Climate data illustrate this forcefully: projecting future temperatures requires understanding the physical mechanisms—greenhouse gas forcing, feedback loops, natural variability—not merely curve-fitting to historical data. The frequency-domain perspective, which reveals the periodic structure and the filtering that shaped the series, serves this understanding.

The examples in this chapter span this range: sunspots, where periodic structure is primary; global temperatures, where trend and its causes are central; and ocean-atmosphere dynamics, where multiple periodicities interact. In each case, we examine the data from both perspectives, showing how they complement each other.

### Chapter Roadmap

This chapter introduces the conceptual and diagnostic foundations for time series analysis. We proceed as follows:

**Examples**: We begin with data—sunspots, global temperatures, the Southern Oscillation—examining each from both time-domain and frequency-domain perspectives. These examples build intuition for what the two views reveal.

**The White Noise Baseline**: White noise, a sequence of independent observations, serves as the reference point. Its autocorrelation function is zero at all nonzero lags; its spectrum is flat. Departures from this baseline indicate structure—the fingerprint of dependence.

**Stationarity and Time-Invariance**: We define stationarity, the condition under which the autocovariance and spectrum are well-defined and estimable from a single realization. Stationarity is the statistical counterpart of time-invariance in systems theory—the condition that a system's behavior does not depend on when we observe it.

**Autocorrelation and Partial Autocorrelation**: We develop the ACF and PACF as diagnostic tools, showing how their patterns reveal the dependence structure and guide model selection.

**Why It Matters for Inference**: We demonstrate that ignoring autocorrelation leads to incorrect variance estimates and misleading confidence intervals. The concept of effective sample size quantifies how much less information autocorrelated data contain than independent observations would.

**Looking Ahead**: We preview the subsequent chapters—time-domain methods (Chapter 13) and frequency-domain methods (Chapter 14)—which develop the modeling and estimation tools sketched here.

Throughout, we maintain the dual perspective: every diagnostic can be viewed in both time and frequency domains, and both views inform our understanding of what the data reveal.
